---
title: 'Craniosynostosis Multivariate Regression Analysis'
author: "Ryan Taylor, Joshua Lukemire, and Julia Wrobel"
date: "11-1-2024"
output:
  slidy_presentation:
    theme: paper
    highlight: haddock
    duration: 60
    footer: "Wrobel Lab"
  ioslides_presentation: default
editor_options:
  chunk_output_type: console
---

## Overview

\newcommand{\bA}{\mathbf{A}}
\newcommand{\bB}{\mathbf{B}}
\newcommand{\bC}{\mathbf{C}}
\newcommand{\bD}{\mathbf{D}}
\newcommand{\bf}{\mathbf{f}}
\newcommand{\bI}{\mathbf{I}}
\newcommand{\bP}{\mathbf{P}}
\newcommand{\bS}{\mathbf{S}}
\newcommand{\bs}{\mathbf{s}}
\newcommand{\bU}{\mathbf{U}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\bV}{\mathbf{V}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bW}{\mathbf{W}}
\newcommand{\bw}{\mathbf{w}}
\newcommand{\bX}{\mathbf{X}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\boldy}{\mathbf{y}}
\newcommand{\bZ}{\mathbf{Z}}
\newcommand{\bbeta}{\boldsymbol{\beta}}
\newcommand{\bgamma}{\mathbf{\gamma}}
\newcommand{\bepsilon}{\mathbf{\epsilon}}
\newcommand{\blambda}{\boldsymbol{\lambda}}
\newcommand{\bLambda}{\boldsymbol{\Lambda}}
\newcommand{\bSigma}{\boldsymbol{\Sigma}}
\newcommand{\bpsi}{\boldsymbol{\psi}}
\newcommand{\tR}{\mathbb{R}}
\newcommand{\Wbar}{\overline{W}}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator*{\argmax}{argmax}
\newcommand{\var}{\mathrm{Var}}

```{r setup, include=FALSE}
# Hide warnings and messages in output;
# Align images in center of page
knitr::opts_chunk$set(
  echo = FALSE,
  fig.align = "center",
  fig.dim = c(7, 5),
  include = TRUE,
  message = FALSE,
  warning = FALSE,
  results = 'markup'
)
# Clear memory
rm(list = ls()); gc()

# Set universal options
options(scipen = 999, # Don't use scientific notation
        knitr.kable.NA = "" # Hide NA's in printed tables
)

# Load packages
if(!require("pacman")) install.packages("pacman")
pacman::p_load(
  
  # Extras
  pwr, gtools, coda, numDeriv, mvtnorm, extraDistr, rstan, refund,
  latex2exp, pROC, glmnet, corrplot, ncvreg, readxl, sf, MASS,
  table1, gtsummary, ggridges, forestplot,
  broom.helpers, broom.mixed, patchwork, colorRamps, modelsummary,
  
  # Staples
  DescTools, tidyverse, magrittr, janitor,
  SASmarkdown, knitr, kableExtra, ggpubr, ggpmisc, stargazer
)

here::i_am("./analysis/2024.11_Collaborators.Rmd")
```

- **Goal: predict volume differences for each region by suture type**

- We use preprocessed Volumetric results from CT and 3D Photometry to create a multivariate regression associating regional volume differences with specific suture fusion diagnoses
  - This model uses a patient's suture fusion type to generate predictions and uncertainty for their 12 region-specific volume differences from their personalized reference head

- Our model shows a clear and reasonable pattern in volume differences associated with each suture fusion type, so we think we could be ready to develop a clinical paper

- However, a significant percentage of variance remains in the data
  - None of the demographics included in the data explain these residuals

- We have some remaining questions about the construction of the personalized normative reference head volume, and think a better understanding of this would help us to better understand the model's performance

## Data Description

```{r load data}
# File saved in "data" folder one level above repo
cranio_raw <- read_xlsx(here::here("../data/VolumetricResults-8-6-2024.xlsx"))

cranio <- cranio_raw %>% clean_names() 

sut <- cranio %>% select(matches("fused")) %>% 
  mutate(i = 1:n()) %>% 
  pivot_longer(cols = -i) %>%
  mutate(suture = if_else(value == 1, name, NA_character_)) %>% 
  group_by(i) %>% 
  summarize(sutures = paste(na.omit(suture), collapse = "/")) %>% 
  ungroup() %>% 
  arrange(i)

cranio %<>% 
  bind_cols(sut %>% select(sutures)) %>% 
  mutate(num_sutures = rowSums(select(., matches("fused"))),
         suture_type = case_when(num_sutures = 2 & 
                                   left_coronal_fused == 1 & 
                                   right_coronal_fused == 1 ~ "bicoronal",
                                 num_sutures > 1 ~ "multiple",
                                 num_sutures == 0 ~ "no_suture",
                                 T ~ sutures)) %>% 
  mutate(suture_type = str_remove_all(suture_type, "_fused"))
```

- We use the Volumetric Results file from 8/6/2024:
  - Each row is one picture
  - Based on the "UID" column, multiple pictures can come from the same patient
  - Demographics include Sex, Race, Ethnicity, Genetic Diagnosis, and Age at image (in days)
  - Separate columns indicate the presence of Metopic, Sagittal, Left coronal, Right coronal, Left lambdoid, and Right Lambdoid fusions
  - Overall Head Volume and Personalized Normative Reference Head Volume are given
  - Region-specific volume differences are given

- We clean the suture type column:
  - If a patient has both left coronal and right coronal sutures, we classify them as "Bicoronal" 
  - If they have another combination of multiple sutures, we classify them as "Multiple"
    - We eventually exclude these 
  - If they have no sutures, we classify them as "No suture"
    - These are the reference group in our analysis
  - Otherwise, we classify them based on the single suture they have

```{r clean data}

# Suture types
sut_type <- cranio %>% 
  group_by(suture_type) %>% 
  summarize(pics = n(),
            patients = n_distinct(uid)) %>% 
  ungroup() %>% 
  arrange(desc(pics)) %>% 
  mutate(pats_included = if_else(suture_type %in% c("multiple", "right_lambdoid", "left_lambdoid"), 0, patients),
         pct_pats = scales::label_percent(accuracy = 0.1)(patients / sum(patients)),
         pct_included = scales::label_percent(accuracy = 0.1)(pats_included / sum(pats_included))) %>% 
  adorn_totals("row") %>% 
  select(-pats_included)

# Summarize counts
cran_summ <- cranio %>% 
  summarize("Rows" = n(),
            "Unique Pictures" = n_distinct(image_identifier),
            "Unique Patients" = n_distinct(uid)) %>% 
  pivot_longer(everything(), names_to = "  ", values_to = " ")


# Filter to regression subset
cranio_reg <- cranio %>% 
  # Exclude rare sutures
  filter(!suture_type %in% c("multiple", "left_lambdoid", "right_lambdoid")) %>% 
  # Filter to one picture per patient
  arrange(uid, age_at_image_days) %>% 
  group_by(uid) %>% 
  slice_tail(n=1) %>% 
  ungroup() %>% 
  mutate(suture_type = factor(suture_type, levels = sut_type$suture_type))
```

## Multivariate Regression {.smaller}

- We run a multivariate linear regression model
  - Linear regression provides an estimate of the average head shape for a patient in each suture group
  - We are interested in modeling how regions' volumes vary together
    - Multivariate regression allows for modeling covariance across regions
      - See appendix for detail
  - Multivariate regression also allows for testing hypotheses about coefficients across regions
  - We can answer "How are volume differences in regions 1, 2, 3, and 4 related between the sagittal group and the metopic group?"

```{r reg}

cranio_reg %<>%
  mutate(head_volume_m_l_std = scale(head_volume_m_l))

vol_reg <- lm(
  formula(paste("cbind(",
                paste(
                  paste0("region_", 1:12, "_volume_difference_m_l"),
                  collapse = ", "),
                ") ~ ", "suture_type + head_volume_m_l_std")),
  data = cranio_reg
)
```

```{r brain shape}

# Manually define center points of regions
brain_points <- tribble(
  ~region, ~x, ~y,
  1, 1, 7,
  2, 2, 8, 
  3, 3, 8,
  4, 4, 7,
  5, 4, 5,
  6, 3, 4, 
  7, 4, 3,
  8, 3, 1,
  9, 2, 1,
  10, 1, 3,
  11, 2, 4,
  12, 1, 5
) %>% 
  st_as_sf(coords = c("x", "y"), remove = F)

# Create shapes around these points
g = st_combine(st_geometry(brain_points)) # make multipoint
v = st_voronoi(g)
v = st_collection_extract(v)

# Add region numbers
brain_shapes <- v[unlist(st_intersects(brain_points, v))] %>% 
  st_as_sf() %>% 
  mutate(region_num = 1:n())

# Create circle
brain_circle <- tibble(x = 2.5, y = 4.5) %>% 
  st_as_sf(coords = c("x", "y")) %>% 
  st_buffer(4.5)

brain_shapes <- st_intersection(brain_shapes, brain_circle)
```

```{r coeffs, fig.dim = c(7, 5), out.extra = 'style = "float:right"'}

# Extract and clean coefficients
reg_coeffs <- coefficients(vol_reg) %>% 
  as.data.frame() %>% 
  rownames_to_column() %>% 
  pivot_longer(-rowname) %>% 
  mutate(region_num = str_extract(name, "[0-9]+") %>% as.numeric(),
         suture_clean = str_remove_all(rowname, "suture_type") %>% 
           str_replace_all("_", " ") %>% 
           str_remove_all("\\(|\\)") %>% 
           str_to_title()) %>%
  mutate(suture_clean = case_when(suture_clean == "Head Volume M L Std" ~ "Head Volume", 
                                  T ~ suture_clean)) %>% 
  mutate(suture_clean = fct_inorder(suture_clean)) %>% 
  left_join(brain_shapes) %>% 
  st_as_sf()

# Plot coefficients
coeff_plot <- ggplot() +
  geom_sf(data = reg_coeffs, aes(fill = value)) +
  geom_sf_text(data = st_centroid(reg_coeffs), aes(label = region_num)) +
  facet_wrap(~suture_clean, nrow = 2) +
  scale_fill_viridis_c(option = "turbo") +
  labs(x = "", y = "", fill = "Estimate", title = "Expected Volume Differences by Suture Type")

coeff_plot
```

- Therefore, we run the model: 
$$y_i^{(d)} = \beta_0 + \sum_{k = 1}^5 \beta_k X_{ik} + \gamma_1 Z_{i1} + \epsilon_i^{(d)}$$

- $y_i^{(d)} =$ Volume Difference for Patient $i$ in region $d$ (mL)
- $X_{ik} =$ Binary indicator variable for Patient $i$'s fusion type: Sagittal, Metopic, Bicoronal, Right Coronal, or Left Coronal
- $Z_{i1} =$ Patient $i$'s Overall Head Volume (mL)

## Model Performance

- $R^2$ ranges from 0.10 to 0.32.

```{r beta table}

beta_tbl <- broom::tidy(vol_reg)

lm_stats <- distinct(beta_tbl, response) %>% 
  mutate(rse = purrr::map_dbl(1:length(summary(vol_reg)),
                              ~summary(vol_reg)[.][[1]]$sigma),
         r_2 = purrr::map_dbl(1:length(summary(vol_reg)),
                              ~summary(vol_reg)[.][[1]]$r.squared),
         adj_r_2 = purrr::map_dbl(1:length(summary(vol_reg)),
                                  ~summary(vol_reg)[.][[1]]$adj.r.squared))

beta_tbl %<>% left_join(lm_stats) %>% 
  select(response, rse, r_2, adj_r_2, everything()) %>% 
  mutate(low_95 = estimate - 1.96*std.error,
         high_95 = estimate + 1.96*std.error,
         region_num = str_extract(response, "[0-9]+") %>% as.numeric(),
         clean_var = str_remove_all(term, "suture_type"),
         term_name = paste0(region_num, ": ", clean_var))

lm_stats$r_2 <- cell_spec(round(lm_stats$r_2, 3), color = "white", bold = T, 
                          background = spec_color(lm_stats$r_2, option = "turbo"))
lm_stats$adj_r_2 <- cell_spec(round(lm_stats$adj_r_2, 3), color = "white", bold = T, 
                              background = spec_color(lm_stats$adj_r_2, option = "turbo"))

kbl(lm_stats, caption = "Model Performance Statistics", digits = 3,
    col.names = c("Region", "RSE", "R Squared", "Adj. R Squared"), escape = F) %>% 
  kable_styling(full_width = F)
```

## Next steps {.bigger}

- **Questions?**

- Outstanding issues for a paper?

## ***Appendix*** {.bigger}

- Data questions

- Correlation plots

- Age effects

- Data summaries

## Data Questions {.smaller}

- We created some summaries of the data in order to understand the volume differences (i.e. our outcome variable) better

- Suppose $\boldsymbol{Y}_i = (Y^{(1)}_i, \ Y_i^{(1)}, \ ..., \ Y_i^{(12)})$ is the $12 \times 1$ vector of region-specific volume differences for patient $i$; $V^{(A)}_i$ is patient $i$'s actual overall head volume, and $V^{(R)}_i$ is the volume of the personalized reference head volume for patient $i$. We expected:
  - $\sum_{j=1}^{12} Y_i^{(j)} = V_i^{(A)} - V_i^{(R)}$ for each patient
    - I.e. the sum of the region-specific volume differences is equal to the overall head volume minus the personalized reference head volume

- In the data, we observe that these metrics are similar, but not identical:

```{r head volume check}

vol_check <- cranio_reg %>% 
  mutate(vol_diff_sum = rowSums(select(., matches("region"))),
         vol_diff_sing = head_volume_m_l - personalized_normative_reference_head_volume_m_l)

ggplot(vol_check, aes(x = vol_diff_sing, y = vol_diff_sum)) +
  geom_point(aes(color = suture_type), alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "blue") +
  stat_poly_eq() +
  scale_color_viridis_d(option = "turbo") +
  labs(x = TeX("$V^{(A)} - V^{(R)}$"),
       y = TeX("$\\sum_J Y^{(J)}$"))
```

## Correlation Plots

- Within each suture type, the correlation matrices for residuals and coefficients take on the same general form. We show the volume difference, residual, and coefficient correlation matrix for sagittal fusions:

```{r cov matrices, results = 'hide', fig.dim = c(15, 15)}

Y <- vol_reg$model[[1]]

all_suts <- setdiff(unique(sort(cranio_reg$suture_type)), "no_suture")

print_set <- all_suts  # c("sagittal")

cor_y_list <- list()
cor_resid_list <- list()
cor_beta_list <- list()
cor_resid_shape_list <- list()

for(s in all_suts){
  
  s_nm <- s %>% str_replace_all("_", " ") %>% 
    str_remove_all("\\(|\\)") %>% 
    str_to_title()
  
  s_set <- which(cranio_reg$suture_type == s)
  
  cor_y <- cor(Y[s_set,]) %>% 
    as.data.frame() %>% 
    rownames_to_column() %>% 
    pivot_longer(cols = -rowname) %>% 
    mutate(across(c(rowname, name), 
                  ~str_extract(., "[0-9]+") %>% as.numeric() %>% as.factor()))
  
  cor_resid <- cor(vol_reg$residuals[s_set,]) %>% 
    as.data.frame() %>% 
    rownames_to_column() %>% 
    pivot_longer(cols = -rowname) %>% 
    mutate(across(c(rowname, name), 
                  ~str_extract(., "[0-9]+") %>% as.numeric() %>% as.factor()))
  
  cy <- ggplot(cor_y) + 
    geom_tile(aes(x = name, y = rowname, fill = value)) +
    scale_fill_viridis_c(option = "turbo", limits = c(-1, 1)) +
    scale_y_discrete(limits = rev) +
    labs(title = paste("Volume Corr.:", s),
         x = "Region", y = "Region", fill = "r")
  
  cr <- ggplot(cor_resid) + 
    geom_tile(aes(x = name, y = rowname, fill = value)) +
    scale_fill_viridis_c(option = "turbo", limits = c(-1, 1)) +
    scale_y_discrete(limits = rev) +
    labs(title = paste("Residual Corr.:", s),
         x = "Region", y = "Region", fill = "r")
  
  # Plot coefficient correlations
  cor_mx <- cov2cor(vcov(vol_reg)) %>% 
    as.data.frame() %>% 
    rownames_to_column() %>% 
    pivot_longer(-rowname, names_to = "colname") %>% 
    mutate(region_num_1 = str_extract(rowname, "[0-9]+") %>% 
             as.numeric() %>% as.factor(),
           region_num_2 = str_extract(colname, "[0-9]+") %>% 
             as.numeric() %>% as.factor(),
           variable_1 = str_match(rowname, "\\:(.*)")[,2] %>% 
             str_remove_all("suture_type") %>% 
             str_replace_all("_", " ") %>% 
             str_remove_all("\\(|\\)") %>% 
             str_to_title(),
           variable_2 = str_match(colname, "\\:(.*)")[,2] %>%  
             str_remove_all("suture_type") %>% 
             str_replace_all("_", " ") %>% 
             str_remove_all("\\(|\\)") %>% 
             str_to_title()
    )
  
  cor_mx_suture <- cor_mx %>% 
    filter(variable_1 == variable_2 &
             variable_1 != "Intercept")
  
  cor_plot_suture <- ggplot(cor_mx_suture %>% filter(variable_1 == s_nm)) +
    geom_tile(aes(x = region_num_1, y = region_num_2, fill = value)) +
    scale_fill_viridis_c(option = "turbo", limits = c(-1, 1)) +
    scale_y_discrete(limits = rev) +
    labs(title = paste("Coefficient Corr.:", s),
         x = "Region", y = "Region", fill = "r")
  
  resid_shapes <- cor_resid %>% 
    left_join(brain_shapes %>% 
                mutate(region_num = as.factor(region_num)), 
              by = c("name" = "region_num")) %>% 
    st_as_sf() %>% 
    mutate(is_d = if_else(name == rowname, "Yes", "No"),
           corr = if_else(name == rowname, NA_real_, value))
  
  resid_shape_plot <- ggplot() +
    geom_sf(data = resid_shapes, aes(fill = corr)) +
    geom_sf_text(data = st_centroid(resid_shapes), aes(label = name)) +
    facet_wrap(~rowname) +
    scale_fill_viridis_c(option = "turbo", na.value = "white", limits = c(-1, 1)) +
    labs(x = "", y = "", fill = "Correlation",
         title = paste("Residual Correlation:", s))
  
  # Save
  cor_y_list[[s]] <- cy
  cor_resid_list[[s]] <- cr
  cor_beta_list[[s]] <- cor_plot_suture
  cor_resid_shape_list[[s]] <- resid_shape_plot
  
}

wrap_plots(cor_y_list)

wrap_plots(cor_resid_shape_list, ncol = 2)
```

## Age

- Age is associated with overall head volume, but this association appears to be captured in the personalized reference head volume, so age is not associated with the volume differences.

```{r age plot, fig.dim = c(10, 8)}

# Age vs overall head volume
age_hv <- cranio_reg %>% 
  filter(age_at_image_days > 0) %>% 
  ggplot(aes(x = age_at_image_days, y = head_volume_m_l)) +
  geom_point(aes(color = suture_type), alpha = 0.25) +
  geom_smooth(formula = y ~ log(x)) +
  stat_cor(digits = 2, p.accuracy = 0.001) +
  scale_color_viridis_d(option = "turbo") +
  labs(title = "Age vs Overall Volume",
       x = "Age (days)",
       y = TeX("$V^{(A)}$ (mL)"))

# Age vs reference volume
age_rv <- cranio_reg %>% 
  filter(age_at_image_days > 0) %>% 
  ggplot(aes(x = age_at_image_days, y = personalized_normative_reference_head_volume_m_l)) +
  geom_point(aes(color = suture_type), alpha = 0.25) +
  geom_smooth(formula = y ~ log(x)) +
  stat_cor(digits = 2, p.accuracy = 0.001) +
  scale_color_viridis_d(option = "turbo") +
  labs(title = "Age vs Overall Reference Volume",
       x = "Age (days)",
       y = TeX("$V^{(R)}$ (mL)"))

# Age vs Overall Difference

age_diff <- cranio_reg %>% 
  filter(age_at_image_days > 0) %>% 
  ggplot(aes(x = age_at_image_days, y = (head_volume_m_l - personalized_normative_reference_head_volume_m_l))) +
  geom_point(aes(color = suture_type), alpha = 0.25) +
  geom_smooth(formula = y ~ log(x)) +
  stat_cor(digits = 2, p.accuracy = 0.001) +
  scale_color_viridis_d(option = "turbo") +
  labs(title = "Age vs Overall - Reference Volume",
       x = "Age (days)",
       y = TeX("$V^{(A)} - V^{(R)}$ (mL)"))

# Overall Volume vs Reference Volume
vol_comp <- cranio_reg %>% 
  filter(age_at_image_days > 0) %>% 
  ggplot(aes(x = head_volume_m_l, y = personalized_normative_reference_head_volume_m_l)) +
  geom_point(aes(color = suture_type), alpha = 0.25) +
  stat_cor(digits = 2, p.accuracy = 0.001) +
  scale_color_viridis_d(option = "turbo") +
  labs(title = "Actual Volume vs Overall Reference Volume",
       x = TeX("$V^{(A)}$ (mL)"),
       y = TeX("$V^{(R)}$ (mL)"))

# Age vs volume differences
age_y <- cranio_reg %>% 
  pivot_longer(cols = c(matches("region"))) %>% 
  filter(age_at_image_days > 0) %>% 
  mutate(region_num = str_extract(name, "[0-9]+") %>% as.numeric()) %>% 
  ggplot(aes(x = age_at_image_days, y = value)) +
  geom_point(aes(color = suture_type), alpha = 0.25) +
  stat_cor(digits = 2, p.accuracy = 0.001) +
  facet_wrap(~region_num) +
  scale_color_viridis_d(option = "turbo") +
  labs(title = "Age vs Region Volume Differences",
       x = "Age (days)",
       y = TeX("$Y_i^{(j)}$ (mL)"))

(age_hv | age_rv) /
(age_diff | vol_comp)

age_y
```

## Age < 1 Year

- If we limit the analysis to infants under 1 year old, the association between age and overall volume is closer to linear.
- We also have a slightly higher association between overall volume and volume differences

```{r age 1 yr}
# Age vs overall volume < 1 yr
age_hv_1 <- cranio_reg %>% 
  filter(age_at_image_days <= 365) %>% 
  ggplot(aes(x = age_at_image_days, y = head_volume_m_l)) +
  geom_point(aes(color = suture_type), alpha = 0.25) +
  geom_smooth(formula = y ~ x) +
  stat_cor(digits = 2, p.accuracy = 0.001) +
  scale_color_viridis_d(option = "turbo") +
  labs(title = "Age vs Overall Volume, < 1 Year",
       x = "Age (days)",
       y = TeX("$V^{(A)}$ (mL)"))

# Overall volume vs Volume differences
age_diff_1 <- cranio_reg %>% 
  filter(age_at_image_days <= 365) %>% 
  pivot_longer(cols = c(matches("region"))) %>% 
  filter(age_at_image_days > 0) %>% 
  mutate(region_num = str_extract(name, "[0-9]+") %>% as.numeric()) %>% 
  ggplot(aes(x = head_volume_m_l, y = value)) +
  geom_point(aes(color = suture_type), alpha = 0.25) +
  stat_cor(digits = 2, p.accuracy = 0.001) +
  facet_wrap(~region_num) +
  scale_color_viridis_d(option = "turbo") +
  labs(title = "Overall Volume vs Region Volume Differences, < 1 Year",
       x = TeX("$V^{(A)}$ (mL)"),
       y = TeX("$Y_i^{(j)}$ (mL)"))

# Reference volume vs Volume differences
vol_comp_1 <- cranio_reg %>% 
  filter(age_at_image_days <= 365) %>% 
  pivot_longer(cols = c(matches("region"))) %>% 
  filter(age_at_image_days > 0) %>% 
  mutate(region_num = str_extract(name, "[0-9]+") %>% as.numeric()) %>% 
  ggplot(aes(x = personalized_normative_reference_head_volume_m_l, y = value)) +
  geom_point(aes(color = suture_type), alpha = 0.25) +
  stat_cor(digits = 2, p.accuracy = 0.001) +
  facet_wrap(~region_num) +
  scale_color_viridis_d(option = "turbo") +
  labs(title = "Reference Volume vs Region Volume Differences, < 1 Year",
       x = TeX("$V^{(R)}$ (mL)"),
       y = TeX("$Y_i^{(j)}$ (mL)"))

age_hv_1

```

## Volume Difference Distributions {.smaller}

- Volume differences are generally right-tailed, and the distributions look similar across regions. Boxplots show there are only a very small number of outliers pulling the distribution so far out to the right.

```{r vol ridge}

vol_ridge <- cranio_reg %>% 
  select(matches("region")) %>% 
  pivot_longer(cols = everything()) %>% 
  mutate(region_num = str_extract(name, "[0-9]+") %>% as.numeric() %>% as.factor()) %>% 
  ggplot(aes(x=value, y=region_num, fill=region_num)) +
  geom_density_ridges() +
  labs(title = "Volume Difference Distributions")

vol_box <- cranio_reg %>% 
  select(matches("region")) %>% 
  pivot_longer(cols = everything()) %>% 
  mutate(region_num = str_extract(name, "[0-9]+") %>% as.numeric() %>% as.factor()) %>% 
  ggplot(aes(x=value, y=region_num, fill=region_num)) +
  geom_jitter(width=0.1, alpha = 0.25) +
  geom_boxplot() +
  labs(title = "Volume Difference Distributions")

vol_ridge | vol_box
```

## Data Summary: Multiple Pictures

```{r data summ}
kable(cran_summ, caption = "All pictures and patients") %>% 
  kable_styling(full_width = F)
```

```{r mult}
# Multiple pictures
multi_pic <- cranio %>% 
  arrange(uid, age_at_image_days) %>% 
  group_by(uid, suture_type) %>% 
  mutate(days_diff = age_at_image_days - lag(age_at_image_days, 1)) %>% 
  summarize(pics = n(),
            avg_day_diff = mean(days_diff, na.rm = T)) %>% 
  ungroup() %>% 
  filter(pics > 1) %>% 
  group_by(suture_type) %>% 
  summarize(patients = n_distinct(uid),
            max_pics = max(pics),
            avg_pics = mean(pics)) %>% 
  ungroup() %>% 
  mutate(pct_data = scales::label_percent(accuracy = 0.01)(patients / cran_summ$` `[3])) %>% 
  arrange(desc(patients))

kable(multi_pic, caption = "Multiple Pictures", 
      col.names = c("Suture Fusion", "Patients", "Max Pics / Patient", "Avg Pics / Patient", "% of Data"),
      digits = 2) %>% 
  kable_styling(full_width = F)
```

- A small percentage of patients had multiple visits, but any one patient was seen five times at most
  - This suggests that we cannot do any longitudinal analysis with the data
  - We also find that no patient's diagnosis changes over time

## Data Summary: Fusions

```{r sutures}
kable(sut_type, caption = "Suture Representation", digits = 2,
      col.names = c("Suture Type", "Pictures", "Patients", "% of Patients", "% of Included")) %>% 
  kable_styling(full_width = F, position = "float_right")
```

- Sagittal is the most common suture fusion type in the data
- Bicoronal fusion is more common than either left or right coronal
- Multiple, Left Lambdoid, and Right Lambdoid fusions are very uncommon

- Based on these summaries, we exclude Multiple, Left Lambdoid, and Right Lambdoid fusions, and then take the last picture taken chronologically for each patient

## Table 1

```{r table 1}

label(cranio_reg$image_type) <- "Image Type"
label(cranio_reg$pt_sex) <- "Sex"
label(cranio_reg$pt_race) <- "Race"
label(cranio_reg$pt_ethnicity) <- "Ethnicity"
label(cranio_reg$age_at_image_days) <- "Age (days)"
label(cranio_reg$head_volume_m_l) <- "Head Volume (mL)"
label(cranio_reg$personalized_normative_reference_head_volume_m_l) <- "Reference Head Volume (mL)"

demo_table <- table1::table1(~ 
                               image_type + pt_sex + pt_race + pt_ethnicity + 
                               age_at_image_days + head_volume_m_l + 
                               personalized_normative_reference_head_volume_m_l | suture_type,
                             data = cranio_reg)

demo_table
```
