---
title: "Malformation Analysis"
author: "Ryan Taylor"
format: 
  html:
    embed-resources: true
    theme: paper
    highlight: haddock
    code-fold: true
    toc: true
    toc-location: left
    code-overflow: wrap
    df-print: kable
    knitr:
      opts_chunk:
        echo: true
        include: true
        message: false
        warning: false
        results: markup
        fig.align: center
        fig.height: 5
        fig.width: 7
        R.options:
          scipen: 999
          knitr.kable.NA: ""
number-sections: true
editor: source
---

::: {.hidden}
$$
\newcommand{\bX}{\mathbf X}
\newcommand{\bY}{\mathbf Y}
\newcommand{\by}{\mathbf y}
\newcommand{\bH}{\mathbf H}
\newcommand{\bh}{\mathbf h}
\newcommand{\bx}{\mathbf x}
\newcommand{\bG}{\mathbf G}
\newcommand{\bg}{\mathbf g}
\newcommand{\bS}{\mathbf S}
\newcommand{\bs}{\mathbf s}
\newcommand{\bP}{\mathbf P}
\newcommand{\bC}{\mathbf C}
\newcommand{\bI}{\mathbf I}
\newcommand{\hatbG}{\widehat \bG}
\newcommand{\ba}{\mathbf a}
\newcommand{\bz}{\mathbf z}
\newcommand{\bb}{\mathbf b}
\newcommand{\bB}{\mathbf B}
\newcommand{\bw}{\mathbf w}
\newcommand{\bW}{\mathbf W}
\newcommand{\bZ}{\mathbf Z}

\newcommand{\balpha}{\boldsymbol{\alpha}}
\newcommand{\bbeta}{\boldsymbol{\beta}}
\newcommand{\bgamma}{\boldsymbol{\gamma}}
\newcommand{\bepsilon}{\boldsymbol{\epsilon}}
\newcommand{\hatbbeta}{\hat{\bbeta}}
\newcommand{\hatbgamma}{\hat{\bgamma}}
\newcommand{\bSigma}{\boldsymbol{\Sigma}}
\newcommand{\bsigma}{\boldsymbol{\sigma}}
\newcommand{\hatVar}{\widehat{\Var}}
\newcommand{\hatCov}{\widehat{\Cov}}
\newcommand{\bLambda}{\boldsymbol{\Lambda}}
\newcommand{\hatbLambda}{\hat{\bLambda}}
\newcommand{\blambda}{\boldsymbol{\lambda}}
\newcommand{\hatblambda}{\hat{\blambda}}
\newcommand{\btheta}{\boldsymbol{\theta}}
\newcommand{\hatbSigma}{\widehat{\boldsymbol{\Sigma}}}
\newcommand{\bmu}{\boldsymbol{\mu}}
\newcommand{\bzeta}{\boldsymbol{\zeta}}

\newcommand{\diag}{\mathrm{diag}}
\newcommand{\tr}{\mathrm{tr}}

\newcommand{\bzero}{\boldsymbol{0}}
\newcommand{\bone}{\boldsymbol{1}}
\def\iid{\,{\buildrel iid \over \sim}\,} 
\newcommand{\reals}{\mathbb{R}}
$$
:::

```{r}
#| label: setup
#| include: false

# Clear memory
rm(list = ls()); gc(verbose = F)

# Load packages
# if(!require("pacman")) install.packages("pacman")
pacman::p_load(
  # Specialty
  itsadug, spdep, concaveman,
  
  # Bayes
  mvtnorm,
  
  # Format
  latex2exp, table1, gtsummary,
  broom.helpers, broom.mixed, colorspace,
  
  # Modeling
  mgcv,
  
  # Useful extensions
  gtools, readxl, data.table, reshape2, ggridges, Matrix,
  
  # Basics
  DescTools, tidyverse, magrittr, janitor, glue, skimr,
  knitr, kableExtra, ggpubr, ggpmisc
)

# Load my Rcpp functions
library(DevCranio)

# Define function for predicting GAM on new data (in homage to get_predictions)
obtain_predictions <- function(model, newdata, 
                               rm_effects = c("~"), # By default match nothing; assume no variable includes tilde
                               ci_crit = 1.96){
  
  x <- newdata
  
  # Get design matrix of basis functions
  nd_mat <- predict(model, newdata = x, type = "lpmatrix")
  
  # Find the columns for effects we want to remove
  rm_cols <- map(rm_effects, 
                 ~grepv(., colnames(nd_mat), fixed = T)) %>% 
    unlist() %>% unique()
  
  # For these columns, set (linear predictor) columns to 0
  nd_mat[, rm_cols] <- 0
  
  # Get coefficients of the model
  m_coef <- coef(model)
  
  # Calculate predictions
  x$fit <- (nd_mat %*% m_coef) %>% as.vector()
  
  # Calculate variances (without calculating all n x n covariances)
  x$se <- (nd_mat %*% vcov(model) * nd_mat) %>% rowSums() %>% sqrt()
  
  # Add confidence interval
  x %<>% mutate(
    ci_ll = fit - ci_crit * se,
    ci_ul = fit + ci_crit * se
  )
  
  return(x)
}

# Set switches
DO_LOAD_MATRICES <- F
DO_PRELIM_MODEL_REGION <- F
DO_PRELIM_MODEL_PIXEL <- F
DO_PRELIM_MODEL_FUSION <- F
DO_INDIV_SOAPS <- F

# Set constants
RM_IGNORE <- "DO|IGNORE|FOLDER|cranio|_dict|so_"
```

# Overview

This document consolidates our work on a Bayesian hierarchical model using a soap-film smoother to describe the patterns of volume abnormalities in children with craniosynostosis.

## Updates

-   Updated estimation of demographic effects on one soap film basis function: unpenalized parameters now have corrected prior variance term and we use the same lambda for all of the separate suture fusion types. (@sec-coeff-gibbs)
-   Wrote out prior assumptions with more lengthy definitions and vectorization / Kronecker products (@sec-joint-model-spec)

## Questions

-   What is the best way to calculate confidence intervals for the basis coefficient predictions? (For the 1-coefficient model, I am currently making a prediction for each individual draw of betas, then taking percentiles from these samples)

## Collaborator Questions

-   7 patients in the data have *sphenoidal involvement* marked as 1. What effect might this have on volumes?
-   18 patients have *previous surgery* marked as 1. Is this specifically indicating a previous surgery related to craniosynostosis?
-   Should we control for image type in later analyses?
-   Looking at a subset of the data, we can see that the images "sub-CNH_0002_dol-1295" and "sub-CNH_Cranio_00002_dol-61" have the exact same pointwise differences, even though the first is marked as having no fusion and the second is marked as having a Sagittal fusion. Are these the same image?

# Covariates Data

Data were sent by collaborators at Colorado (primarily Antonio Porras, Alejandro Cruz Guerrero, and Connor Elkhill).

Covariates are stored in a "merged_covariates" csv file that Josh Lukemire created from these data. Each image's malformation map is stored as a $500 \times 500$ matrix in its own RData file saved in a data folder in Josh's OneDrive.

We clean the covariate data set for easier analysis and to identify rare subsets of the data that should be excluded from our early work.

```{r}
#| label: load-data

# Set directory for references
here::i_am("2025.08_Malformation_Analysis.qmd")

# Set data folder
DATA_FOLDER <- here::here("../../data/Lukemire, Joshua's files - data/")
COVARS_FOLDER <- here::here("../../data/SphericalMaps_database/SphericalMaps_database/")

# Set subfolder with image data
MATRIX_FOLDER <- file.path(DATA_FOLDER, "unsigned_distances")

# List all files with any extension (so no folders)
matrix_files <- list.files(MATRIX_FOLDER, pattern = "\\.")

# Read in covariates data
patient_images <- read_csv(file.path(DATA_FOLDER, "merged_covariates.csv"))
```

First, we ensure that all images in the covariate data are represented in the data folder and vice versa.

```{r}
#| label: clean-file-names

### Clean cranio covariates file

## Create new data set with easier names
cranio <- patient_images %>% 
  clean_names()

## Check if files in data were found in data folder
cranio %<>%
  mutate(
    fname = paste0(id_fmt, "_desc-unsigneddist.RData"),
    fname_exists = if_else(fname %in% matrix_files, 1, 0)
  )

# Print a table to check we have all files
fname_summ <- tibble(
  in_data_not_folder = sum(1 - cranio$fname_exists),
  in_folder_not_data = sum(!matrix_files %in% cranio$fname)
)

kbl(fname_summ, caption = "Check Files vs. Data Availability") %>% 
  kable_styling(full_width = F)

```

## Suture Fusion Categories

Next, we create a cleaner categorization of each image's suture fusions. Information about the suture fusions for each image are initially stored in a series of binary columns: *metopic_fused, sagittal_fused, left_coronal_fused, right_coronal_fused, left_lambdoid_fused, right_lambdoid_fused*. We combine these into one column and create categories for "Bicoronal" and "Multiple."

```{r}

#| label: clean-sutures

## Classify sutures

# Currently stored in binary columns with names ending in "fused"
# Reshape data to identify all combinations of sutures
fusion_df <- cranio %>% 
  select(id_fmt, matches("fused")) %>% 
  pivot_longer(cols = -id_fmt) %>% 
  mutate(
    name = name %>% 
      str_remove("_fused") %>% 
      str_replace_all("_", " ") %>% 
      str_to_title()
  ) %>% 
  mutate(fusion = if_else(value == 1, name, NA_character_)) %>% 
  group_by(id_fmt) %>% 
  summarize(num_fusions = sum(value, na.rm = T),
            fusions = paste(na.omit(fusion), collapse = "/")) %>% 
  ungroup()

# (1) count number of sutures
# (2) identify bicoronal (the only useful instance of multiple sutures)
# (3) identify other instances of multiple sutures
# (3) assign remaining suture types
cranio %<>%
  left_join(fusion_df, by = "id_fmt") %>% 
  mutate(fusion_type = case_when(
    num_fusions == 0 ~ "Normative",
    fusions == "Left Coronal/Right Coronal" ~ "Bicoronal",
    num_fusions > 1 ~ "Multiple",
    T ~ fusions
  )) %>% 
  mutate(fusion_type = fct_infreq(fusion_type)) %>% 
  mutate(fusion_type = fct_relevel(fusion_type, "Multiple", after = Inf))

# Summarize frequencies of each suture type
fusion_summ <- cranio %>% 
  group_by(fusion_type) %>% 
  summarize(pics = n(),
            patients = n_distinct(id_base)) %>% 
  ungroup() %>% 
  arrange(desc(pics)) %>% 
  mutate(pics_pct = pics / sum(pics),
         patients_pct = patients / sum(patients)) %>% 
  adorn_totals("row")

kbl(fusion_summ, caption = "Suture Fusions: Full Data") %>% 
  kable_styling(full_width = F)
```

## Longitudinal Visits {#sec-longitudinal}

Next, we look at the availability of longitudinal data. No longitudinal data exist for patients with no suture fusions. Patients with suture fusions have an average of about 1.5 visits per person, and this average is fairly consistent even when we limit to the number of visits taken before the 1 year mark.

```{r}
#| label: visits

# Identify which visit each image is from and how many visits they had
cranio %<>%
  arrange(id_base, age) %>% 
  group_by(id_base) %>% 
  mutate(visit_num = 1:n()) %>% 
  # Count visits total and visits under age limit (currently 1 year)
  mutate(visits = max(visit_num)) %>% 
  ungroup()

# Summarize visits on average and max by fusion group
visits_summ <- cranio %>%
  group_by(fusion_type) %>% 
  summarize(patients = n_distinct(id_base),
            pics = n(),
            visits_avg = mean(visits),
            visits_max = max(visits)) %>%
  ungroup() %>% 
  arrange(desc(visits_avg))
  
kbl(visits_summ, caption = "Longitudinal Visits: Full Data") %>% 
  kable_styling(full_width = F)
```

## Exclusion Criteria {#sec-exclusion}

Next, we consider exclusion criteria.

First, there are a small number of patients who were labeled by the collaborators as "Craniosynostosis," but have no suture fusions. Close inspection reveals that these are 6 of the 7 patients in the data with "sphenoidal involvement." Therefore, we exclude these 7 images.

```{r}
#| label: group-and-sphenoidal

## Investigate any issues with fusion type
# Sanity check: fusion category against collaborators' "group" column
fusion_group_summ <- tabyl(cranio, fusion_type, group)

kbl(fusion_group_summ, caption = "Collaborators' Group vs. Fusions") %>% 
  kable_styling(full_width = F)

# 6 in "Craniosynostosis" group with no sutures all have sphenoidal involvement
sphen_summ <- cranio %>% 
  group_by(fusion_type, group, sphenoidal_involvement) %>% 
  summarize(n = n()) %>% 
  ungroup() %>% 
  arrange(desc(n))

kbl(sphen_summ, caption = "Sphenoidal Involvement Summary") %>% 
  kable_styling(full_width = F)
```

Second, it is worth investigating the category of "Multiple" fusions. We find that other than the "bicoronal" combination of fusions (left coronal and right coronal fusions but nothing else), no other combination has a sample size of more than 14. Therefore, we exclude 81 patients with "Multiple" fusions.

Third, we likewise find that *Right Lambdoid* and *Left Lambdoid* suture fusions are very rare in our data, with 10 and 7 images respectively. Therefore, we exclude lambdoid suture fusions.

```{r}
#| label: multiple-sutures

## Investigate multiple sutures

fusion_multiple_summ <- cranio %>% 
  filter(fusion_type == "Multiple") %>% 
  group_by(num_fusions, fusions) %>% 
  summarize(pics = n(),
            patients = n_distinct(id_base)) %>% 
  ungroup() %>% 
  arrange(desc(pics))

kbl(fusion_multiple_summ, caption = "Summary of Multiple Fusions") %>% 
  kable_styling(full_width = F)

max_multiple_count <- max(fusion_multiple_summ$pics)
```

Fourth, we find that 18 images in the data are associated with a "previous surgery." Since we are more interested in the patterns associated with suture fusions before surgical intervention, we exclude these images. Notably, 6 of these 18 images have multiple suture fusions, so they would already be excluded.

```{r}
#| label: previous-surgery

## Investigate previous surgery
surg_summ <- cranio %>% 
  filter(previous_surgery == 1) %>% 
  group_by(fusion_type, group, sphenoidal_involvement) %>% 
  summarize(n = n()) %>% 
  ungroup() %>% 
  arrange(desc(n)) %>% 
  adorn_totals("row")

kbl(surg_summ, caption = "Images with Previous Surgery") %>% 
  kable_styling(full_width = F)
```

### Age Summary {#sec-age-summary}

Fifth, we check the distribution of age in our data. We determine that over 98\% of the data falls under the age of 5 years, so we exclude those above this age, since we have very sparse observations to estimate an age effect.

```{r}
#| label: age-summary

# Set colors for fusions to remain consistent in future plots
fusion_color_dict <- distinct(cranio, fusion_type) %>% 
  arrange(fusion_type) %>% 
  mutate(color = c(palette.colors(palette = "Set1")[-c(6, 9)],
                   palette.colors(palette = "Set1")[c(6, 9)]))

# Plot age densities
age_plot <- ggplot(cranio) +
  geom_density_ridges(aes(x = age, 
                          y =  fusion_type, 
                          fill = fusion_type)) +
  scale_fill_discrete(guide = "none", 
                      palette = fusion_color_dict$fusion_color) +
  labs(x = "Age (Days)", y = "Fusion Type")

age_plot

# Summarize densities
have_fusions <- cranio %>% 
  filter(fusion_type != "Normative")

have_fusions_quant <- have_fusions %$%
  quantile(age, probs = c(0.9, 0.95, 0.975, 0.99, 1)) %>% 
  enframe(name = "Percentile (with Fusions)", value = "Age") %>% 
  mutate(Years = Age / 365.25)

kbl(have_fusions_quant, 
    caption = "Max. Age Options: Patients with Fusions") %>% 
  kable_styling(full_width = F)

# Set maximum age as 3 years old
cranio_max_age <- 365 * 3

# Determine what percentile this corresponds to among patients with fusion
max_age_summ <- tibble("Age Cap" = cranio_max_age, 
                       "Percentile" = mean(have_fusions$age <= cranio_max_age))

kbl(max_age_summ, caption = "Age cap summary") %>% kable_styling(full_width = F)
```

# Pointwise Difference Matrix Data {#sec-load-matrices}

Next, we load the malformation maps into one data set. Each malformation map is stored as a $500 \times 500$ matrix of pointwise absolute value differences, and is saved in its own RData file. We load each RData file in the remaining subset of images selected above. Then, we reshape the data from wide to long, so that each row contains one pixel's row, column, and pointwise difference value. We then store this long data as a tibble nested in the "data" column. Finally, we rearrange these data so that we can work with a data set with one row for each pixel, containing all participants' pointwise differences in that pixel. In addition, we merge each pixel's region label to this pointwise matrix data, so that we can associate this analysis with previous analyses.

```{r}
#| results: hide

# Clear memory space
rm(list = setdiff(ls(), c(lsf.str(), ls(pattern = eval(RM_IGNORE)))))
gc(verbose = F)
```


```{r}
#| label: load-matrix-files

# Define function to read each file
load_matrix_data <- function(x){
  
  # Path to data file
  full_path <- file.path(MATRIX_FOLDER, x)
  
  # Load RData file (pointwise_diff)
  load(full_path)
  
  # Convert matrix to long format and remove values outside mask
  reduced_data <- pointwise_diff %>%
    reshape2::melt(varnames = c("row", "col"), value.name = "diff") %>%
    as_tibble() %>%
    drop_na(diff)
  
  return(list(reduced_data))
}

# Toggle whether to create matrix data
if(DO_LOAD_MATRICES){
  # Create minimal matrix data
  cranio_matrix <- cranio %>%
    select(fname, age, sex, fusion_type) %>% 
    rowwise() %>%
    mutate(
      raw_data = load_matrix_data(fname)
    ) %>%
    ungroup()
  
  # Rearrange so each row is a pixel
  cranio_matrix %<>% 
    unnest(raw_data) %>%
    group_by(row, col) %>%
    nest() %>% 
    ungroup()
  
  # Load regions and mask
  load(paste0(DATA_FOLDER, "region_labels_and_mask.RData"))
  
  cranio_matrix %<>%
    left_join(region_labels,
              by = c("row" = "X", "col" = "Y")) %>% 
    filter(value != 0) %>% 
    rename(region = value)
  
  saveRDS(cranio_matrix, 
          file = here::here("intermediate", 
                            "covars_and_matrices.rds"))
} else{ 
  cranio_matrix <- readRDS(file = here::here("intermediate", 
                                             "covars_and_matrices.rds")) }

# Flip axes to match overhead view
cranio_matrix %<>%
  mutate(row = -row, 
         col = -col)
```

We use these matrices to identify duplicates in the data. Currently, these are still included in our analyses.

```{r}
#| label: check-duplicate-data

# Find instances where average of all observations is the same
dup_avgs <- cranio_matrix %>% 
  unnest(data) %>% 
  select(fname, fusion_type, age, row, col, diff) %>% 
  group_by(fname, fusion_type, age) %>% 
  nest() %>% 
  mutate(diff_avg = map_dbl(data, ~mean(.$diff))) %>% 
  select(-data) %>% 
  group_by(diff_avg) %>% 
  mutate(n = n()) %>% 
  ungroup() %>% 
  filter(n > 1)

cranio_dup_fnames <- unique(dup_avgs$fname)

# Reshape these to 1 per set of duplicates
dups_collected <- dup_avgs %>% 
  group_by(diff_avg, n) %>% 
  summarize(fnames = paste(na.omit(unique(fname)), collapse = "/"),
            fusions = paste(na.omit(unique(fusion_type)), collapse = "/"),
            ages = paste(na.omit(unique(age)), collapse = "/")) %>% 
  ungroup() %>% 
  arrange(desc(n), diff_avg)

# Add information about these duplicate pairs
dups_collected %<>% 
  # Check study
  mutate(is_cnh = if_else(str_detect(fnames, "CNH"), 1, 0),
         is_chco = if_else(str_detect(fnames, "CHCO"), 1, 0))

dup_by_study <- dups_collected %>% 
  group_by(is_cnh, is_chco) %>% 
  summarize(n = n()) %>% 
  ungroup()

```

### Data Subset {#sec-subset-data}

```{r}
#| label: filtered-data

### Create subset data
# Exclude rare fusion categories: multiple, lambdoid
# Exclude sphenoidal involvement (in Craniosynostosis group without suture)
# Exclude patients with previous surgery (10)
# Exclude patients above age cap of 3 yrs
# Only keep last picture taken
cranio_sub <- cranio %>% 
  filter(!fusion_type %in% c("Multiple", 
                             "Left Lambdoid", "Right Lambdoid") &
           sphenoidal_involvement %in% c(NA, 0) &
           previous_surgery %in% c(NA, 0) &
           age <= cranio_max_age &
           visit_num == visits) %>% 
  # Reduce number of columns
  select(id_base, fname, fusion_type,
         age, sex) %>% 
  arrange(id_base, age) %>% 
  # Make sex variable categorical; drop unused fusion types
  mutate(sex = factor(sex),
         fusion_type = fct_drop(fusion_type))

# Clear space
rm(cranio)
```

```{r}
#| label: filter-to-subset

# Filter pointwise data to same subset
cranio_matrix <- cranio_matrix %>% 
  unnest(data) %>% 
  filter(fname %in% cranio_sub$fname) %>% 
  group_by(row, col, region) %>% 
  nest() %>% 
  ungroup()
```

We generate a summary of the fusion types present in the remaining observations, to compare against the full data.

```{r}
#| label: summarize-cleaned-data

# Summarize counts of images and patients by fusion type
fusion_summ_sub <- cranio_sub %>% 
  group_by(fusion_type) %>% 
  summarize(pics = n(),
            patients = n_distinct(id_base)) %>% 
  ungroup() %>% 
  arrange(desc(pics)) %>% 
  mutate(pics_pct = pics / sum(pics),
         patients_pct = patients / sum(patients)) %>% 
  adorn_totals("row")

kbl(fusion_summ_sub, caption = "Fusions in Subset Data") %>% 
  kable_styling(full_width = F)
```

We also create a Table 1, to identify what representation we have across key demographics.

```{r}
#| label: table-1

table1(~age + sex | fusion_type,
       data = cranio_sub)
```


## Growth Matrix Visualizations {#sec-matrix-viz}

First, to understand the first way we can aggregate the data spatially, we plot the collaborators' defined regions by pixel.

### Region Map

```{r}
#| label: plot-regions

# Create dataset of only region assignment for each pixel
matrix_region <- cranio_matrix %>% 
  select(row, col, region) %>% 
  distinct()

# Identify regions' centroids for labels
matrix_region_center <- matrix_region %>% 
  group_by(region) %>% 
  summarize(row_center = mean(row),
            col_center = mean(col)) %>% 
  ungroup()

# Plot regions
plot_regions <- ggplot() +
  geom_raster(data = matrix_region, 
              aes(x = row, y = col, fill = factor(region))) +
  geom_text(data = matrix_region_center, 
            aes(x = row_center, y = col_center, label = factor(region))) +
  coord_fixed() +
  theme_void() +
  scale_fill_brewer(palette = "Set3", guide = "none") +
  labs(title = "Region Map", fill = "Region")

plot_regions
```

### Fusion Averages

Next, we summarize the growth by suture fusion and plot these averages on the brain shape.

```{r}
#| results: hide

# Clear memory space
rm(list = setdiff(ls(), c(lsf.str(), ls(pattern = eval(RM_IGNORE)))))
gc(verbose = F)
```

```{r}
#| label: map-fusions

### Summarize differences by fusions

# Create dataset of average cell for each fusion
matrix_fusion <- cranio_matrix %>% 
  # Extract only fusion type and outcome from nested data
  hoist(data, "fusion_type", "diff") %>% 
  select(-c(data, region)) %>% 
  unnest_longer(col = c(fusion_type, diff)) %>% 
  # Summarize average by pixel and fusion type
  group_by(row, col, fusion_type) %>% 
  summarize(diff_avg = mean(diff)) %>% 
  ungroup()

# Plot brain map by fusion type
plot_fusion <- ggplot(matrix_fusion) +
  geom_raster(aes(x = row, y = col, fill = diff_avg)) +
  facet_wrap(~fusion_type) +
  coord_fixed() +
  theme_void() +
  scale_fill_viridis_c(option = "turbo") +
  labs(title = "Fusion Average Growth",
       fill = "Avg. Growth") +
  theme(legend.position = "bottom")

# Print fusion maps
plot_fusion
```

### Fusion - Age Averages

We add to this summary by showing how these averages change on average over time, dividing age into 6 bins of roughly 2-month windows of ages. We also compare these results to the sample sizes in each of these age-fusion bins.

```{r}
#| label: plot-fusion-and-age

### Investigate age

# Define age category cutoffs
age_breaks <- c(0, 61, 122, 183, 366, cranio_max_age)

# Label these with spaces so they can wrap in plots
age_break_labels <- map_chr(2:length(age_breaks), 
                            ~paste0(age_breaks[.-1], 
                                    " - ", 
                                    if_else(. == length(age_breaks), "", "<"), 
                                    age_breaks[.]))

# Create dataset with fusion and age info
matrix_age_fusion <- cranio_matrix %>% 
  hoist(data, "fusion_type", "age", "diff") %>% 
  select(-c(data, region)) %>% 
  unnest_longer(col = c(fusion_type, age, diff)) %>% 
  # Add age categories
  mutate(age_bin = cut(age, 
                       breaks = age_breaks, labels = age_break_labels, 
                       right = FALSE, include.lowest = T)) %>% 
  # Summarize average in these bins
  group_by(row, col, fusion_type, age_bin) %>% 
  summarize(diff_avg = mean(diff)) %>% 
  ungroup()

# Plot brain map by fusion type and age
plot_fusion_age <- ggplot(matrix_age_fusion) +
  geom_raster(aes(x = row, y = col, fill = diff_avg)) +
  facet_grid(fusion_type ~ age_bin,
             labeller = label_wrap_gen(width = 10)) +
  coord_fixed() +
  scale_fill_viridis_c(option = "turbo") +
  theme_void() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "Fusion - Age Average Growth",
       fill = "Avg. Growth") +
  theme(legend.position = "bottom")

# Print fusion maps
plot_fusion_age

## Determine sample size for these averages

# Count images per fusion and age bin
size_fusion_age <- cranio_sub %>% 
  mutate(age_bin = cut(age, 
                       breaks = age_breaks, labels = age_break_labels, 
                       right = FALSE, include.lowest = T)) %>% 
  group_by(fusion_type, age_bin) %>% 
  summarize(n = n()) %>% 
  ungroup() %>% 
  pivot_wider(names_from = age_bin, values_from = n, id_cols = fusion_type)

# Print table
kbl(size_fusion_age, caption = "Fusion-Age Sample Size") %>% 
  kable_styling(full_width = F)
```

### Age Trajectory {#sec-age-smooth}

In order to gain a better understanding of the specific shape of the age effect, we plot average growth by age, stratified by region and fusion type.

```{r}
#| label: age-effect

# Check shape of age effect by region and fusion

# Summarize by age, region, and fusion type; have option to round to nearest month
matrix_age_region <- cranio_matrix %>% 
  hoist(data, "fusion_type", "age", "diff") %>% 
  select(-data) %>% 
  unnest_longer(col = c(fusion_type, age, diff)) %>% 
  mutate(age_round_month = round(age / 30)) %>% 
  group_by(fusion_type, age, region) %>% 
  summarize(diff_avg = mean(diff)) %>% 
  ungroup()

plot_age <- ggplot(matrix_age_region) +
  # Plot 1 line over age for each region-fusion combo
  geom_smooth(aes(x = age, y = diff_avg, 
                  color = fusion_type, fill = fusion_type),
              alpha = 0.2) +
  facet_wrap(~region) +
  scale_color_discrete(palette = fusion_color_dict$color) +
  scale_fill_discrete(palette = fusion_color_dict$color) +
  labs(title = "Fusion - Region Average Growth by Age",
       x = "Age (Days)",
       y = "Avg. Growth",
       color = "Fusion Type",
       fill = "Fusion Type") +
  theme_minimal() +
  theme(legend.position = "bottom")

# Print plot
plot_age
```

Finally, we consider how each fusion type might differ between the two natal sexes recorded in the data.

```{r}
#| label: fusion-by-sex

# Take average pointwise differences in cell by fusion and sex
matrix_sex_fusion <- cranio_matrix %>% 
  hoist(data, "fusion_type", "sex", "diff") %>% 
  select(-c(data, region)) %>% 
  unnest_longer(col = c(fusion_type, sex, diff)) %>% 
  # Summarize average in these bins
  group_by(row, col, fusion_type, sex) %>% 
  summarize(diff_avg = mean(diff)) %>% 
  ungroup()

# Plot brain map by fusion type and age
plot_fusion_sex <- ggplot(matrix_sex_fusion) +
  geom_raster(aes(x = row, y = col, fill = diff_avg)) +
  facet_grid(fusion_type ~ sex,
             labeller = label_wrap_gen(width = 9)) +
  coord_fixed() +
  scale_fill_viridis_c(option = "turbo") +
  theme_void() + 
  theme(axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "Fusion-Sex Average Growth",
       fill = "Avg. Growth") +
  theme(legend.position = "bottom")

# Print fusion maps
plot_fusion_sex
```

# Matrix Boundary

To prepare for preliminary models, we construct the boundary of our matrix data, differentiating the area in which we have data and the area outside of it. To construct this boundary, we need to address two issues: first, we need to define the boundary such that none of our data overlap with it; and second, we need to order the points that make up the boundary so that the boundary can be constructed by simply connecting these points with line segments *in order*. Therefore, we construct this boundary in a few steps. First, we take the concave hull of all points in our matrix; this approach is not be generalizable to the case where there are multiple loops to account for in the soap film, but it works for our case, where we have one continuous boundary. Next, we add a buffer of 1 unit, so that none of our data will overlap with the boundary. We then convert this shape to a boundary line and "sample" points at 1-unit intervals, ensuring that we have even coverage of the resulting boundary shape. These points are passed as a list to the soap film function. The boundary resulting from this process is depicted below.

```{r}
#| label: cranio-boundary

## Create boundary around cranium shape

# Create polygon around pixels in cranium shape
cranio_bound_shape <- cranio_matrix %>% 
  distinct(row, col) %>% 
  st_as_sf(coords = c("row", "col")) %>% 
  concaveman(concavity = 1)

# Create boundary as points
cranio_bound_matrix <- cranio_bound_shape %>% 
  # Add buffer so boundary is beyond data
  st_buffer(1) %>% 
  # Convert to line
  st_boundary() %>% 
  # Take points along this line at density = 1 unit
  st_line_sample(density = 1) %>% 
  st_coordinates() %>% 
  as_tibble() %>% 
  rename(row = X, col = Y) %>% 
  # Add order of appearance in data
  mutate(appearance_order = 1:n())

# Visualize how boundary compares to all points
bound_viz <- ggplot() +
  geom_point(data = cranio_matrix %>% select(row, col),
             aes(x = row, y = col), alpha = 0.5, size = 0.0001) +
  geom_point(data = cranio_bound_matrix,
            aes(x = row, y = col, color = appearance_order)) +
  scale_color_viridis_c(option = "mako") +
  coord_fixed() +
  labs(title = "Boundary Sanity Check",
       color = "Order of Appearance") +
  theme_minimal() +
  theme(legend.position = "bottom")

bound_viz
```

# Preliminary Models {#sec-prelim-models}

## Region-Level Independent

Next, we get a better understanding of how these demographics interact with fusion type through preliminary regression models. First, we regress the region-level average growth for each subject on fusion indicators, age, and sex. Based on the plot above, we include age as a smooth term; all other covariates are included as indicator variables. We also run a multiple linear regression with age as a linear term.

We denote the average growth for patient $i$ in region $r$ as $G_{ir}$. We let $f_r(\cdot)$ be a cubic regression spline estimated on observations from region $r$; we choose a cubic regression spline for faster estimation of many regressions, especially when we extend to a pixel-level model below. Then these models can be written:

$$
\begin{gather*}
\begin{split}
G_{ir} = {} & \beta_{0, r} + \beta_{1,r} [\text{Sagittal}]_i + \beta_{2,r} [\text{Metopic}]_i + \\
& \beta_{3,r} [\text{Right Coronal}]_i + \beta_{4,r} [\text{Left Coronal}]_i + \\
& \beta_{5,r} [\text{Left and Right Coronal}]_i + \\
& \beta_{6,r} [\text{Sex: Male}]_i + f_r(Age_i) + \epsilon_{ir}
\end{split} \\ \\
\begin{split}
G_{ir} = {} & \beta_{0, r} + \beta_{1,r} [\text{Sagittal}]_i + \beta_{2,r} [\text{Metopic}]_i + \\
&\beta_{3,r} [\text{Right Coronal}]_i + \beta_{4,r} [\text{Left Coronal}]_i + \\
& \beta_{5,r} [\text{Left and Right Coronal}]_i + \\
& \beta_{6,r} [\text{Sex: Male}]_i + \beta_{7,r}(Age_i) + \epsilon_{ir}
\end{split} \\ \\
\begin{split}
\epsilon_{ir} \iid {} & N(0, \sigma^2_r)
\end{split}
\end{gather*}
$$

```{r}
#| results: hide

# Clear memory space
rm(list = setdiff(ls(), c(lsf.str(), ls(pattern = eval(RM_IGNORE)))))
gc(verbose = F)
```

```{r}
#| label: region-model

# Add fusion variables for regression
fusion_dict <- tibble(
  fusion_type = c("Normative", 
                  "Sagittal", "Metopic", 
                  "Right Coronal", "Bicoronal", "Left Coronal"),
  fused_Sagittal = c(0, 1, 0, 0, 0, 0),
  fused_Metopic = c(0, 0, 1, 0, 0, 0),
  fused_RCoronal = c(0, 0, 0, 1, 1, 0),
  fused_LCoronal = c(0, 0, 0, 0, 1, 1)
)

# Recreate pixel map with region labels
matrix_region <- cranio_matrix %>% 
  select(row, col, region) %>% 
  distinct() %>% 
  st_as_sf(coords = c("row", "col")) %>% 
  group_by(region) %>% 
  summarize(geometry = st_union(geometry)) %>% 
  ungroup() %>% 
  mutate(geometry = st_convex_hull(geometry)) %>% 
  mutate(centroid = st_centroid(geometry))

save(matrix_region, file = here::here("intermediate", "region_shape_object.rda"))

if(DO_PRELIM_MODEL_REGION){
  
  ## Run models for average pointwise differences in each region
  region_models <- cranio_matrix %>% 
    # Rearrange to incorporate subject data
    unnest(data) %>% 
    # Take average in each region for each image
    group_by(fname, age, sex, fusion_type, region) %>% 
    summarize(diff_avg = mean(diff)) %>% 
    ungroup() %>%
    # Add fusion indicators
    left_join(fusion_dict, by = "fusion_type") %>% 
    # Nest region-specific data
    group_by(region) %>% 
    nest() %>% 
    ungroup() %>% 
    # Fit model with age as a spline (cubic regression basis for fast estimation)
    mutate(model = map(data,
                       ~gam(diff_avg ~ 
                              fused_Sagittal + fused_Metopic +
                              fused_RCoronal * fused_LCoronal +
                              sex + s(age, bs = "cr"),
                            data = .x, method = "REML")),
           # Also try linear age
           model_linear = map(data,
                              ~lm(diff_avg ~ 
                                    fused_Sagittal + fused_Metopic +
                                    fused_RCoronal * fused_LCoronal +
                                    sex + age,
                                  data = .x)))
  
  saveRDS(
    region_models, 
    file = here::here("intermediate", 
                      "prelim_models_region.rds"))
  
} else { 
  region_models <- readRDS(
    file = here::here("intermediate", 
                      "prelim_models_region.rds")) }

region_models %<>% 
  ## Add bicoronal variance to final coefficients; output coefficient summary
  mutate(across(matches("model"),
                ~map(.,
                     function(m){
                       
                       # Make constrast vector identifying all coronal variables
                       is_coronal <- as.numeric(str_detect(names(coef(m)), "Coronal"))                        
                       
                       # Compute variance of bicoronal combination
                       bicoronal_v <- crossprod(is_coronal,  vcov(m)) %*% is_coronal
                       bicoronal_e <- crossprod(is_coronal, coef(m))
                       
                       # Add bicoronal as row in tidy summary
                       bicoronal_row <- tibble(
                         term = "fused_Bicoronal",
                         estimate = bicoronal_e[1,1],
                         std.error = sqrt(bicoronal_v[1,1])) %>% 
                         mutate(p.value = 2 * pnorm(abs(estimate / std.error), 
                                                    lower.tail = FALSE))
                       
                       # Output coefficient table (smaller than model)
                       return(tidy(m, parametric = TRUE) %>% 
                                bind_rows(bicoronal_row))
                     }),
                .names = "{.col}_coeffs")) %>% 
  # Rename individual model terms before merging
  mutate(model_coeffs = map(model_coeffs, 
                            ~rename_with(., 
                                         .cols = -term,
                                         .fn = ~paste0(., "_smooth"))),
         model_linear_coeffs = map(model_linear_coeffs, 
                                   ~rename_with(., 
                                                .cols = -term,
                                                .fn = ~paste0(., "_linear")))) %>% 
  # Merge model summaries
  mutate(coeffs_combo = map2(model_coeffs, model_linear_coeffs,
                             ~full_join(.x, .y, by = "term"))) %>% 
  # Clean model summary
  mutate(coeffs_combo_cln = map(coeffs_combo, 
                                ~filter(., term != "fused_RCoronal:fused_LCoronal") %>% 
                                  arrange(-str_detect(term, "fused")) %>% 
                                  mutate(term = fct_inorder(term))))

# Merge model summaries with matrix for plotting
model_region_matrix <- region_models %>% 
  select(region, coeffs_combo_cln) %>% 
  unnest(coeffs_combo_cln) %>% 
  full_join(matrix_region, by = "region")

# Plot parametric coefficients
plot_region_smooth <- ggplot(model_region_matrix %>% filter(term != "age")) +
  geom_sf(aes(geometry = geometry, fill = estimate_smooth), linewidth = 0) +
  geom_sf_text(aes(geometry = centroid, label = region)) +
  facet_wrap(~term) +
  scale_fill_viridis_c(option = "plasma") +
  theme_void() +
  labs(title = "Region-Level Smooth Age Model",
       fill = "Coefficient") +
  theme(legend.position = "bottom")

# Plot parameters from alternate version of the model
plot_region_linear <- ggplot(model_region_matrix) +
  geom_sf(aes(geometry = geometry, fill = estimate_linear), linewidth = 0) +
  geom_sf_text(aes(geometry = centroid, label = region)) +
  facet_wrap(~term) +
  scale_fill_viridis_c(option = "plasma") +
  theme_void() +
  labs(title = "Region-Level Linear Age Model",
       fill = "Coefficient") +
  theme(legend.position = "bottom")

# Print these plots
plot_region_linear

plot_region_smooth

## Plot smoothed effect
# Define range for predictions
newdata_cranio_age <- tibble(
  fused_Sagittal = 0,
  fused_Metopic = 0,
  fused_RCoronal = 0,
  fused_LCoronal = 0,
  sex = 0,
  age = seq(min(cranio_sub$age), 
            max(cranio_sub$age),
            length.out = 30)
)

# Calculate predicted values and Conf Ints
model_region_smooth <- region_models %>% 
  select(region, model) %>% 
  # Add predictions from smooth term
  mutate(smooth_pred = map(
    model, 
    ~obtain_predictions(
      model = .x, newdata = newdata_cranio_age))) %>% 
  select(-model) %>% 
  unnest(smooth_pred)

# Add linear age effect to compare
model_region_linear <- region_models %>% 
  select(region, model_linear) %>% 
  mutate(pred_data = list(newdata_cranio_age),
         lin_pred = map(
    model_linear,
    ~predict(.x, newdata = newdata_cranio_age, se.fit = TRUE) %>% 
      as.data.frame())) %>% 
  select(-model_linear) %>% 
  unnest(c(pred_data, lin_pred)) %>% 
  mutate(ci_ll = fit - 1.96 * se.fit,
         ci_ul = fit + 1.96 * se.fit)

# Combine smooth and linear predictions
model_region_combo <- model_region_smooth %>% 
  select(region, age, fit, ci_ll, ci_ul) %>% 
  mutate(type = "Smooth") %>% 
  bind_rows(
    model_region_linear %>% 
      select(region, age, fit, ci_ll, ci_ul) %>% 
      mutate(type = "Linear")
  )

# Plot 1 smoothed effect per region
plot_region_smooth_pred <- ggplot(model_region_combo) +
  geom_line(aes(x = age, y = fit, 
                color = factor(region), linetype = rev(type))) +
  geom_ribbon(aes(x = age, 
                  ymin = ci_ll, ymax = ci_ul, 
                  fill = factor(region),
                  linetype = type),
              alpha = 0.3) +
  facet_wrap(~type, nrow = 1) +
  scale_linetype(guide = "none") +
  scale_color_discrete(palette = "Set3") +
  labs(x = "Age (Days)", y = "Predicted Growth (No Fusion Female)", 
       color = "Region", fill = "Region") +
  theme_minimal() +
  theme(legend.position = "bottom")

plot_region_smooth_pred

```

## Pixel-Level Independent

To get a more detailed spatial perspective, we then run the same regression model on each individual pixel in the data.

We denote the growth for patient $i$ in matrix row $x$ and column $y$ as $G_{i}(x,y)$ and let $f^{(x,y)}(\cdot)$ be a cubic regression spline function estimated on observations at pixel $(x,y)$; then this model can be written:

$$
\begin{gather*}
\begin{split}
G_{i}^{(x,y)} = {} & \beta_{0}^{(x,y)} + \beta_{1}^{(x,y)} [\text{Sagittal}]_i + \beta_{2}^{(x,y)} [\text{Metopic}]_i + \\
& \beta_{3}^{(x,y)} [\text{Right Coronal}]_i + \beta_{4}^{(x,y)} [\text{Left Coronal}]_i + \\
& \beta_{5}^{(x,y)} [\text{Left and Right Coronal}]_i + \\
& \beta_{6}^{(x,y)} [\text{Sex: Male}]_i + f^{(x,y)}(Age_i) + \epsilon_{i}^{(x,y)}
\end{split} \\ \\
\epsilon_{i}^{(x,y)} \iid{} N(0, \sigma^{2 \ (x,y)})
\end{gather*}
$$

```{r}
#| results: hide

# Clear memory space
rm(list = setdiff(ls(), c(lsf.str(), ls(pattern = eval(RM_IGNORE)))))
gc(verbose = F)
```

```{r}
#| label: pixel-model

if(DO_PRELIM_MODEL_PIXEL){
  
  ## Run models for pointwise differences in each pixel
  pixel_models <- cranio_matrix %>% 
    mutate(
      model = map(data,
                  function(d){
                    d_reg <- d %>% left_join(fusion_dict, by = "fusion_type")
                    
                    mdl <- bam(diff ~ 
                                 fused_Sagittal + fused_Metopic +
                                 fused_RCoronal * fused_LCoronal +
                                 sex + s(age, bs = "cr"),
                               data = d_reg)
                    
                    return(mdl)
                  })) %>% 
    mutate(coeffs = map(
      model,
      function(m){
        
        # Make constrast vector identifying all coronal variables
        is_coronal <- as.numeric(str_detect(names(coef(m)), "Coronal"))                        
        
        # Compute variance of bicoronal combination
        bicoronal_v <- crossprod(is_coronal, vcov(m)) %*% is_coronal
        bicoronal_e <- crossprod(is_coronal, coef(m))
        
        # Add bicoronal as row in tidy summary
        bicoronal_row <- tibble(
          term = "fused_Bicoronal",
          estimate = bicoronal_e[1,1],
          std.error = sqrt(bicoronal_v[1,1])) %>% 
          mutate(statistic = estimate / std.error,
                 p.value = 2 * pnorm(abs(estimate / std.error), 
                                     lower.tail = FALSE))
        
        # Output coefficient table (smaller than model)
        return(tidy(m, parametric = TRUE) %>% 
                 bind_rows(bicoronal_row))
      }))
  
  # Combine pixel-level models and add region column
  model_pixel_summ <- pixel_models %>% 
    select(row, col, coeffs) %>% 
    unnest(coeffs) %>% 
    # Remove raw bicoronal interaction term (replaced with Bicoronal combo)
    filter(term != "fused_RCoronal:fused_LCoronal") %>% 
    arrange(row, col, -str_detect(term, "fused")) %>% 
    mutate(term = fct_inorder(term))
  
  # Extract smother predictions
  model_pixel_smooth <- pixel_models %>% 
    select(row, col, region, model) %>% 
    # Add plot of smoothed term
    mutate(smooth_pred = map(
      model, 
      ~obtain_predictions(
        model = .x, newdata = newdata_cranio_age))) %>% 
    select(-model) %>% 
    unnest(smooth_pred)
  
  save(model_pixel_summ, model_pixel_smooth,
       file = here::here("intermediate", 
                         "prelim_model_pixels_summ.rda"))
} else{ load(file = here::here("intermediate", 
                               "prelim_model_pixels_summ.rda")) }

# Plot pixel-level coefficients
plot_pixel <- ggplot(model_pixel_summ) +
  geom_raster(aes(x = row, y = col, fill = estimate)) +
  facet_wrap(~term) +
  coord_fixed() +
  scale_fill_viridis_c(option = "plasma") +
  theme_void() +
  labs(title = "Pixel-Level Model",
       fill = "Coefficient") +
  theme(legend.position = "bottom")

plot_pixel

# Plot 1 age smoothed effect per region
plot_pixel_pred <- ggplot(model_pixel_smooth) +
  geom_line(aes(x = age, y = fit, 
                group = interaction(row, col),
                color = factor(region)),
            alpha = 0.05, linewidth = 0.05) +
  labs(x = "Age (Days)", y = "Predicted Growth (No Fusion Female)", 
       color = "Region") +
  theme_minimal() +
  scale_color_discrete(palette = "Set3") +
  guides(color = guide_legend(override.aes = list(alpha = 1, linewidth = 0.5))) +
  theme(legend.position = "bottom")

plot_pixel_pred
```

## Fusion-Independent GAM {#sec-prelim-gam}

Next, for a preliminary spatial model, we split the data only by fusion type. Based on the plots above, the age and sex effects are basically uniform across all of the pixels, so we include these as linear terms that do not vary spatially. Otherwise, we only include a tensor spline interaction between row and column indices. Our model does not reflect the brain boundary. It is a generalized additive model with an intercept and a tensor product smooth of row and column indices.

We denote the growth for patient $i$ with suture fusion type $s$ in matrix row $x$ and column $y$ as $G_{si}(x,y)$. Let $f_{s}(\cdot)$ be a cubic regression spline estimated on observations from patients with suture fusion $s$ and let $h_s(\cdot, \cdot)$ be a tensor product smooth with a cubic regression spline basis; then this model can be written:

$$
\begin{gather*}
\begin{split}
G_{si}(x,y) = {} & \beta_{0, s} + \beta_{1,s} [\text{Sex: Male}]_i + \\
& f_{s}(Age_i) + h_s(x, y) + \epsilon_{si}(x,y)
\end{split} \\ \\
\epsilon_{si}(x,y) \iid{} N(0, \sigma^2_s).
\end{gather*}
$$

Predictions from this model are shown below. In order to show how this model is inadequate for our data, we compare these predictions to the boundary of the matrix data.

```{r}
#| results: hide

# Clear memory space
rm(list = setdiff(ls(), c(lsf.str(), ls(pattern = eval(RM_IGNORE)))))
gc(verbose = F)
```

```{r}
#| label: preliminary-gam

if(DO_PRELIM_MODEL_FUSION){
  
  # Run preliminary spatial models
  fusion_models <- cranio_matrix %>% 
    unnest(data) %>% 
    # Split data by fusion type
    group_by(fusion_type) %>% 
    nest() %>% 
    ungroup() %>% 
    # Fit model with linear sex effect, smoothed age term, and row-col tensor surface
    mutate(model = map(data,
                       ~bam(diff ~ sex + s(age, bs = "cr") + 
                              te(row, col),
                            data= .x)))
  
  saveRDS(
    fusion_models, 
    file = here::here("intermediate", "prelim_models_fusion_te.rds"))
} else { 
  fusion_models <- readRDS(
    file = here::here("intermediate", "prelim_models_fusion_te.rds")) }

# Set new data for GAM
newdata_fusiongam <- tibble(
  fused_Sagittal = 0,
  fused_Metopic = 0,
  fused_RCoronal = 0,
  fused_LCoronal = 0,
  sex = 0,
  age = 365,
  row = list(seq(from = (min(cranio_bound_matrix$row) - 30), 
                 to = (max(cranio_bound_matrix$row) + 30), 
                 length.out = 50)),
  col = list(seq(from = (min(cranio_bound_matrix$col) - 30), 
                 to = (max(cranio_bound_matrix$col) + 30), 
                 length.out = 50))
) %>% 
  unnest_longer(row) %>% unnest_longer(col)

fusion_models  %<>% 
  # Predict from this model for spatial pattern (1 year old male)
  mutate(pred_data = map(
    model,
    ~obtain_predictions(.x, newdata = newdata_fusiongam)))

# Flatten predictions into their own data
fusion_model_summ <- fusion_models %>% 
  select(fusion_type, pred_data) %>% 
  unnest(pred_data)

# Plot these predictions (in a rectangle shape with brain mask)
plot_model_fusion <- ggplot() +
  geom_raster(data = fusion_model_summ, 
              aes(x = row, y = col, fill = fit)) +
  geom_sf(data = cranio_bound_shape, fill = NA, linewidth = 1) +
  facet_wrap(~fusion_type, nrow = 2) +
  scale_fill_viridis_c(option = "turbo") +
  scale_alpha(range = c(0.5, 1), guide = "none") +
  theme_void() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(fill = "Pred. Growth\n(1 y.o. F)") +
  theme(legend.position = "bottom")

plot_model_fusion

ggsave(here::here("Presentation Images", "tensor_smooth_fusion.png"),
       plot_model_fusion,
       height = 4, width = 8, units = "in")

# Add row and column to age smooth new-data (to be cancelled out)
newdata_age_gam <- newdata_cranio_age %>% 
  mutate(row = 0, col = 0)

## Extract age smoothed predictions
model_fusion_smooth <- fusion_models %>% 
  select(fusion_type, model) %>% 
  # Add plot of smoothed term
  mutate(smooth_pred = map(
    model, 
    ~obtain_predictions(.x, 
                       newdata = newdata_age_gam,
                       rm_effects = c("row", "col")))) %>% 
  select(-model) %>% 
  unnest(smooth_pred)

# Plot 1 smoothed effect per fusion type
plot_fusion_pred <- ggplot(model_fusion_smooth) +
  geom_line(aes(x = age, y = fit, color = fusion_type)) +
  geom_ribbon(aes(x = age, ymin = ci_ll, ymax = ci_ul, 
                  group = fusion_type, fill = fusion_type),
              alpha = 0.5) +
  labs(x = "Age (Days)", y = "Predicted Growth (Male)", 
       color = "Fusion Type", fill = "Fusion Type") +
  scale_fill_discrete(palette = fusion_color_dict$color) +
  theme_minimal() +
  theme(legend.position = "bottom")

plot_fusion_pred

## Compile parametric coefficients (Intercept and sex)
model_fusion_params <- fusion_models %>% 
  select(fusion_type, model) %>% 
  mutate(params = map(model, ~tidy(., parametric = TRUE))) %>% 
  select(-model) %>% 
  unnest(params)

# Print table of these parametric coefficients
kbl(model_fusion_params, 
    caption = "Preliminary 2D GAM Parametric Coefficients") %>% 
  kable_styling(full_width = F)

```

# Individual Soap Film Models

Next, we introduce the soap film smoother. To fit this model, we first need to create the boundary that determines the extent of the model and the set of knots in $\mathbb{R}^2$ that will define the smoothing functions.

## Soap Film Grid  {#sec-soap-film-grid}

We construct the boundary shape above. Here, we set up a $10 \times 10$ grid of knots and crop this grid to only the knots within the boundary. We visualize these knots to ensure that the process works: every knot included in our model is within the boundary, and every knot not included is outside of it.

```{r}
#| label: soap-infrastructure

## Create boundary as list
bound_list <- list(
  list(row = cranio_bound_matrix$row,
       col = cranio_bound_matrix$col)
)

## Create knots
grid_n <- 10

# Identify row knot points
grid_row <- seq(min(cranio_bound_matrix$row), max(cranio_bound_matrix$row), 
                length.out = grid_n)

# Identify column knot points
grid_col <- seq(min(cranio_bound_matrix$col), max(cranio_bound_matrix$col), 
                length.out = grid_n)

# Expand to grid of knots
grid_all <- expand.grid("row" = grid_row, "col" = grid_col)

# Crop to cranial boundary
so_grid_k <- grid_all %>% 
  st_as_sf(coords = c("row", "col")) %>% 
  st_filter(cranio_bound_shape) %>% 
  st_coordinates() %>% 
  as_tibble() %>% 
  rename(row = X, col = Y)

# Visualize grid
grid_viz <- ggplot() +
  geom_point(data = cranio_bound_matrix, aes(x = row, y = col), color = "blue") +
  geom_point(data = grid_all, aes(x = row, y = col)) +
  geom_point(data = so_grid_k, aes(x = row, y = col), color ="red") +
  coord_fixed() +
  labs(title = "Grid Sanity Check")

grid_viz
```

## Smoother Object {#sec-soap-film-smoothcon}

Next, we use the set of row and column indices (shared among all images) to create a design matrix containing 2-dimensional basis functions that make up a smoother that can be used for every individual image. We use the `mgcv::smoothCon` function.

```{r}
#| label: soap-film-basis

# Keep only what we need to build smoother
so_matrix_pts <- cranio_matrix %>% 
  select(row, col)

save(so_matrix_pts, file = here::here("intermediate", "matrix_indices.rda"))

# Construct soap film smoother object
cranio_soap <- smoothCon(
  s(
    row, col, 
    bs = "so", 
    xt = list(
      bnd = bound_list)
  ),
  data = so_matrix_pts, knots = so_grid_k
)

# Keep the actual smooth list
cranio_soap <- cranio_soap[[1]]
```

From this smoother object, we extract the relevant matrices that will allow us to fit the model. Let $m$ be the number of pixels in each image and $K$ be the number of basis functions evaluated on these pixels. Then we extract the $m \times K$ design matrix $\bB$, the $K_b \times K_b$ penalty matrix for the boundary spline ($\bS_b$) and the $K_f \times K_f$ penalty matrix for the interior spline ($\bS_f$). In `mgcv`, The smoother object stores each penalty matrix as a block-diagonal size $K \times K$ matrix, where $K = K_b + K_f$. We denote these $\bS^{(b)}$ and $\bS^{(f)}$:

$$
\begin{aligned}
\bS^{(b)} &= \begin{bmatrix}
\bS_b & \smash{\overbrace{\bzero}^{K_b \times K_f}} \\
\smash{\underbrace{\bzero}_{K_f \times K_b}} & \smash{\underbrace{\bzero}_{K_f \times K_f}}
\end{bmatrix} \\ \\ \\ \\
\bS^{(f)} &= \begin{bmatrix}
\smash{\overbrace{\bzero}^{K_b \times K_b}} & \smash{\overbrace{\bzero}^{K_b \times K_f}} \\
\smash{\underbrace{\bzero}_{K_f \times K_b}} & \bS_f
\end{bmatrix}
\\ \\
\end{aligned}
$$

In order to isolate the effect of each separate spline, we also remove the block zero matrices to convert each penalty matrix into the original individual matrices $S_b$ and $S_f$.

To estimate coefficients for this model, we multiply these penalty matrices by smoothing parameters $\lambda_b$ and $\lambda_f$. We then use the "total penalty matrix"

$$
\bS^{(\lambda)}_T = \lambda_b \bS^{(b)} + \lambda_f \bS^{(f)} = 
\begin{bmatrix}
\lambda_b \bS_b & \smash{\overbrace{\bzero}^{K_b \times K_f}} \\
\smash{\underbrace{\bzero}_{K_f \times K_b}} & \lambda_f \bS_f
\end{bmatrix}.
\\ \\
$$

```{r}
#| results: hide

# Clear memory space
rm(cranio_bound_shape)
rm(list = setdiff(ls(), c(lsf.str(), ls(pattern = eval(RM_IGNORE)))))
gc(verbose = F)
```

```{r}
#| label: soap-film-matrices

# Define model matrix
so_X <- cranio_soap$X

# Identify sample size
so_n <- nrow(so_X)

# Identify boundary and film penalty matrices
S_bound0 <- cranio_soap$S[[1]]
S_film0 <- cranio_soap$S[[2]]

# Identify different parts of penalty matrix
cols_bound <- which(apply(S_bound0, 2, sum) != 0)
cols_film <- which(apply(S_film0, 2, sum) != 0)

# Get interior "film" design and penalty matrices
so_X_film <- so_X[, cols_film]
S_film <- S_film0[cols_film, cols_film]

save(so_X_film, 
     file = here::here("intermediate", "design_matrix_interior.rData"))
```

### Soap Film Basis Functions {#sec-soap-basis-fns}

After constructing the basis, it is useful to visualize the basis functions to understand the effect each has on predictions.

First, we identify the outermost points in our data, on which the boundary splines are estimated. These points are the "nearest neighbors" to the boundary points defined earlier. We keep the boundary points in the order in which they are defined in the boundary object, so that they can be connected by line segments to form the boundary. For each boundary point, we calculate the Euclidean distance from the previous point and take the cumulative sum of these distances to determine the distance around the boundary. We use this measure as the X axis to plot the basis that is used to estimate the boundary condition, i.e. the value of the outcome on the boundary.

Then, we plot the "boundary-induced" basis functions: these are the 2-dimensional functions that are defined over the entire matrix. Each boundary-induced function essentially forms the minimal surface over the area that respects its single corresponding boundary basis function. This is described in section 4.1 in the soap film paper (p. 940):

First, suppose values on the boundary follow a function $\bG_i^{(b)}(r) = f_i^{(b)}(r) + \epsilon_i^{(b)}(r)$, where $r$ is the distance along the boundary loop from an arbitrary starting point. 

Second, represent that function with the basis expansion $f_i^{(b)}(r) = \sum_{j=1}^{K_b} \gamma^{(b)}_{ij} \delta_j(r)$. To estimate $\bgamma^{(b)}_{i}$, include a penalty functional $\bgamma_{i}^{(b) \top} \bS_b \bgamma^{(b)}_{i}$.

These basis functions $\boldsymbol{\delta}(r)$ are plotted below:

```{r}
#| label: set-up-basis-visualizations

# Link knots to interior basis functions
so_grid_k %<>% mutate(basis_fn = cols_film)

# Identify sample of interior knots to plot (3 groups of 5)
so_plot_knots <- cols_film[c(1:5, 
                             (round(length(cols_film)/2) + c(-2:2)),
                             ((length(cols_film) - 4):length(cols_film)))]

### Link boundary points with nearest matrix point
# Boundary is buffered 1 unit away from matrix; need to match it with real points

# Convert boundary to sf for nearest neighbor matching
# Add in distances determined in smoother object
bound_points <- cranio_bound_matrix %>% 
  mutate(bound_r = cranio_soap$sd$bnd[[1]]$d[-1]) %>% 
  st_as_sf(coords = c("row", "col"), remove = F)

# Convert points within matrix to sf for nearest neighbor
matrix_sf <- so_matrix_pts %>% st_as_sf(coords = c("row", "col"), remove = F)

# Identify nearest matrix neighbor to each boundary point
bound_points %<>%
  mutate(nearest_cell = st_nearest_feature(bound_points, matrix_sf)) %>% 
  st_drop_geometry() %>% 
  select(nearest_cell, bound_r)

# Match matrix points to their order on the boundary
matrix_pts_df <- so_matrix_pts %>%
  mutate(cell_id = 1:n()) %>% 
  left_join(bound_points,
            by = c("cell_id" = "nearest_cell")) %>% 
  # When we get multiple matches, take first point on boundary
  arrange(cell_id, bound_r) %>% 
  group_by(cell_id) %>% 
  slice(1) %>% 
  ungroup() %>% 
  select(-cell_id)

# Create data of basis functions and pixel locations for plotting
basis_df <- cranio_soap$X %>% 
  data.frame() %>% 
  bind_cols(matrix_pts_df) %>% 
  pivot_longer(-c(row, col, bound_r), 
               names_to = "basis_fn") %>% 
  mutate(basis_fn = as.numeric(str_remove(basis_fn, "X")),
         fn_cat = case_when(basis_fn %in% cols_bound ~ "Boundary", 
                            basis_fn %in% cols_film ~ "Interior"))
```

```{r}
#| label: PDE-matrix

# Reconstruct PDE derivative matrix D
soap_D <- t(cranio_soap$sd$P) %*%
  cranio_soap$sd$L %*% cranio_soap$sd$U %*%
  cranio_soap$sd$Q

```


```{r}
#| label: boundary-cyclic-basis-viz

# Plot boundary cyclic 1-dimensional splines
soap_basis_cyclic <- ggplot(data = basis_df %>% 
                              filter(!is.na(bound_r) & fn_cat == "Boundary")) +
  geom_line(aes(x = bound_r, y = value, color = factor(basis_fn))) +
  scale_color_brewer(palette = "Paired") +
  labs(x = "Distance Along Boundary", 
       y = "Function Value", 
       color = "Basis Function",
       title = "Boundary Cyclic Basis Functions") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(nrow = 2, byrow = T))

soap_basis_cyclic
```

Third, for each $j = 1, \ldots, K_b$, determine the basis-specific boundary condition by setting $\gamma^{(b)}_j = 1$ and $\gamma^{(b)}_{j'} = 0, \ j' \neq j$. *I think this means all we're left with is * $\delta_j(r)$. Solve for the function $a_j(x,y)$ which satisfies this boundary condition based on $\delta_j(r)$ and satisfies the equation $\frac{\partial^2 a_j}{\partial x^2} + \frac{\partial^2 a_j}{\partial y^2} = 0$. 

Then the "boundary-condition-induced-function" $\alpha_i^{(b)}(x,y)$ has the basis function representation $\alpha_i^{(b)}(x,y) = \sum_{j=1}^{K_b} \gamma^{(b)}_{ij} a_j(x,y)$. The basis functions $\ba(x,y)$ are plotted below:

```{r}
#| label: boundary-2d-function-viz

# Plot boundary-induced 2-dimensional basis functions
soap_basis_bd <- ggplot() +
  geom_raster(data = basis_df %>% filter(fn_cat == "Boundary"),
              aes(x = row, y = col, fill = value)) +
  scale_fill_viridis_c() +
  labs(x = NULL, y = NULL, fill = "Fn. Value",
       title = "Boundary Basis Functions") +
  coord_fixed() +
  theme_void() +
  theme(legend.position = "bottom") +
  facet_wrap(~basis_fn)

soap_basis_bd
```

Last, we plot a sample of interior basis functions. *I think* each interior basis function is defined in a similar form as the boundary-induced functions:

First, suppose the residual values after accounting for the boundary values follow a function $\bG_i(x,y) - \alpha_i^{(b)}(x,y) = h_i^{(f)}(x,y) + \epsilon^{(f)}(x,y)$.

Second, represent that function with the basis expansion $h_i^{(f)}(x,y) = \sum_{l=1}^{K_f} \gamma^{(f)}_{il} z_l(x,y)$.

Third, for each basis function $z_l(x,y)$, set up a corresponding 2-dimensional function $\rho_{l}(x,y)$ such that $\rho_{l}(x,y) = 0$ on the boundary, $\frac{\partial^2 \rho_l}{\partial x^2} + \frac{\partial^2 \rho_l}{\partial y^2} = 0$, and $\rho_{l}(x,y)$ rises to a singularity at one corresponding knot $(x^*_{l}, y^*_{l})$.

Fourth, calculate the function $z_l(x,y)$ that solves the equation  $\frac{\partial^2 z_l}{\partial x^2} + \frac{\partial^2 z_l}{\partial y^2} = \rho_l(x,y)$.

Fifth, to estimate the basis coefficients $\bgamma_i^{(f)}$, include a penalty functional $\bgamma_i^{(f) \top} \bS_f \bgamma_i^{(f)}$.

Then the interior function has the basis function representation $h_i^{(f)}(x,y) = \sum_{l=1}^{K_f} \gamma_{il}^{(f)} z_l(x,y)$. A sample of the basis functions $\bz(x,y)$ are plotted below:

```{r}
#| label: basis-interior-viz

# Plot interior basis functions with corresponding knots
soap_basis_int <- ggplot() +
  geom_raster(data = basis_df %>% filter(basis_fn %in% so_plot_knots),
              aes(x = row, y = col, fill = value)) +
  geom_point(data = so_grid_k %>% select(-basis_fn), 
             aes(x = row, y = col), 
             color ="black", size = 0.01) +
  geom_point(data = so_grid_k %>% filter(basis_fn %in% so_plot_knots), 
             aes(x = row, y = col), 
             color ="red", size = 0.01) +
  scale_fill_viridis_c() +
  labs(x = NULL, y = NULL, fill = "Fn. Value",
       title = "Interior Basis Functions") +
  coord_fixed() +
  theme_void() +
  theme(legend.position = "bottom") +
  facet_wrap(~basis_fn, ncol = 5)

soap_basis_int
```

Finally, for one soap film smoother, we have the basis representation 
$$
\begin{aligned}
\bG_i(x,y) &= a_i^{(b)}(x,y) + h_i^{(f)}(x,y) + \epsilon_i(x,y) \\
&= \sum_{j=1}^{K_b} \gamma^{(b)}_{ij} a_j(x,y) + \sum_{l=1}^{K_f} \gamma^{(f)}_{il} z_l(x,y) + \epsilon_i(x,y) \\
&= \begin{bmatrix} \ba (x,y) & \bz (x,y) \end{bmatrix} \begin{bmatrix} \bgamma^{(b)}_{i} \\ \bgamma^{(f)}_{i} \end{bmatrix}
+ \epsilon_i(x,y) \\
&= \mathcal{B}(x,y) \bgamma_i \\
\implies \bG_i &= \bB \bgamma_i.
\end{aligned}
$$

Notably, this shows that the basis functions $\bB(x,y)$ can be common across all individuals. In this case, the basis coefficients $\bgamma_i$ are the only thing varying across subjects, and each coefficient has the same interpretation across all subjects.

### Smoother Model Components {#sec-soap-film-components}

Then, we estimate the coefficients for this model for patient $i$ using the closed-form expression

$$
\hatbgamma_i = (\bB^\top \bB + \bS^{(\lambda)}_T)^{-1}\bB^\top \bG_i
$$

We run a grid search to find the optimal smoothing penalty constants $\lambda_b$ and $\lambda_f$ (called "sp" in mgcv). We use the range $10^{-9}, ..., 10^{-2}$. We then select the model that achieves the minimum generalized cross-validation ("GCV") score:
$$
\begin{gather*}
GCV = \frac{m \sum_{x,y} [G_i(x,y) - \hat{f_i}(x,y)]^2}{[m - tr(\bH)]^2} \\
\bH = \bB (\bB^\top \bB + \bS^{(\lambda)}_T)^{-1}\bB^\top \\
\tr(\bH) = \tr((\bB^\top \bB + \bS^{(\lambda)}_T)^{-1}\bB^\top \bB)
\end{gather*}
$$

Last, we reshape our data to include one row for each patient, with a nested data set of row indices, column indices, and pointwise differences.

```{r}
#| results: hide

# Clear memory space
rm(so_grid_k, so_matrix_pts, 
   bound_points, matrix_pts_df, basis_df)
```

```{r}
#| label: soap-film-gcv-constants

# Define grid for penalty term optimization
lambdas_film <- 10^(-7:-1)
lambdas_bound <- lambdas_film

## Calculate other necessary objects that only depend on X and penalty constant:
# 1. "visor" matrix for each penalty constant (hat without the first X)
# 2. Trace of hat matrix (with X cycled to end)

# First, only interior film
so_constants_film <- tibble("sp" = lambdas_film) %>% 
  mutate(visor = map(sp,
                     ~tcrossprod(solve(crossprod(so_X_film) + 
                                         .x * S_film), 
                                 so_X_film))) %>% 
  mutate(hat_trace = map_dbl(visor, ~sum(diag(.x %*% so_X_film))))

# Next, boundary and interior
so_constants <- tibble(
  expand.grid("sp_film" = lambdas_film, "sp_bound" = lambdas_bound)
  ) %>% 
  mutate(visor = map2(sp_film, sp_bound,
                      ~tcrossprod(solve(crossprod(so_X) +
                               (.x * S_bound0 + .y * S_film0)),
                               so_X))) %>% 
  mutate(hat_trace = map_dbl(visor, ~sum(diag(.x %*% so_X)))) %>% 
  mutate(constant_row = 1:n())

## Reshape data to have pointwise differences for each image
cranio_models <- cranio_matrix %>% 
  unnest(data) %>% 
  select(fname, fusion_type, row, col, diff) %>% 
  group_by(fname, fusion_type) %>% 
  nest() %>% 
  ungroup()

# Test with small subset of data
so_test_df <- cranio_models %>% 
  filter(!fname %in% cranio_dup_fnames) %>% 
  group_by(fusion_type) %>% 
  slice(1) %>% 
  ungroup()

# Identify range of outcome values for consistent plotting
so_diff_range <- (
  cranio_models %>% 
    select(data) %>% 
    mutate(max_diff = map_dbl(data, ~max(.$diff))) %>% 
    summarize(max_d = quantile(max_diff, 0.999)) %>% 
    pull(max_d)
) * c(-0.1, 1)
```

```{r}
#| results: hide

# Clear memory space
rm(cranio_matrix, newdata_cranio_age)
rm(list = setdiff(ls(), c(lsf.str(), ls(pattern = eval(RM_IGNORE)))))
gc(verbose = F)
```

## Model Fitting without Boundary

### Data subset nested Loop {#sec-film-loop}

To test the model fit process, we take a subset of the data, only keeping the first appearance of each suture fusion type. We then run a nested for loop to fit the soap film model on each image and for each potential smoothing penalty constant. Plots of the GCV and details for the optimal fits are below.

```{r}
#| label: prelim-film-loop

# Run loop over individuals and potential penalty constants
film_loop_df <- NULL

for(i in 1:nrow(so_test_df)){
  for(j in 1:nrow(so_constants_film)){
    
    # Data to fit model on
    d <- so_test_df$data[[i]]
    
    # File name to identify output
    f <- so_test_df$fname[i]
    
    # Matrix helpful for influence and coefficients
    reg_mx_film <- so_constants_film$visor[[j]]
    
    trace_H_film <- so_constants_film$hat_trace[[j]]
    
    # Outcome vector
    soap_y <- d$diff
    
    # Coefficient estimates
    soap_b_film <- reg_mx_film %*% soap_y
    
    # Fitted values and square residuals
    soap_fit <- so_X_film %*% soap_b_film
    soap_sqresid <- (soap_y - soap_fit)^2
    
    # GCV score
    soap_gcv <- (so_n * sum(soap_sqresid)) / (so_n - trace_H_film)^2
    
    out_dl <- tibble_row(
      "fname" = f,
      "sp" = so_constants_film$sp[[j]],
      "beta" = list(soap_b_film), 
      "fit" = list(soap_fit), 
      "gcv" = soap_gcv)
    
    film_loop_df %<>% bind_rows(out_dl)
  }}

# Add fusion type to this output
film_loop_df %<>% 
  left_join(so_test_df %>% select(-data))
  

# Plot GCV from test models
film_loop_gcv_plot <- ggplot(film_loop_df) +
  geom_line(aes(x = sp, y = gcv)) +
  facet_wrap(~fusion_type, scales = "free_y") +
  scale_x_continuous(transform = "log10")

film_loop_gcv_plot

# Take model with lowest GCV for each patient
film_loop_optim <- film_loop_df %>% 
  arrange(fname, gcv, desc(sp)) %>% 
  group_by(fname) %>% 
  slice(1) %>% 
  ungroup() %>% 
  left_join(so_test_df, by = c("fname", "fusion_type"))

kbl(film_loop_optim %>% 
      select(fname, fusion_type, sp, gcv), 
    caption = "First Subset: Optimal Fit for each Patient") %>% 
  kable_styling(full_width = F)

# Extract data and fitted values for plotting
film_loop_optim_fit <- film_loop_optim %>% 
  select(fname, fusion_type, fit, data) %>% 
  unnest(c(fit, data))

# Plot model with lowest GCV for each patient
plot_loop_film <- ggplot(film_loop_optim_fit) +
  geom_raster(aes(x = row, y = col, fill = fit)) +
  scale_fill_viridis_c(option = "turbo", limits = so_diff_range) +
  labs(x = NULL, y = NULL, fill = "Pred. Growth",
       title = "Interior using Loop") +
  coord_fixed() +
  theme_void() +
  theme(legend.position = "bottom") +
  facet_wrap(~fusion_type, nrow = 1)

# Plot actual observations
plot_so_test_data <- ggplot(so_test_df %>% unnest(data)) +
  geom_raster(aes(x = row, y = col, fill = diff)) +
  scale_fill_viridis_c(option = "turbo", limits = so_diff_range) +
  labs(x = NULL, y = NULL, fill = "Growth",
       title = "True Data") +
  coord_fixed() +
  theme_void() +
  theme(legend.position = "bottom") +
  facet_wrap(~fusion_type, nrow = 1)

# Combine and print
loop_film_both <- ggarrange(plot_loop_film, 
                            plot_so_test_data, 
                            ncol = 1)
loop_film_both

```

### Data subset vectorized {#sec-film-map}

Next, we try to replicate this process by nesting everything within a dataset.

```{r}
#| results: hide

# Clear memory space
rm(list = setdiff(ls(), c(lsf.str(), ls(pattern = eval(RM_IGNORE)))))
gc(verbose = F)
```

```{r}
#| label: prelim-film-map

# Add model fits to data frame
film_map_df <- so_test_df %>%
  # Expand data to one row per patient and penalty constant
  mutate(constants = list(so_constants_film)) %>%
  unnest(constants) %>% 
  # Estimate coefficients
  mutate(coeffs = map2(visor, data,
                       ~.x %*% .y$diff)) %>% 
  # Fitted values
  mutate(fit = map(coeffs, ~ so_X_film %*% .x)) %>% 
  # Square residuals
  mutate(ssr = map2_dbl(data, fit,
                        ~ sum((.x$diff - .y)^2))) %>% 
  # GCV Score
  mutate(gcv = (so_n * ssr) / (so_n - hat_trace)^2)

# Extract best model by GCV
film_map_optim <- film_map_df %>% 
  arrange(fname, gcv, desc(sp)) %>% 
  group_by(fname) %>% 
  slice(1) %>% 
  ungroup()

kbl(film_map_optim %>% 
      select(fname, fusion_type, sp, gcv), 
    caption = "First Subset: Optimal Fit by Map/Mutate") %>% 
  kable_styling(full_width = F)

# Extract data and fitted values for plotting
film_map_optim_fit <- film_map_optim %>% 
  select(fname, fusion_type, fit, data) %>% 
  unnest(c(fit, data))

# Plot model with lowest GCV for each patient
plot_map_nobound <- ggplot(film_map_optim_fit) +
  geom_raster(aes(x = row, y = col, fill = fit)) +
  scale_fill_viridis_c(option = "turbo", limits = so_diff_range) +
  labs(x = NULL, y = NULL, 
       fill = "Pred. Growth",
       title = "Predictions: Interior Only") +
  coord_fixed() +
  theme_void() +
  theme(legend.position = "bottom") +
  facet_wrap(~fusion_type, nrow = 1)

# Combine and print
map_film_both <- ggarrange(plot_map_nobound, 
                           plot_so_test_data, 
                           ncol = 1)
map_film_both

```

## Model Fitting with Boundary

### Data subset nested loop {#sec-soap-loop}

To test the model fit process, we take a subset of the data, only keeping the first appearance of each suture fusion type. We then run a nested for loop to fit the soap film model on each image and for each potential smoothing penalty constant. Plots of the GCV and details for the optimal fits are below.

```{r}
#| results: hide

# Clear memory space
rm(list = ls(pattern = "_film"))
rm(list = setdiff(ls(), c(lsf.str(), ls(pattern = eval(RM_IGNORE)))))
gc(verbose = F)
```

```{r}
#| label: prelim-soap-loop

# Run loop over individuals and potential penalty constants
soap_loop_df <- NULL

for(i in 1:nrow(so_test_df)){
  
  # Data to fit model on
  d <- so_test_df$data[[i]]
  
  # File name to identify output
  f <- so_test_df$fname[i]
  
  # Outcome vector
  soap_y <- d$diff
  
  df_i <- NULL
  
  for(j in 1:nrow(so_constants)){
    
    # Matrix helpful for influence and coefficients
    reg_mx <- so_constants$visor[[j]]
    
    trace_H <- so_constants$hat_trace[j]
    
    # Coefficient estimates
    soap_b <- reg_mx %*% soap_y
    
    # Fitted values and square residuals
    soap_fit <- so_X %*% soap_b
    soap_sqresid <- (soap_y - soap_fit)^2
    
    # GCV score
    soap_gcv <- (so_n * sum(soap_sqresid)) / (so_n - trace_H)^2
    
    # Add to image dataset
    df_row_j <- tibble_row(
      "sp_bound" = so_constants$sp_bound[[j]],
      "sp_film" = so_constants$sp_film[[j]],
      "beta" = list(soap_b), 
      "fit" = list(soap_fit), 
      "gcv" = soap_gcv)
    
    df_i %<>% bind_rows(df_row_j)
  }
  
  # Keep all GCV values
  df_i_gcv <- df_i %>% 
    select(-c(beta, fit)) %>% 
    rename_with(.cols = everything(), ~paste0(., "_all"))
  
  # Sort so best GCV / most smooth is on top
  df_i %<>% 
    arrange(gcv, desc(sp_film), desc(sp_bound))
  
  # Extract best model; add all gcv scores
  out_df <- df_i %>% 
    slice(1) %>% 
    mutate("fname" = f, .before = everything()) %>% 
    mutate(all_gcv = list(df_i_gcv))
  
  # Append to final data
  soap_loop_df %<>% bind_rows(out_df)
}

# Add patient info to output
soap_loop_df %<>% left_join(so_test_df, by = c("fname"))

# Plot GCV from test models
soap_loop_gcv_plot <- ggplot(soap_loop_df %>% 
                               unnest(all_gcv)) +
  geom_raster(aes(x = sp_bound_all, y = sp_film_all, fill = gcv_all)) +
  facet_wrap(~fusion_type, scales = "free_y") +
  scale_x_continuous(transform = "log10") +
  scale_y_continuous(transform = "log10") +
  scale_fill_continuous(transform = "log")

soap_loop_gcv_plot

# Print model with lowest GCV
kbl(soap_loop_df %>% 
      select(fname, fusion_type, sp_bound, sp_film, gcv), 
    caption = "First Subset: Optimal Fit for each Patient",
    scientific = TRUE,
    digits = 8) %>% 
  kable_styling(full_width = F)

# Extract data and fitted values for plotting
soap_loop_fit <- soap_loop_df %>% 
  select(fname, fusion_type, fit, data) %>% 
  unnest(c(fit, data))

# Plot model with lowest GCV for each patient
plot_loop_soap <- ggplot(soap_loop_fit) +
  geom_raster(aes(x = row, y = col, fill = fit)) +
  scale_fill_viridis_c(option = "turbo", limits = so_diff_range) +
  labs(x = NULL, y = NULL, 
       fill = "Pred. Growth",
       title = "Predictions Using Loop") +
  coord_fixed() +
  theme_void() +
  theme(legend.position = "bottom") +
  facet_wrap(~fusion_type, nrow = 1)

# Combine and print
loop_soap_both <- ggarrange(plot_loop_soap, 
                            plot_so_test_data, 
                            ncol = 1)
loop_soap_both

```

### Data subset vectorized {#sec-soap-map}

Next, we replicate this process by nesting everything within a dataset.

```{r}
#| results: hide

# Clear memory space
rm(list = setdiff(ls(), c(lsf.str(), ls(pattern = eval(RM_IGNORE)))))
gc(verbose = F)
```

```{r}
#| label: prelim-soap-map

bound_cols <- which(apply(cranio_soap$S[[1]], 2, sum) != 0)
film_cols <- which(apply(cranio_soap$S[[2]], 2, sum) != 0)

# Add model fits to data frame
soap_map_df <- so_test_df %>%
  # Map over every image
  mutate(
    model = map(
      data,
      function(d){
        
        # Fit estimates for every combination of penalty constants
        fits <- so_constants %>% 
          select(-visor) %>% 
          # Estimate coefficients
          mutate(coeffs = map(constant_row,
                              ~so_constants$visor[[.x]] %*% d$diff)) %>% 
          # Fitted values
          mutate(fit = map(coeffs, ~ so_X %*% .x),
                 fit_bound = map(coeffs, 
                                 ~ so_X[, bound_cols] %*% .x[bound_cols]),
                 fit_int = map(coeffs, 
                               ~ so_X[, film_cols] %*% .x[film_cols])) %>% 
          # Sum of square residuals
          mutate(ssr = map_dbl(fit, ~sum((d$diff - .x)^2))) %>% 
          # GCV Score
          mutate(gcv = (so_n * ssr) / (so_n - hat_trace)^2)
        
        # Keep the GCV scores for all combinations of constants
        fits_gcv <- fits %>% 
          select(sp_bound, sp_film, gcv) %>% 
          rename_with(.cols = everything(), ~paste0(.x, "_all"))
        
        # Keep the optimal model (lowest GCV -> smoothes interior -> smoothest boundary)
        image_model <- fits %>%
          arrange(gcv, desc(sp_film), desc(sp_bound)) %>% 
          slice(1) %>% 
          select(-c(hat_trace, constant_row)) %>% 
          # Append all GCV scores
          mutate(all_scores = list(fits_gcv)) %>% 
          tibble_row()
        
        return(image_model)
      }
    )
  )

# Extract all model components
soap_map_df %<>% 
  unnest_wider(model) %>% 
  unnest(c(coeffs, fit, fit_bound, fit_int))

kbl(soap_map_df %>% 
      select(fname, fusion_type, sp_bound, sp_film, gcv), 
    caption = "First Subset: Optimal Fit by Map/Mutate",
    digits = 8) %>% 
  kable_styling(full_width = F)

# Extract data and fitted values for plotting
so_test_fit <- soap_map_df %>% 
  select(fname, fusion_type, fit, fit_bound, fit_int, data) %>% 
  unnest(c(fit, fit_bound, fit_int, data)) %>% 
  mutate(across(matches("fit"), as.numeric)) %>% 
  # Compare separate components
  mutate(fit_sep = fit_bound + fit_int,
         fit_comp = (fit - fit_sep)^2)

# Full value range


# Plot model with lowest GCV for each patient
plot_map_soap <- ggplot(so_test_fit) +
  geom_raster(aes(x = row, y = col, fill = fit)) +
  scale_fill_viridis_c(option = "turbo", limits = so_diff_range) +
  labs(x = NULL, y = NULL, 
       fill = "Pred. Growth",
       title = "Sample Predictions") +
  coord_fixed() +
  theme_void() +
  theme(legend.position = "bottom") +
  facet_wrap(~fusion_type, nrow = 1)

# Combine and print
map_soap_both <- ggarrange(plot_map_soap, 
                           plot_so_test_data, 
                           ncol = 1)
map_soap_both

### Compare predictions with only bound, only interior, both splines, and truth
plot_map_build <- so_test_fit %>% 
  pivot_longer(c(fit_int, fit_bound, fit, diff)) %>% 
  mutate(name_cln = fct_recode(fct_inorder(name),
                               "Data" = "diff",
                               "Interior" = "fit_int",
                               "Boundary" = "fit_bound",
                               "Soap Film" = "fit"),
         ) %>% 
  ggplot() +
  geom_raster(aes(x = row, y = col, fill = value)) +
  scale_fill_continuous_divergingx(palette = "RdYlBu", rev = TRUE,
                                   l3 = 0, p3 = 2) +
  labs(x = NULL, y = NULL, fill = "Pred. Growth", 
       title = "Predictions") +
  facet_grid(name_cln ~ fusion_type) + 
  coord_fixed() +
  theme_void() +
  theme(legend.position = "bottom")

plot_map_build
```

For this subset, we make predictions using only the interior spline, only the boundary spline, and both, and compare all three to the actual data. This shows an interesting result: when we exclude the boundary spline entirely and estimate coefficients, the resulting predicted values are almost all non-negative. When we estimate the coefficients using both splines and then model their contributions separately, the contribution from the interior is often negative, resulting in predictions near 0. This is because the basis functions in the design matrix are not just defined on the boundary; they are defined over the entire space, based on the boundary condition described by a cyclic boundary spline.

## Fitting Coefficients for All Data {#sec-soap-film-individual}

Once we have set up the soap film model, we estimate coefficients for each individual, $\hatbgamma_i$. These are stored in a nested tibble, which we save as an RDS file.

```{r}
#| results: hide

# Clear memory space
rm(so_test_df)
rm(list = setdiff(ls(), c(lsf.str(), ls(pattern = eval(RM_IGNORE)))))
gc(verbose = F)
```

```{r}
#| label: individual-soap-models

if(DO_INDIV_SOAPS){

# Add to existing models data set
cranio_models %<>%
# Map over every image
  mutate(
    model = map(
      data,
      function(d){
        
        # Fit estimates for every combination of penalty constants
        fits <- so_constants %>% 
          select(-visor) %>% 
          # Estimate coefficients
          mutate(coeffs = map(constant_row,
                              ~so_constants$visor[[.x]] %*% d$diff)) %>% 
          # Fitted values
          mutate(fit = map(coeffs, ~ so_X %*% .x)) %>% 
          # Sum of square residuals
          mutate(ssr = map_dbl(fit, ~sum((d$diff - .x)^2))) %>% 
          # GCV Score
          mutate(gcv = (so_n * ssr) / (so_n - hat_trace)^2)
        
        # Keep the GCV scores for all combinations of constants
        fits_gcv <- fits %>% 
          select(sp_bound, sp_film, gcv) %>% 
          rename_with(.cols = everything(), ~paste0(.x, "_all"))
        
        # Keep the optimal model (lowest GCV -> smoothes interior -> smoothest boundary)
        image_model <- fits %>%
          arrange(gcv, desc(sp_film), desc(sp_bound)) %>% 
          slice(1) %>% 
          select(-c(hat_trace, constant_row)) %>% 
          # Append all GCV scores
          mutate(all_scores = list(fits_gcv)) %>% 
          tibble_row()
        
        return(image_model)
      }
    )
  )
  
  saveRDS(cranio_models, 
          file = here::here("intermediate", "individual_soap_film_df.rds"))
  
} else {
  cranio_models <- readRDS(
    file = here::here("intermediate", "individual_soap_film_df.rds"))
}

```

We create a dataset containing only the demographics and estimated coefficients for each individual, and save this along with the design matrix $\bB$, for predictions from these soap film models based on demographics.

```{r}
#| results: hide

# Clear memory space
rm(so_constants)
rm(list = setdiff(ls(), c(lsf.str(), ls(pattern = eval(RM_IGNORE)))))
gc(verbose = F)
```

```{r}
#| label: soap-individuals-data

# Extract all model components
cranio_models %<>% 
  unnest_wider(model) %>% 
  unnest(c(coeffs, fit))

# Extract coefficient estimates
cranio_models_beta <- cranio_models %>% 
  select(fname, coeffs) %>% 
  mutate(coeffs = map(coeffs, as.vector)) %>% 
  rename(beta = coeffs) %>% 
  left_join(cranio_sub, by = "fname") %>% 
  left_join(fusion_dict, by = "fusion_type") %>% 
  select(-beta, everything(), beta) %>% 
  unnest_wider(beta, names_sep = "_", simplify = T)

# Save coefficients as csv
write_csv(cranio_models_beta, 
          here::here("intermediate", "individual_soap_coeffs.csv"))

# Save estimator requirements as RData
save(cranio_models_beta, so_X,
     file = here::here("intermediate", "individual_soap_film_fits.rda"))
```

# Bayesian Soap Film {#sec-bayes-soap}

Next, we fit a model on one individual using a Gibbs sampler. Recall that we defined the "total penalty matrix" as 
$$
\bS^{(\lambda)}_T =
\begin{bmatrix}
\lambda_b \bS_b & \bzero \\
\bzero & \lambda_f \bS_f
\end{bmatrix}.
$$
Note that neither $S_b$ nor $S_f$ are full-rank, but we can define the generalized inverse of $\bS^{(\lambda)}_T$ as
$$
\bS^{(\lambda) \ -}_T =
\begin{bmatrix}
\frac{1}{\lambda_b} \bS_b^- & \bzero\\
\bzero & \frac{1}{\lambda_f} \bS_f^-
\end{bmatrix}.
$$
This generalized inverse is not positive-definite, but we can use it in an improper prior.

We also use the notation 
$$
\bgamma_i = \begin{pmatrix} \smash{\overbrace{\bgamma^{(b)}_{i}}^{K_b \times 1}} \\ \smash{\underbrace{\bgamma^{(f)}_{i}}_{K_f \times 1}} \end{pmatrix},
\\ \\ \\
$$
where $\bgamma_b$ are the coefficients corresponding to the boundary spline basis and $\bgamma_f$ are the coefficients corresponding to the interior "film" spline basis.

Then we assume the (conjugate) priors:

$$
\begin{aligned}
\bG_i & \sim \mathrm{N}_m ( \bB^\top \bgamma_i, \sigma^2 \bI_m) \\
\bgamma_i & \sim \mathrm{N}_K \Big( \bzero, \sigma^2 \bS^{(\lambda) \ -}_T \Big) \\
\sigma^2 & \sim \mathrm{Inv. Gamma} (a, b) \\
\lambda_{b} & \sim \mathrm{Gamma} (c_b, d_b) \\
\lambda_{f} & \sim \mathrm{Gamma} (c_f, d_f) \\
\lambda_{b} & \perp \lambda_{f}
\end{aligned}
$$

We choose simple hyperparameters: $a = 1, \ b = 1,  \ c_b = c_f = 1, d_b = d_f = 1$. We have not yet checked how informative they are.

These priors lead to the full conditionals:

$$
\begin{aligned} 
\ [\lambda_{b} | \bgamma^{(b)}_{i}, \sigma^2] & \sim \mathrm{Gamma} \Bigg( \frac{K_b}{2} + c_b, \ \frac{1}{2 \sigma^2} \bgamma^{(b) \top}_{i} \bS_b \bgamma^{(b)}_{i} + d_b \Bigg) \\
[\lambda_{f} | \bgamma^{(f)}_{i}, \sigma^2] & \sim \mathrm{Gamma} \Bigg( \frac{K_f}{2} + c_f, \ \frac{1}{2 \sigma^2} \bgamma^{(f) \top}_{i} \bS_f \bgamma^{(f)}_{i} + d_f \Bigg) \\
[\sigma^2 | \bgamma_i, \lambda_{b}, \lambda_{f}] & \sim \mathrm{Inv. Gamma} \Bigg( \frac{m}{2} + \frac{K}{2} + a, \ \frac{1}{2} \Big[ \bG_i^\top \bG_i  - 2 \bgamma_i^\top \bB^\top \bG_i + \bgamma_i^\top \Big( \bB^\top \bB + \bS^{(\lambda)}_T \Big) \bgamma_i \Big] + b \Bigg) \\
[\bgamma_i | \sigma^2, \lambda_{b}, \lambda_{f}] & \sim \mathrm{N}_K \Bigg( \Big( \bB^\top \bB + \bS^{(\lambda)}_T \Big)^{-1} \bB^\top \bG_i, \ \sigma^2 \Big( \bB^\top \bB + \bS^{(\lambda)}_T \Big)^{-1} \Bigg)
\end{aligned}
$$
## Gibbs Penalized Regression function

```{r}
#| label: bayesian-regression-function

### Write Gibbs sampler for Bayesian linear regression

# Notes:
# - Assumes that each penalty matrix is the full K x K but is block diagonal
# - Assumes at least one set of penalized terms
# - Assumes one distinct smoothing parameter lambda for each S penalty matrix
# - If there are unpenalized terms, assumes one matrix is unpenalized
# - Assumes that the one unpenalized matrix is the first in the list

lm_penalized_gibbs <- function(
    y, X, # Data
    S, has_unpenalized = FALSE, # Penalty matrices
    sigmasq0, lambda0, beta0, # Default parameters
    a_pri, b_pri, c_pri, d_pri, # Prior hyperparameters
    iters = 1e4, burn_pct = 0.1 # MCMC parameters
){
  
  # Set up time tracking
  time_check <- Sys.time()
  time_pts <- c("start" = time_check)
  timer <- c("setup" = 0,
             "lambda_s" = 0,
             "beta_update_c" = 0,
             "x_inverse" = 0,
             "beta_moments" = 0,
             "beta_draw" = 0,
             "sigma2_update" = 0,
             "lambdas_update" = 0)
  
  ### Reformat penalty matrix
  if(is.matrix(S)){ S <- list(S) }
  
  ### Calculate dimensions
  
  ## Design matrix
  K <- ncol(X)
  N <- nrow(X)
  
  stopifnot(length(y) == N)
  
  ## Iterations of burn-in
  iters_burn <- ceiling(iters * burn_pct)
  
  ## Penalty dimensions
  
  # number of penalties
  S_blocks <- length(S)
  num_penals <- S_blocks - as.numeric(has_unpenalized)

  # non-zero columns in each penalty matrix
  S_block_cols <- map(S, ~which(apply(., 2, sum) != 0))
  
  # number of parameters for each penalty
  S_block_Ks <- map_dbl(S_block_cols, length)
  
  # First number of parameters might not impact any lambda; if so, remove it
  if(has_unpenalized){ S_block_Ks <- S_block_Ks[-1] }
  
  ### Initialize hyperparameters
  
  # Assign initial values
  sigmasq <- sigmasq0
  lambda <- rep_len(lambda0, num_penals)
  beta <- rep_len(beta0, K)
  
  # Initialize beta moments before update
  beta_expect <- beta
  beta_var <- diag(K)
  xtx_plus_pen <- diag(K)
  
  # Compute hyperparameters that don't update (gamma / Igamma shapes)
  a_post <- N/2 + K/2 + a_pri
  c_post <- S_block_Ks/2 + c_pri
  
  ### Create objects to save output
  
  # Posterior samples
  beta_out <- array(0, dim = c(iters - iters_burn, K))
  sigmasq_out <- array(0, dim = c(iters - iters_burn, 1))
  lambda_out <- array(0, dim = c(iters - iters_burn, num_penals))
  
  # Burn-in period
  beta_burn <- array(0, dim = c(iters_burn, K))
  sigmasq_burn <- array(0, dim = c(iters_burn, 1))
  lambda_burn <- array(0, dim = c(iters_burn, num_penals))
  
  timer["setup"] <- timer["setup"] + 
    difftime(Sys.time(), time_check, units = "secs")
  time_check <- Sys.time()
  
  ### Run MCMC loop
  
  for(iter in 1:iters){
    
    # Output on checkpoints
    if(iter %% iters_burn == 0) cat("at turn", iter, "\n")
    
    ## (1) Update beta (coefficients)
    
    # Full penalty matrix including lambdas
    lambda_S_list <- if(has_unpenalized){
      # If there are unpenalized terms, do not multiply 1st matrix by any lambda
      c(
        S[1],
        map2(lambda, S[-1],
             function(l_const, s_mx){
               l_const * s_mx
             })
        )
    } else {
      # If no unpenalized terms, scale by one lambda for each penalty matrix
      map2(lambda, S,
           function(l_const, s_mx){
             l_const * s_mx
           })
    }
    
    # Sum together these individual (lambda * S) block diagonal matrices
    lambda_S <- Reduce("+", lambda_S_list)
    
    timer["lambda_s"] <- timer["lambda_s"] + 
      difftime(Sys.time(), time_check, units = "secs")
    time_check <- Sys.time()
    
    # Call to Cpp function for beta update
    penalized_beta_inplace(
      y = y, X = X, lambda_S = lambda_S, sigmasq = sigmasq,
      beta_expect, 
      beta_var, 
      xtx_plus_pen
    )
    
    timer["beta_update_c"] <- timer["beta_update_c"] + 
      difftime(Sys.time(), time_check, units = "secs")
    time_check <- Sys.time()
    
    # Generate new beta sample
    beta <- t(rmvnorm(n = 1, beta_expect, beta_var))
    
    timer["beta_draw"] <- timer["beta_draw"] + 
      difftime(Sys.time(), time_check, units = "secs")
    time_check <- Sys.time()
    
    ## (2) Update sigma^2 (residual variance)
    
    # 2nd parameter of inverse gamma
    b_post <- 
      b_pri - crossprod(y, X) %*% beta + 
      0.5 * (sum(y^2) + crossprod(beta, xtx_plus_pen) %*% beta)
    
    # Generate new sigma squared sample
    precision <- rgamma(1, a_post, b_post)
    sigmasq <- 1 / precision
    
    timer["sigma2_update"] <- timer["sigma2_update"] + 
      difftime(Sys.time(), time_check, units = "secs")
    time_check <- Sys.time()
    
    ## (3) Update lambda (beta function smoothness factor)
    
    # Calculate posterior gamma shape
    d_post <- map(
      # Exclude any unpenalized terms from posterior
      if(has_unpenalized){ S[-1] } else { S },
      function(S_j){
        (1 / (2 * sigmasq)) * crossprod(beta, S_j) %*% beta + d_pri
      })
    
    # Update lambdas
    lambda <- map2_dbl(c_post, d_post,
                       function(c_new, d_new){
                         rgamma(1, c_new, d_new)
                       })
    
    timer["lambdas_update"] <- timer["lambdas_update"] + 
      difftime(Sys.time(), time_check, units = "mins")
    time_check <- Sys.time()
    
    ## Save results
    iter_post <- iter - iters_burn
    
    if(iter_post > 0){
      beta_out[iter_post,] <- beta
      sigmasq_out[iter_post,] <- sigmasq
      lambda_out[iter_post,] <- lambda
    } else{
      beta_burn[iter,] <- beta
      sigmasq_burn[iter,] <- sigmasq
      lambda_burn[iter,] <- lambda
    }
  }
  
  time_pts <- c(time_pts, "end" = Sys.time())
  timer["total"] <- difftime(time_pts["end"], time_pts["start"], units = "secs")
  timer <- enframe(timer, name = "step", value = "secs")
  
  return(list("beta" = beta_out,
              "sigma_sq" = sigmasq_out,
              "lambda" = lambda_out,
              "burn_beta" = beta_burn,
              "burn_sigma_sq" = sigmasq_burn,
              "burn_lambda" = lambda_burn,
              "start-stop" = time_pts,
              "timing" = timer))
}
```

## Fit on individual model {#sec-bayes-individual}

```{r}
#| label: bayes-soap-film-sample
#| results: hide

### Run Gibbs sampler for soap film

# Set data to train on, especially outcome
data_eg <- so_test_fit %>% 
  filter(fusion_type == "Sagittal") %>% 
  filter(fname == first(fname))

y_mx <- data_eg$diff

# Set penalty matrices
S_mx <- cranio_soap$S

# Set default parameters
def_pars <- c(
  "beta" = 0.0,
  "sigma_sq" = var(y_mx),
  "lambda" = 0.1,
  "a" = 1.0,
  "b" = 1.0,
  "c" = 1.0,
  "d" = 1.0
)

# Set iterations
bayes_soap_iters <- 1e3

set.seed(70)

# Run function
soap_bayes1 <- lm_penalized_gibbs(
  y = y_mx, X = so_X, S = S_mx,
  beta0 = def_pars["beta"], 
  sigmasq0 = def_pars["sigma_sq"], 
  lambda0 = def_pars["lambda"], 
  a_pri = def_pars["a"],
  b_pri = def_pars["b"],
  c_pri = def_pars["c"],
  d_pri = def_pars["d"],
  iters = bayes_soap_iters, 
  burn_pct = 0.1
)
```

### MCMC Diagnostics {#sec-mcmc-diagnostics}

Trace plots show that the parameters converge to a stable distribution quickly and are well-differentiated.

```{r}
#| label: MCMC-diagnostics

# Check timing
mcmc_time <- soap_bayes1$timing %>% 
  mutate(pct = as.numeric(secs) / last(secs))

kbl(mcmc_time %>% mutate(pct = scales::label_percent(accuracy = 0.1)(pct)), 
    caption = "Loop Timing") %>% 
  kable_styling(full_width = F)

# Make accessible all parameters together

soap_bayes1_tr <- imap(
  soap_bayes1 %>% 
    list_modify("timing" = zap(), 
                "start-stop" = zap()),
  function(x, xn){
    x <- as_tibble(x)
    if(ncol(x) == 1){
      names(x) <- xn
    } else {
      names(x) <- paste0(xn, ".", 1:ncol(x))
    }
    x %<>% 
      mutate(iter = 1:n(), .before = everything()) %>% 
      pivot_longer(-iter, 
               names_pattern = "(burn_)?(.*)",
               names_to = c("draw", "name"))
    
    x
  }
) %>% 
  unname() %>% 
  list_rbind() %>% 
  mutate(draw_cat = if_else(draw == "burn_", 
                            "Burn-In", "Posterior"))

# Add default parameters to this table
trace_default <- soap_bayes1_tr %>% 
  filter(draw_cat == "Burn-In") %>% 
  distinct(draw_cat, name) %>% 
  filter(!str_detect(name, "beta")) %>% 
  mutate(
    iter = 0,
    value = case_when(
      str_detect(name, "sigma_sq") ~ def_pars["sigma_sq"],
      str_detect(name, "lambda") ~ def_pars["lambda"],
      T ~ NA
    ))

soap_bayes1_tr %<>% bind_rows(trace_default)

# Trace plot of sigma
sigma_trace <- ggplot(soap_bayes1_tr %>% 
                        filter(str_detect(name, "sigma_sq"))) +
  geom_line(aes(x = iter, y = value)) +
  facet_wrap(~ name + draw_cat, ncol = 2, scales = "free", 
             labeller = \(x) label_value(x, multi_line = F))

sigma_trace

# Trace plot of lambdas
lambda_trace <- ggplot(soap_bayes1_tr %>% 
                         filter(str_detect(name, "lambda"))) +
  geom_line(aes(x = iter, y = value), alpha = 0.5) +
  facet_wrap(~ name + draw_cat, ncol = 2, scales = "free", 
             labeller = \(x) label_value(x, multi_line = F)) +
  theme(legend.position = "bottom")

lambda_trace
```

```{r}
#| label: soap-beta-trace
#| fig.width: 10
#| fig.height: 50

# Trace plot of betas
beta_trace <- ggplot(soap_bayes1_tr %>% 
                         filter(str_detect(name, "beta"))) +
  geom_line(aes(x = iter, y = value), alpha = 0.5) +
  theme(legend.position = "bottom") +
  facet_wrap(~ name + draw_cat, ncol = 2, scales = "free", 
             labeller = \(x) label_value(x, multi_line = F))

beta_trace
```


### Visualize predictions {#sec-bayes-soap-plot}

```{r}
#| label: bayes-soap-predictions

# Extract posterior mean coefficients
soap_bayes1_betas <- colMeans(soap_bayes1$beta)

# Calculate predictions from these coefficients
soap_bayes1_preds <- so_X %*% soap_bayes1_betas

# Add predictions to data
data_eg %<>% mutate(fit_bayes = as.numeric(soap_bayes1_preds))

data_toplot <- data_eg %>% 
  select(row, col, diff, fit_gcv = fit, fit_bayes) %>% 
  pivot_longer(-c(row, col))

# Plot truth and predictions from Bayesian example
soap_bayes1_plot <- ggplot(data_toplot) +
  geom_raster(aes(x = row, y = col, fill = value)) +
  scale_fill_viridis_c(option = "turbo", limits = so_diff_range) +
  facet_wrap(~name, ncol = 1) +
  labs(x = NULL, y = NULL, 
       fill = "Growth",
       title = "Bayesian Prediction Comparison") +
  coord_fixed() +
  theme_void() +
  theme(legend.position = "bottom")

soap_bayes1_plot
```

# Basis Coefficient Individual Models {#sec-basis-coeff-models}

Once we have a working Bayesian soap film model for one individual, we switch to the other level of the hierarchy: for each basis coefficient, we estimate how that basis coefficient can be estimated as a function of a patient's age and suture fusion type.

Specifically, we assume that each basis coefficient $\gamma_k$ for individual $i$, $i = 1, \ldots, n, \ k = 1, \ldots, K$, can be estimated as the sum of a sex effect added flexible function of age categorized by suture type:

$$
\begin{aligned}
E[\bgamma_{ik}] = {} & \beta_{0k} + \beta_{1k} [\text{Sex}_i] + \\
& f_{0k} ( \text{Age}_i, \bbeta_k^{(0)}, \lambda_{k}^{(0)} ) + \\ 
& f_{1k} ( \text{Age}_i, \bbeta_k^{(S)}, \lambda_{k}^{(S)} ) * \bone [\text{Sagittal}] + \\
& f_{2k} ( \text{Age}_i, \bbeta_k^{(M)}, \lambda_{k}^{(M)} ) * \bone [\text{Metopic}] + \\
& f_{3k} ( \text{Age}_i, \bbeta_k^{(RC)}, \lambda_{k}^{(RC)} ) * \bone [\text{Right Coronal}] + \\
& f_{4k} ( \text{Age}_i, \bbeta_k^{(LC)}, \lambda_{k}^{(LC)} ) * \bone [\text{Left Coronal}] + \\
& f_{5k} ( \text{Age}_i, \bbeta_k^{(BC)}, \lambda_{k}^{(BC)} ) * \bone [\text{Bicoronal}]\\
\end{aligned}
$$

## Basis Coefficient Individual Plots {#sec-basis-coeff-plots}

We can check these assumptions based on the coefficients we estimated for each individual using GCV.

The assumptions appear to be reasonable, although there may be significant variance $\eta_{ik}$. This could be especially problematic for larger ages, where there are significantly fewer observations and are often no observations for less common suture fusion types.

```{r}
#| label: set-up-coeff-data

## Load all coefficients previously fit, along with demographics
gcv_coeffs <- read_csv(here::here("intermediate", "individual_soap_coeffs.csv"))

cranio_coeff_df <- gcv_coeffs %>% 
  mutate(fusion_type = fct_infreq(fusion_type),
         sex = factor(sex, levels = c(0,1)))

```

```{r}
#| label: plot-coeff-demographics
#| fig.width: 10
#| fig.height: 15

coeff_toplot <- cranio_coeff_df %>% 
  pivot_longer(cols = matches("beta"),
               names_to = "beta_id",
               values_to = "beta_val") %>% 
  mutate(beta_num = as.numeric(str_remove_all(beta_id, "beta_")))

coeff_plots <- ggplot(coeff_toplot %>% 
                        filter(beta_num %in% so_plot_knots),
                      aes(x = age, y = beta_val, color = fusion_type)) +
  geom_point(alpha = 0.2) +
  geom_smooth(alpha = 0.5) +
  geom_rug(inherit.aes = F, aes(x = age), color = "gray50") +
  scale_color_discrete(palette = fusion_color_dict$color) +
  facet_grid(beta_id ~ fusion_type, scales = "free_y") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "Age (Days)", y = "Soap Basis Coefficient")

coeff_plots
```

## Coefficient Model Design Matrix {#sec-coefficient-design}

First, we set up the design matrix and penalty matrices, which are the same for each basis coefficient. We currently use 10 knots for each suture fusion-specific smooth. Therefore, the design matrix currently includes 2 parametric covariates (1 intercept and 1 Sex indicator), 9 basis functions for a reference smooth, and $9 * 5 = 45$ basis functions for the smooths associated with Sagittal, Metopic, Left Coronal, Right Coronal, and Bicoronal. In total, the design matrix $\bX$ is $n \times P$, where $P = 56$.

Note that these smooths account for an identifiability constraint. Since we are including multiple basis expansions, each covariate in the smooths needs to sum to zero. In practice, this leads us to subtract the column mean from each column of the design matrix, which reduces the rank of each smooth's columns by 1. Therefore, though we use 10 knots for each smooth, we only keep 9 columns.

We set up the design matrix and the penalty matrices to match the Gibbs sampler function we have already written to fit a penalized linear regression model, which in turn matches the penalty matrices in `mgcv`. Therefore, each penalty matrix is $P \times P$ and block-diagonal, with one $9 \times 9$ penalty matrix included and the remainder of the matrix left as 0.

First, we construct a full "reference" matrix and penalty matrix using `mgcv::smooth.construct2`.

```{r}
#| label: create-coefficient-smooth

### Create cubic regression basis splines for age

# Set rank
knots_per_fusion <- 10

# Construct smooth without constraint or factor levels
cranio_coeff_smooth <- smooth.construct2(
  s(age, bs = "cr", k = knots_per_fusion),
  data = cranio_coeff_df, knots = NULL
)

# Attach name of the dataset to this smooth
attr(cranio_coeff_smooth, "data") <- "cranio_coeff_df"
```

Second, we write a function that takes this smooth object, data, the name of the factor variable we want to construct separate splines for, and the names of the unpenalized "parametric" covariates, and constructs a full design matrix and penalty matrix list for these.

```{r}
#| label: coefficient-design-matrix

### Write functions to set up design matrix 
##  - Given a constructed smooth (with constraint applied), factor variable, and new data:
## .. and generate full design matrix for new data,
## .. assuming "reference coding" with reference spline and difference functions

## 1) Given smooth, subtract column means and drop last column
apply_sweep_and_drop <- function(sm, drop_col = NULL, newdata = NULL){

  # Extract design matrix
  sm_X0 <- sm$X

  # Calculate column means
  sm_means <- colMeans(sm_X0)

  # Identify number of columns
  sm_k <- length(sm_means)
  
  # Determine whether column to drop is provided
  if(is.null(drop_col)) drop_col <- sm_k
  
  # Determine whether to sweep and drop from original or new matrix
  if(is.null(newdata)){
    nd_X0 <- sm_X0
  } else {
    nd_X0 <- PredictMat(sm, data = newdata)
  }

  # Subtract these from design matrix AND remove last column
  X_ref <- sweep(nd_X0[, -drop_col], 2, STATS = as.array(sm_means[-drop_col]))

  # Locate full penalty matrix
  sm_S0 <- sm$S[[1]]

  # Filter to subset corresponding to remaining design matrix
  S_ref <- sm_S0[-drop_col, -drop_col]

  # Export
  out <- list("X" = X_ref,
              "S" = S_ref,
              "means" = sm_means,
              "col_dropped" = drop_col)

  out
}

## 2) Given constructed smooth and a factor, make full design and penalty
##    - assume "reference" coding for splines
make_reference_coded_splines <- function(sm, drop_col = NULL, newdata = NULL, 
                                         by_var){
  
  # Apply constraint to smooth object
  constrained_list <- apply_sweep_and_drop(sm = sm, 
                                           drop_col = drop_col, 
                                           newdata = newdata)
  
  # Get data with factor variable: either from the smooth or a new data frame
  if(is.null(newdata)){
    nd <- get(attr(sm, "data"))
  } else {
    nd <- newdata
  }
  
  # Define vector of factor assignments in data
  by_vec <- nd[[by_var]]
  
  # Define list of effect values for which we need separate splines
  by_effects <- levels(by_vec)[-1]
  
  # Generate one design matrix for each effect
  X_by <- map(by_effects, 
              ~ as.numeric(by_vec == .x) * constrained_list$X)
  
  # Merge with original spline basis
  X_splines <- Reduce(cbind, x = X_by, init = constrained_list$X)
  
  out <- list("X" = X_splines)
  
  # Construct penalty matrix for all splines together
  if(is.null(newdata)){
    # Define indicator matrix list 
    # Each matrix is 0 except for a 1 on the diagonal
    # This 1 represents the block where the penalty matrix S is
    spline_indicator_list <- map(1:length(levels(by_vec)),
                                 .f = function(j, z = rep(0, length(levels(by_vec)))){
                                   z[j] <- 1
                                   diag(z, nrow = length(z))
                                 })
    
    # For each matrix in the list, take kronecker product with S from the smooth
    S_splines <- map(spline_indicator_list,
                     ~kronecker(.x, constrained_list$S))
    
    out[["S"]] <- S_splines
  }
  out
}

## 3) Add parametric covariates
setup_reference_model <- function(sm, drop_col = NULL, by_var, newdata = NULL, 
                                  param_formula = NULL){
  
  # Make spline matrix
  ref_splines <- make_reference_coded_splines(sm = sm, drop_col = drop_col, 
                                              by_var = by_var, newdata = newdata)
  
  # If there is no parametric formula, add unpenalized intercept
  if(is.null(param_formula)){
    design_param <- matrix(1, nrow = nrow(ref_splines$X))
    
    # Include 1 x 1 matrix for unpenalized intercept
    S_unpenal <- matrix(1, 1, 1)
  } else{
    # Otherwise, create matrix columns from formula
    
    # Get data with parametric variables: if no new data, get it from smooth
    if(is.null(newdata)){ 
      nd <- get(attr(sm, "data")) 
    } else {
      nd <- newdata
    }
    
    # Create parametric matrix from this new data
    design_param <- model.matrix(param_formula, nd)
    colnames(design_param) <- NULL
    
    # Create identity matrix for these unpenalized parameters
    S_unpenal <- diag(nrow = ncol(design_param))
  }
  
  # Merge model matrix with spline matrix for full design matrix
  X_design <- cbind(design_param, ref_splines$X)
  
  out <- list("X" = X_design)
  
  if(is.null(newdata)){
  ## Create full penalty / precision matrix
  
  # Add 0's to unpenalized matrix to match size of penalties
  S_unpenal_full <- as.matrix(bdiag(S_unpenal,
                                    diag(0, nrow = nrow(ref_splines$S[[1]]))))
  
  # Add 0's for penalized matrices at beginning of other matrices
  S_full_size <- map(ref_splines$S, 
                     ~as.matrix(
                       bdiag(
                         diag(0, nrow = nrow(S_unpenal)), 
                         .x)))
  
  # Append unpenalized chunk to beginning of penalty matrix list
  S_full <- c(list(S_unpenal_full), S_full_size)
  
  out[["S"]] <- S_full 
  }
  
  out
}
```

```{r}
#| results: hide

# Clear memory space
rm(list = setdiff(ls(), c(lsf.str(), ls(pattern = eval(RM_IGNORE)))))
gc(verbose = F)
```

## Single Basis Coefficient Bayesian Additive Model {#sec-coeff-gibbs}

We use this function to construct the design matrix and penalty matrix list for the Gibbs sampler function. 

To test this model, we use the 38th soap film basis coefficient. This coefficient's basis function is plotted above, and it is in a region that visibly differs between Normative, Sagittal, and Metopic presentations.


```{r}
#| label: fit-basis-coefficient-estimators
#| results: hide

### Run Gibbs sampler for coefficient models

## Create design and penalty matrices using function above
coeff_ingredients <- setup_reference_model(
  sm = cranio_coeff_smooth,
  by_var = "fusion_type", param_formula = ~sex
)

# Combine suture fusion penalties so we estimate the same lambda
coeff_S_combo <- list(coeff_ingredients$S[[1]],
                      Reduce("+", coeff_ingredients$S[-1]))

# Set 1 covariate as outcome 
# - (plotted above, should be lower for sagittal and higher for metopic)
coeff_y <- cranio_coeff_df$beta_38

# Set default parameters
def_pars_coeff <- c(
  "beta" = 0.0,
  "sigma_sq" = var(coeff_y),
  "lambda" = 0.1,
  "a" = 1.0,
  "b" = 1.0,
  "c" = 1.0,
  "d" = 1.0
)

# Set iterations
bayes_coeff_iters <- 5e3

set.seed(700)

# Run function
coeff_bayes1 <- lm_penalized_gibbs(
  y = coeff_y, X = coeff_ingredients$X, 
  S = coeff_S_combo, has_unpenalized = TRUE,
  beta0 = def_pars_coeff["beta"], 
  sigmasq0 = def_pars_coeff["sigma_sq"], 
  lambda0 = def_pars_coeff["lambda"], 
  a_pri = def_pars_coeff["a"],
  b_pri = def_pars_coeff["b"],
  c_pri = def_pars_coeff["c"],
  d_pri = def_pars_coeff["d"],
  iters = bayes_coeff_iters, 
  burn_pct = 0.1
)
```

## Basis Coefficient MCMC Diagnostics {#sec-coeff-mcmc-diagnostics}

We plot the diagnostics for the parameters from this Gibbs sampler.

In particular, we can see that the variance of some of the coefficients is very large. These coefficients correspond to large ages for rarer suture fusions like Left Coronal, Right Coronal, and Bicoronal. The variance highlights a potential issue with sample size. For these rare suture fusions, we do not have many observations at higher ages. This is not a problem for estimating Normative growth or even Sagittal or Metopic growth. However, if we want the same basis expansion for each suture fusion type, we need to leave the knots in the same place, which takes away from our ability to estimate more detailed patterns at younger ages. On the other hand, we could add more knots, but this will add to the computation of estimating this model.

```{r}
#| label: coeff-MCMC-diagnostics

# Check timing
mcmc_coeff_time <- coeff_bayes1$timing %>% 
  mutate(pct = as.numeric(secs) / last(secs))

kbl(mcmc_coeff_time %>% mutate(pct = scales::label_percent(accuracy = 0.1)(pct)), 
    caption = "Loop Timing") %>% 
  kable_styling(full_width = F)

# Make accessible all parameters together

coeff_bayes1_tr <- imap(
  coeff_bayes1 %>% 
    list_modify("timing" = zap(), 
                "start-stop" = zap()),
  function(x, xn){
    x <- as_tibble(x)
    if(ncol(x) == 1){
      names(x) <- xn
    } else {
      names(x) <- paste0(xn, ".", 1:ncol(x))
    }
    x %<>% 
      mutate(iter = 1:n(), .before = everything()) %>% 
      pivot_longer(-iter, 
               names_pattern = "(burn_)?(.*)",
               names_to = c("draw", "name"))
    
    x
  }
) %>% 
  unname() %>% 
  list_rbind() %>% 
  mutate(draw_cat = if_else(draw == "burn_", 
                            "Burn-In", "Posterior"),
         param_num = as.numeric(str_extract(name, "[0-9]+"))) %>% 
  mutate(param_num = coalesce(param_num, 1))

# Add default parameters to this table
trace_coeff_default <- coeff_bayes1_tr %>% 
  filter(draw_cat == "Burn-In") %>% 
  distinct(draw_cat, name, param_num) %>% 
  mutate(
    iter = 0,
    value = case_when(
      str_detect(name, "sigma_sq") ~ def_pars_coeff["sigma_sq"],
      str_detect(name, "lambda") ~ def_pars_coeff["lambda"],
      str_detect(name, "beta") ~ def_pars_coeff["beta"],
      T ~ NA
    ))

coeff_bayes1_tr %<>% bind_rows(trace_coeff_default)

# Trace plot of sigma
sigma_trace_coeff <- ggplot(coeff_bayes1_tr %>% 
                        filter(str_detect(name, "sigma_sq"))) +
  geom_line(aes(x = iter, y = value)) +
  facet_grid(rows = name ~ draw_cat, scales = "free") +
  labs(title = "Sigma Squared Trace")

sigma_trace_coeff

# Trace plot of lambdas
lambda_trace_coeff <- ggplot(coeff_bayes1_tr %>% 
                         filter(str_detect(name, "lambda"))) +
  geom_line(aes(x = iter, y = value), alpha = 0.5) +
  facet_grid(param_num ~ draw_cat, scales = "free") +
  labs(title = "Lambda Trace")

lambda_trace_coeff
```

```{r}
#| label: coeff-MCMC-betas
#| fig.width: 10
#| fig.height: 20

# Trace plot of betas
beta_trace_coeff <- ggplot(coeff_bayes1_tr %>% 
                             filter(str_detect(name, "beta"))) +
  geom_line(aes(x = iter, y = value), alpha = 0.5) +
  facet_grid(param_num ~ draw_cat, scales = "free") +
  labs(title = "Beta Trace")

beta_trace_coeff
```

### Visualize coefficient predictions {#sec-coeff-pred-plot}

We plot the coefficient functions of age by suture (including the estimated intercept).

For each suture fusion type, we construct an evenly spaced 100-length sequence from the minimum age to the maximum age. We use the previously defined function to convert this to a design matrix that fits with our model.

To estimate a credible interval, we generate predictions for every draw of betas from the Gibbs sampler, and then take the 5th and 95th percentiles of these predictions for each row of our constructed data.

These credible intervals are very large for rare suture fusion types for high ages, which is consistent with the large variances of coefficients shown in the trace plots above. To get a better idea of the patterns estimated by the coefficient functions for smaller ages, we reduce the size of the y axis. We also include the locations of the knots in our model to gain a better understanding of what level of detail we currently have from our GAM.

```{r}
#| label: bayes-coefficient-GAM-predictions

### Illustrate coefficient functions from these splines

## Create new data

# Create data shell
coeff_newdata <- cranio_coeff_df %>% 
  mutate(min_age= min(age),
         max_age = max(age)) %>% 
  distinct(fusion_type, min_age, max_age) %>% 
  rowwise() %>% 
  mutate(age = list(seq(min_age, max_age, length.out = 50))) %>% 
  ungroup() %>% 
  mutate(sex = list(c(0, 1))) %>% 
  unnest(age) %>% unnest(sex) %>% 
  mutate(sex = factor(sex, levels = levels(cranio_coeff_df$sex)))

# Run function to convert this to design matrix
coeff_new_design <- setup_reference_model(
  sm = cranio_coeff_smooth, newdata = coeff_newdata,
  by_var = "fusion_type", param_formula = ~sex
)$X

## Calculate predictions
coeff_bayes1_preds <- tcrossprod(coeff_new_design, coeff_bayes1$beta)

coeff_bayes1_pred_int <- t(apply(coeff_bayes1_preds, 1,
                                 FUN = function(x) matrix(
                                   c(mean(x), 
                                     quantile(x, 0.05), 
                                     quantile(x, 0.95)),
                                   nrow = 1
                                 )))

colnames(coeff_bayes1_pred_int) <- c("est", "ll", "ul")

# Add predictions to data
coeff_newdata %<>% bind_cols(as.data.frame(coeff_bayes1_pred_int))

## Plot truth and predictions from Bayesian example
coeff_bayes1_plot <- ggplot(coeff_newdata %>% filter(sex == 0),
                            aes(x = age, y = est, ymin = ll, ymax = ul)) +
  geom_line(aes(color = fusion_type)) +
  geom_point(inherit.aes = F, 
             data = enframe(cranio_coeff_smooth$xp),
             aes(x = value, y = 0),
             fill = "yellow", shape = 23) +
  geom_ribbon(aes(fill = fusion_type), alpha = 0.1) +
  facet_wrap(~fusion_type, scales = "free_y") +
  scale_color_discrete(palette = fusion_color_dict$color) +
  scale_fill_discrete(palette = fusion_color_dict$color) +
  labs(x = "Age (Days)", y = "Est. Basis Coefficient", 
       color = "Fusion Type",
       fill = "Fusion Type",
       title = "Bayesian Hierarchical Coefficient Functions") +
  theme_minimal() +
  theme(legend.position = "bottom")

coeff_bayes1_plot

# Add y limit so that patterns are more visible
coeff_bayes1_plot_lim <- coeff_bayes1_plot +
  coord_cartesian(ylim = max(coeff_y)*c(-0.5, 1.25))

coeff_bayes1_plot_lim
```

# Combined Hierarchical Model {#sec-joint-model}

## Model Specification {#sec-joint-model-spec}

### Data Description

First, let $\mathcal{S} \subset \mathbb{R}^2$ be a compact set captured in each image. Suppose we observe $M$ locations $\{ s_1, \ldots, s_M \} = \bS \subseteq \mathcal{S}$. For each location $s_m \ (m = 1, \ldots, M)$, note that there is a corresponding set of coordinates $(x_m, y_m) \in \mathbb{R}^2$.

Second, let $G_i(s_m)$ represent the measurement of growth in image $i$ at location $s_m$. Therefore, each image consists of the observations $\bG_i(\bS) = [G_i(s_1), \ldots, G_i(s_M)]^\top$. For simplicity, we'll write $\bG_i = \bG_i(\bS)$, and $\bG_i \text{ is } M \times 1$. Next, suppose we observe growth measurements on the same set of locations $\bS$ for $N$ individuals, so that we have a full data matrix of $\bG = [\bG_1, \ldots, \bG_N]^\top$, where $\bG \text{ is } N \times M$:

$$
\bG = \begin{bmatrix}
\bG_1^\top \\
\vdots \\
\bG_N^\top
\end{bmatrix} = 
\begin{bmatrix}
G_1(s_1) & G_1(s_2) & \ldots & G_1(s_M) \\
G_2(s_1) & G_2(s_2) & \ldots & G_2(s_M) \\
\vdots & & \ddots & \vdots \\
G_N(s_1) & G_N(s_2) & \ldots & G_N(s_M)
\end{bmatrix}
$$

Third, suppose that along with the observed growth measurements, we also have $C$ covariates of demographic information related to each image, $\bX_i$, so that $\bX$ is an $N \times C$ matrix of fixed covariates.

### Model Form

We treat each image's $M$ growth measurements $\bG_i(\mathcal{S})$ as one functional outcome, and we would like to estimate a mean function $\mu_i(\mathcal{S})$ for these outcomes based on demographic covariates. We also need to account for image-level random effects that are likely spatially correlated, which we could write as $\delta_i(\mathcal{S})$. Generally, this model takes the form:

$$
\begin{aligned}
\bG_i(\bS) &= \mu_i(\bS, \bX_i) + \delta_i(\bS) + \epsilon_i(\bS) \\
\mu_i(s_m, \bX_i) &= \beta^*_0(s_m) + \beta_1^*(s_m, X_{i1}) + ... + \beta_C^*(s_m, X_{iC}) \\
E[\delta_i(s_m)] &= 0 \\
E[\epsilon_i(s_m)] &= 0 \\
Var[\epsilon_i(s_m)] &= \sigma^2
\end{aligned}
$$

To estimate this model, we use a Bayesian hierarchical framework. In this framework, it is easiest to rewrite the model slightly such that $\eta_i = \mu_i + \delta_i$. Then:

$$
\begin{aligned}
G_i(s_m) &= \eta_i(s_m) + \epsilon_i(s_m) \\
\epsilon_i(s_m) & \iid N(0, \sigma^2)
\end{aligned}
$$

### Mean Estimation

To estimate $\eta_i(s_m)$, we use a spline expansion: $\eta_i(s_m) = \sum_{k=1}^K \gamma_{ik} B_k(s_m)$. Since $\mathcal{S}$ is a compact set, we use a soap film smoother expansion to define each $B_k(\cdot)$, and calculate the value of $B_k(s_m)$ using the corresponding coordinates $B_k(x_m, y_m)$. We define a $10 \times 10$ grid of knots, 60 of which fall within $\mathcal{S}$. We have no prior conditions for the values of $G$ along the boundary of $\mathcal{S}$, so we also estimate the boundary values with a cyclic spline of dimension 10. Together, this results in $K_f = 60$ and $K_b = 9$ basis functions, for a total of $K = 69$ parameters in the spline expansion. Therefore, we can rewrite the above equation as 

$$
\begin{aligned}
G_i(s_m) &= \sum_{k=1}^{K = 69} \gamma_{ik} B_k(s_m) + \epsilon_i(s_m) \\
\implies \bG_i(\bS) &= \sum_{k=1}^{K = 69} \gamma_{ik} B_k(\bS) + \epsilon_i(\bS) \\ \\
&= \bB(\bS) \bgamma_i \ + \bepsilon_i(\bS) \\
\implies [ \bG_i | \bgamma_i, \sigma^2 ] & \iid N_M(\bB \bgamma_i, \ \sigma^2 \bI_M) \\
\implies [ vec(\bG^\top) | \bgamma, \sigma^2 ] &\sim N_{N*M} ([\bI_N \otimes \bB] \ vec(\bgamma^\top), \ \sigma^2 \bI_{N*M}),
\end{aligned}
$$

where

$$
\bgamma = \begin{bmatrix}
\bgamma_1^\top \\
\vdots \\
\bgamma_N^\top
\end{bmatrix} 
= 
\begin{bmatrix}
\gamma_{11} & \ldots & \gamma_{1K} \\
\vdots & \ddots & \vdots \\
\gamma_{N1} & \ldots & \gamma_{NK}
\end{bmatrix}.
$$

### Soap Film Basis Function Estimation

At this level, we have modeled each individual image's spatial pattern across $\bS$ with the parameters $\bgamma$. Next, we account for the effect of the demographics $\bX$ on the growth $\bG$ by estimating each $\gamma_{ik}$ as a function of demographics. 

In our case, it is not appropriate to model $\gamma_{ik}$ as a linear function of the demographics $\bX_i$. Instead, we will model $\gamma_{ik}$ as a linear function of some untransformed covariates as well as a second spline expansion over Age, separated by suture fusion type. To make our model general enough to accomodate any linear combination of nonlinear functions of covariates $\bX_i$, let $\bW$ be the $N \times Q$ matrix containing untransformed covariates as well as realizations of basis spline expansions, and we assume

$$ E[\bgamma_i] = \bbeta^\top \bW_i^\top.$$

In addition, we want to penalize the complexity of each $\eta_i$ function over $\mathcal{S}$. This complexity can be quantified using one penalty matrix capturing the variation around the boundary, $\bP_b$, and one matrix capturing the variation within the interior space of the soap film, $\bP_f$. These can be combined to one "total" penalty matrix inducing smoothing across $\bgamma$ terms, along with smoothing constants $\lambda_b$ and $\lambda_f$:

$$
\bP_T^{(\gamma)} = \begin{bmatrix}
\lambda_b \bP_b & \bzero \\
\bzero & \lambda_f \bP_f
\end{bmatrix}
$$

To discourage the complexity captured by this matrix, we set the prior:

$$ 
\begin{aligned}
\ [\bgamma_i | \bbeta, \sigma^2, \lambda_b,  \lambda_f] & \iid N_K (\bbeta^\top \bW_i^\top, \ \sigma^2 \bP_T^{(\gamma) -}) \\
\implies [ vec(\bgamma) | \bbeta, \sigma^2, \lambda_b,  \lambda_f] & \iid N_K ([\bI_K \otimes \bW] \ vec(\bbeta), \ \sigma^2 [\bI_N \otimes \bP_T^{(\gamma) -}]);
\end{aligned}
$$

and 

$$
\bbeta = \begin{bmatrix} 
\bbeta_1 & \ldots & \bbeta_K 
\end{bmatrix} = 
\begin{bmatrix}
\beta_{11} & \ldots & \beta_{1K} \\
\vdots & \ddots & \vdots \\
\beta_{Q1} & \ldots & \beta_{QK} 
\end{bmatrix}.
$$

Note also that for a random matrix $Y$, we can use a commutation matrix to convert $vec(Y)$ to $vec(Y^\top)$, and a commutation matrix is orthogonal, so we can show that for any random matrix $Y$ such that $vec(Y)$ is multivariate normal distributed, $p[vec(Y) = y] = p[KY = Ky]$. This implies that $p[vec(\bgamma)] = p[vec(\bgamma^\top)]$. Therefore, although $vec(\bG^\top)$ relies on $vec(\bgamma^\top)$, we can use the distribution of $vec(\bgamma)$ in our full conditionals.

### Demographic Effect Estimation

Last, we need to specify how we will use $\bbeta$ and $\bW$ to estimate the relationships between the covariates in $\bX$ and $\bgamma$. 

First, suppose that there are $Q_U$ covariates in $\bX$ that we assume to be linearly related to $\bgamma$. Let $\tilde{\bX}$ be the $N \times (Q_U+1)$ matrix including a column vector of 1's and these $Q_U$ linear terms. 

Next, in our model, one covariate in $\bX$ is the Age of the subject in each image. We estimate a non-linear function of this covariate using a spline expansion: $h_0^*(Age_i) = \sum_{l=1}^L \alpha_{0l} h_l(Age_i)$. Then, for each of the $(R-1)$ suture fusion types, we estimate a fusion-specific deviation from this "reference" function: $h_r^*(Age_i)= I(\text{Fusion}_i = r) *[\sum_{l=1}^L \alpha_{rl} h_l(Age_i)]$

Putting these functions together with $\tilde{\bX}$, we have:

$$
\begin{aligned}
\bW &= \begin{bmatrix}
\tilde{\bX} & \bh_0^* & \bh_1^* & \ldots & \bh_{R-1}^*
\end{bmatrix} \\
&= 
\begin{bmatrix}
1 & \tilde{X}_{11} & \ldots & \tilde{X}_{1Q_U} & h_1(Age_1) & \ldots & h_L(Age_1) & I(\text{Fusion}_1 = 1)*h_1(Age_1) & ... &  I(\text{Fusion}_1 = R-1)*h_L(Age_1) \\
\vdots \\
1 & \tilde{X}_{N1} & \ldots & \tilde{X}_{NQ_U} & h_1(Age_N) & \ldots & h_L(Age_N) & I(\text{Fusion}_N = 1)*h_1(Age_i) & ... &  I(\text{Fusion}_N = R-1)*h_L(Age_N)
\end{bmatrix}
\end{aligned}.
$$

Likewise, we can write

$$
\bbeta_k^\top = \begin{bmatrix}
\theta_{0k} & \theta_{1k} & \ldots & \theta_{Q_U,k} & \alpha_{01k} & \ldots & \alpha_{0Lk} & \ldots & \alpha_{R-1, L,k}
\end{bmatrix},
$$

where $\btheta$ represents linear coefficients for untransformed covariates, and $\balpha$ represents coefficients that define the nonlinear spline expansions of age by fusion suture type.

Thus, $\bW$ is $N \times Q$ and $\bbeta$ is $Q \times K$, where $Q = Q_U + 1 + R*L$. 

Using these definitions, we want to penalize the complexity of each nonlinear function defined by one set $\balpha_{lk} = \{\alpha_{r1k}, \ldots, \alpha_{rLk}\}$. Similar to above, the complexity of the function defined by a nonlinear function can be quantified using an $L \times L$ penalty matrix $\bP_\alpha$ and a corresponding penalty constant $\lambda_\alpha$. 

For our model, we make two simplifying assumptions for $\balpha$: we use the same spline expansion for each suture fusion type, and we assume that all of these expansions share one $\lambda_\alpha$. Therefore, 
$$\bP_T^{(\alpha)} = \lambda_\alpha [\bI_R \otimes \bP_\alpha]$$.

To combine $\balpha_k$ with $\btheta_k$, we also need to establish a prior variance for $\btheta_k$ that leaves no penalty for complexity on $\btheta_k$. This is accomplished by using an identity matrix to represent the prior covariance of $\btheta_k$. Then:

$$
\bP_T^{(\beta)} = \begin{bmatrix}
\bI_{Q_U} & \bzero \\
\bzero & \bP_T^{(\alpha)}
\end{bmatrix} = 
\begin{bmatrix}
\bI_{Q_U} & \bzero & \ldots & \bzero \\
\bzero & \lambda_\alpha \bP_\alpha & \ldots &\bzero \\
\vdots & & \ddots & \vdots \\
\bzero & \bzero & \ldots & \lambda_\alpha \bP_\alpha
\end{bmatrix}
$$

Last, we assume no prior covariance between $\bbeta_k$ and $\bbeta_{k'}$, $k \neq k'$, and we add a prior scaling term $\tau^2$ to allow the magnitude of $\bbeta$ to match the magnitude of $\bgamma$.  Therefore, we assume the prior:

$$
\begin{aligned}
\ [\bbeta_k | \tau^2, \lambda_\alpha] & \iid N_Q (\bzero, \tau^2 \bP_T^{(\beta) -}) \\
\implies \ [vec(\bbeta) | \tau^2, \lambda_\alpha] & \sim N_{Q*K} (\bzero, \tau^2 [\bP_T^{(\beta) -} \otimes \bI_K])
\end{aligned}
$$

### Hyperpriors

Finally, we assume conjugate priors for the hyperparameters $\sigma^2, \ \lambda_b, \ \lambda_f, \, \tau^2, \text{ and } \lambda_\alpha$:

$$
\begin{aligned}
\sigma^2 & \sim \text{Inv. Gamma}(a_\sigma, b_\sigma) \\
\lambda_b & \sim \text{Gamma}(a_{\gamma b}, b_{\gamma b}) \\
\lambda_f & \sim \text{Gamma}(a_{\gamma f}, b_{\gamma f}) \\
\tau^2 & \sim \text{Inv. Gamma}(a_\tau, b_\tau) \\
\lambda_\alpha & \sim \text{Gamma}(a_\alpha, b_\alpha)
\end{aligned}
$$

