---
title: "Malformation Analysis"
author: "Ryan Taylor"
format: 
  html:
    embed-resources: true
    theme: paper
    highlight: haddock
    code-fold: true
    toc: true
    toc-location: left
    code-overflow: wrap
    df-print: kable
    knitr:
      opts_chunk:
        echo: true
        include: true
        message: false
        warning: false
        results: markup
        fig.align: center
        fig.height: 5
        fig.width: 7
        R.options:
          scipen: 999
          knitr.kable.NA: ""
number-sections: true
editor: source
---

\newcommand{\iid}{\,{\buildrel iid \over \sim}\,}

```{r}
#| label: setup
#| include: false

# Clear memory
rm(list = ls()); gc(verbose = F)

# Load packages
# if(!require("pacman")) install.packages("pacman")
pacman::p_load(
  # Specialty
  itsadug, spdep, concaveman,
  
  # Format
  latex2exp, table1, gtsummary,
  broom.helpers, broom.mixed, progressr, cli,
  
  # Modeling
  mgcv,
  
  # Useful extensions
  gtools, readxl, data.table, biglm, reshape2, ggridges,
  
  # Basics
  DescTools, tidyverse, magrittr, janitor, glue, skimr,
  knitr, kableExtra, ggpubr, ggpmisc
)

# Define function for predicting GAM on new data
GetPredictions_SimpleGAM <- function(model, newdata){
  
  # Get design matrix of basis functions
  nd_mat <- predict(model, newdata = newdata, type = "lpmatrix")
  
  # Get coefficients of the model
  m_coef <- coef(model)
  
  
}

# Set switches
DO_LOAD_MATRICES <- F
DO_PRELIM_MODEL_REGION <- F
DO_PRELIM_MODEL_PIXEL <- F
DO_PRELIM_MODEL_FUSION <- F

```

# Overview

This document consolidates our work on a Bayesian hierarchical model using a soap-film smoother to describe the patterns of volume abnormalities in children with craniosynostosis.

## Updates

-   Smoothed plots of age by fusion and region (@sec-age-smooth)
-   Written regression equations (@sec-prelim-models)
-   Preliminary models specific to region, pixel, and fusion now include smoothed age term (as a cubic regression spline), and plots show the shape of these splines
-   Intercepts are included in all models
-   Preliminary fusion-specific GAM plots include shading to show areas within the intracranial mask as opposed to outside of it (@sec-prelim-gam)
-   Test soap film smoother model fit to one subject, including boundary and grid of knots filtered to boundary (@sec-soap-film-1)

## Questions

-   Is Moran's I statistic relevant to show whether any given subset of the data actually has spatial dependence? I'm especially wondering whether the normative patients' growth has significant spatial dependence, or whether there is spatial dependence for very young patients.

## Collaborator Questions

-   7 patients in the data have *sphenoidal involvement* marked as 1. What effect might this have on volumes?
-   18 patients have *previous surgery* marked as 1. Is this specifically indicating a previous surgery related to craniosynostosis?
-   Should we control for image type in later analyses?

# Covariates Data

Data were sent by collaborators at Colorado (primarily Antonio Porras, Alejandro Cruz Guerrero, and Connor Elkhill).

Covariates are stored in a "merged_covariates" csv file that Josh Lukemire created from these data. Each image's malformation map is stored as a $500 \times 500$ matrix in its own RData file saved in a data folder in Josh's OneDrive.

We clean the covariate data set for easier analysis and to identify rare subsets of the data that should be excluded from our early work.

```{r}
#| label: load data

# Set directory for references
here::i_am("2025.08_Malformation_Analysis.qmd")

# Set data folder
DATA_FOLDER <- here::here("../../data/Lukemire, Joshua's files - data/")

# Set subfolder with image data
MATRIX_FOLDER <- file.path(DATA_FOLDER, "unsigned_distances")

# List all files with any extension (so no folders)
matrix_files <- list.files(MATRIX_FOLDER, pattern = "\\.")

# Read in covariates data
patient_images <- read_csv(file.path(DATA_FOLDER, "merged_covariates.csv"))
```

First, we ensure that all images in the covariate data are represented in the data folder and vice versa.

```{r}
#| label: clean file names

### Clean cranio covariates file

## Create new data set with easier names
cranio <- patient_images %>% 
  clean_names()

## Check if files in data were found in data folder
cranio %<>%
  mutate(
    fname = paste0(id_fmt, "_desc-unsigneddist.RData"),
    fname_exists = if_else(fname %in% matrix_files, 1, 0)
  )

# Print a table to check we have all files
fname_summ <- tibble(
  in_data_not_folder = sum(1 - cranio$fname_exists),
  in_folder_not_data = sum(!matrix_files %in% cranio$fname)
)

kbl(fname_summ, caption = "Check Files vs. Data Availability") %>% 
  kable_styling(full_width = F)

```

## Suture Fusion Categories

Next, we create a cleaner categorization of each image's suture fusions. Information about the suture fusions for each image are initially stored in a series of binary columns: *metopic_fused, sagittal_fused, left_coronal_fused, right_coronal_fused, left_lambdoid_fused, right_lambdoid_fused*. We combine these into one column and create categories for "Bicoronal" and "Multiple."

```{r}

#| label: clean sutures

## Classify sutures

# Currently stored in binary columns with names ending in "fused"
# Reshape data to identify all combinations of sutures
fusion_df <- cranio %>% 
  select(id_fmt, matches("fused")) %>% 
  pivot_longer(cols = -id_fmt) %>% 
  mutate(
    name = name %>% 
      str_remove("_fused") %>% 
      str_replace_all("_", " ") %>% 
      str_to_title()
  ) %>% 
  mutate(fusion = if_else(value == 1, name, NA_character_)) %>% 
  group_by(id_fmt) %>% 
  summarize(num_fusions = sum(value, na.rm = T),
            fusions = paste(na.omit(fusion), collapse = "/")) %>% 
  ungroup()

# (1) count number of sutures
# (2) identify bicoronal (the only useful instance of multiple sutures)
# (3) identify other instances of multiple sutures
# (3) assign remaining suture types
cranio %<>%
  left_join(fusion_df, by = "id_fmt") %>% 
  mutate(fusion_type = case_when(
    num_fusions == 0 ~ "None",
    fusions == "Left Coronal/Right Coronal" ~ "Bicoronal",
    num_fusions > 1 ~ "Multiple",
    T ~ fusions
  )) %>% 
  mutate(fusion_type = fct_infreq(fusion_type))

# Summarize frequencies of each suture type
fusion_summ <- cranio %>% 
  group_by(fusion_type) %>% 
  summarize(pics = n(),
            patients = n_distinct(id_base)) %>% 
  ungroup() %>% 
  arrange(desc(pics)) %>% 
  mutate(pics_pct = pics / sum(pics),
         patients_pct = patients / sum(patients)) %>% 
  adorn_totals("row")

kbl(fusion_summ, caption = "Suture Fusions: Full Data") %>% 
  kable_styling(full_width = F)
```

## Longitudinal Visits {#sec-longitudinal}

Next, we look at the availability of longitudinal data. No longitudinal data exist for patients with no suture fusions. Patients with suture fusions have an average of about 1.5 visits per person, and this average is fairly consistent even when we limit to the number of visits taken before the 1 year mark.

```{r}
#| label: visits

# Identify which visit each image is from and how many visits they had
cranio %<>%
  arrange(id_base, age) %>% 
  group_by(id_base) %>% 
  mutate(visit_num = 1:n()) %>% 
  # Count visits total and visits under age limit (currently 1 year)
  mutate(visits = max(visit_num)) %>% 
  ungroup()

# Summarize visits on average and max by fusion group
visits_summ <- cranio %>%
  group_by(fusion_type) %>% 
  summarize(patients = n_distinct(id_base),
            pics = n(),
            visits_avg = mean(visits),
            visits_max = max(visits)) %>%
  ungroup() %>% 
  arrange(desc(visits_avg))
  
kbl(visits_summ, caption = "Longitudinal Visits: Full Data") %>% 
  kable_styling(full_width = F)
```

## Exclusion Criteria {#sec-exclusion}

Next, we consider exclusion criteria.

First, there are a small number of patients who were labeled by the collaborators as "Craniosynostosis," but have no suture fusions. Close inspection reveals that these are 6 of the 7 patients in the data with "sphenoidal involvement." Therefore, we exclude these 7 images.

```{r}
#| label: group and sphenoidal

## Investigate any issues with fusion type
# Sanity check: fusion category against collaborators' "group" column
fusion_group_summ <- tabyl(cranio, fusion_type, group)

kbl(fusion_group_summ, caption = "Collaborators' Group vs. Fusions") %>% 
  kable_styling(full_width = F)

# 6 in "Craniosynostosis" group with no sutures all have sphenoidal involvement
sphen_summ <- cranio %>% 
  group_by(fusion_type, group, sphenoidal_involvement) %>% 
  summarize(n = n()) %>% 
  ungroup() %>% 
  arrange(desc(n))

kbl(sphen_summ, caption = "Sphenoidal Involvement Summary") %>% 
  kable_styling(full_width = F)
```

Second, it is worth investigating the category of "Multiple" fusions. We find that other than the "bicoronal" combination of fusions (left coronal and right coronal fusions but nothing else), no other combination has a sample size of more than 14. Therefore, we exclude 81 patients with "Multiple" fusions.

Third, we likewise find that *Right Lambdoid* and *Left Lambdoid* suture fusions are very rare in our data, with 10 and 7 images respectively. Therefore, we exclude lambdoid suture fusions.

```{r}
#| label: multiple sutures

## Investigate multiple sutures

fusion_multiple_summ <- cranio %>% 
  filter(fusion_type == "Multiple") %>% 
  group_by(num_fusions, fusions) %>% 
  summarize(pics = n(),
            patients = n_distinct(id_base)) %>% 
  ungroup() %>% 
  arrange(desc(pics))

kbl(fusion_multiple_summ, caption = "Summary of Multiple Fusions") %>% 
  kable_styling(full_width = F)

max_multiple_count <- max(fusion_multiple_summ$pics)
```

Fourth, we find that 18 images in the data are associated with a "previous surgery." Since we are more interested in the patterns associated with suture fusions before surgical intervention, we exclude these images. Notably, 6 of these 18 images have multiple suture fusions, so they would already be excluded.

```{r}
#| label: previous surgery

## Investigate previous surgery
surg_summ <- cranio %>% 
  filter(previous_surgery == 1) %>% 
  group_by(fusion_type, group, sphenoidal_involvement) %>% 
  summarize(n = n()) %>% 
  ungroup() %>% 
  arrange(desc(n)) %>% 
  adorn_totals("row")

kbl(surg_summ, caption = "Images with Previous Surgery") %>% 
  kable_styling(full_width = F)
```

### Data Subset {#sec-subset-data}

```{r}
#| label: filtered data

### Create subset data
# Exclude rare fusion categories: multiple, lambdoid
# Exclude sphenoidal involvement (in Craniosynostosis group without suture)
# Exclude patients with previous surgery (10)
# Only keep last picture taken
cranio_sub <- cranio %>% 
  filter(!fusion_type %in% c("Multiple", 
                             "Left Lambdoid", "Right Lambdoid") &
           sphenoidal_involvement %in% c(NA, 0) &
           previous_surgery %in% c(NA, 0) &
           visit_num == visits) %>% 
  # Reduce number of columns
  select(id_base, fname, fusion_type,
         age, sex) %>% 
  arrange(id_base, age)
```

### Age Summary {#sec-age-summary}

Fifth, we check the distribution of age in our data. We determine that over 98\% of the data falls under the age of 5 years, so we exclude those above this age, since we have very sparse observations to estimate an age effect.

```{r}
#| label: age summary

# Plot age densities
age_plot <- ggplot(cranio_sub) +
  geom_density_ridges(aes(x = age, 
                          y =  fusion_type, fill = fusion_type)) +
  scale_fill_discrete(guide = "none") +
  labs(x = "Age (Days)", y = "Fusion Type")

age_plot

# Summarize densities
have_fusions <- cranio_sub %>% 
  filter(fusion_type != "None")

have_fusions_quant <- have_fusions %$%
  quantile(age, probs = c(0.9, 0.95, 0.975, 0.99, 1)) %>% 
  enframe(name = "Percentile (with Fusions)", value = "Age") %>% 
  mutate(Years = Age / 365.25)

kbl(have_fusions_quant, 
    caption = "Max. Age Options: Patients with Fusions") %>% 
  kable_styling(full_width = F)

# Set maximum age as 5 years old
max_age <- 365 * 5

# Determine what percentile this corresponds to among patients with fusion
max_age_summ <- tibble("Age Cap" = max_age, 
                       "Percentile" = mean(have_fusions$age <= max_age))

max_age_summ
```

```{r}
#| label: subset by age

### Exclude patients above this age cap
cranio_sub %<>% filter(age <= max_age)
```

We generate a summary of the fusion types present in the remaining observations, to compare against the full data.

```{r}
#| label: summarize cleaned data

# Summarize counts of images and patients by fusion type
fusion_summ_sub <- cranio_sub %>% 
  group_by(fusion_type) %>% 
  summarize(pics = n(),
            patients = n_distinct(id_base)) %>% 
  ungroup() %>% 
  arrange(desc(pics)) %>% 
  mutate(pics_pct = pics / sum(pics),
         patients_pct = patients / sum(patients)) %>% 
  adorn_totals("row")

kbl(fusion_summ_sub, caption = "Fusions in Subset Data") %>% 
  kable_styling(full_width = F)
```

# Pointwise Difference Matrix Data {#sec-load-matrices}

Next, we load the malformation maps into one data set. Each malformation map is stored as a $500 \times 500$ matrix of pointwise absolute value differences, and is saved in its own RData file. We load each RData file in the remaining subset of images selected above. Then, we reshape the data from wide to long, so that each row contains one pixel's row, column, and pointwise difference value. We then store this long data as a tibble nested in the "data" column. Finally, we rearrange these data so that we can work with a data set with one row for each pixel, containing all participants' pointwise differences in that pixel. In addition, we merge each pixel's region label to this pointwise matrix data, so that we can associate this analysis with previous analyses.

```{r}
#| label: load matrix files

# Clear memory space
rm(list = setdiff(ls(), ls(pattern = "DO|FOLDER|cranio")))
rm(cranio); gc(verbose = F)

# Define function to read each file
load_matrix_data <- function(x){
  
  # Path to data file
  full_path <- file.path(MATRIX_FOLDER, x)
  
  # Load RData file (pointwise_diff)
  load(full_path)
  
  # Convert matrix to long format and remove values outside mask
  reduced_data <- pointwise_diff %>%
    reshape2::melt(varnames = c("row", "col"), value.name = "diff") %>%
    as_tibble() %>%
    drop_na(diff)
  
  return(list(reduced_data))
}

# Toggle whether to create matrix data
if(DO_LOAD_MATRICES){
# Create minimal matrix data
cranio_matrix <- cranio_sub %>%
  select(fname, age, sex, fusion_type) %>% 
  rowwise() %>%
  mutate(
    raw_data = load_matrix_data(fname)
  ) %>%
  ungroup() %>% 
  # Rearrange so each row is a pixel
  unnest(raw_data) %>%
  group_by(row, col) %>%
  nest() %>% 
  ungroup()

# Load regions and mask
load(paste0(DATA_FOLDER, "region_labels_and_mask.RData"))

cranio_matrix %<>%
  left_join(region_labels,
            by = c("row" = "X", "col" = "Y")) %>% 
  filter(value != 0) %>% 
  rename(region = value)

saveRDS(cranio_matrix, 
        file = here::here("intermediate", 
                          "covars_and_matrices.rds"))
} else{ 
  cranio_matrix <- readRDS(file = here::here("intermediate", 
                                             "covars_and_matrices.rds")) }

# Flip axes to match overhead view
cranio_matrix %<>%
  mutate(row = -row, 
         col = -col)
```

## Growth Matrix Visualizations {#sec-matrix-viz}

```{r}
#| label: cranio boundary

## Create boundary around cranium shape

# Create polygon around pixels in cranium shape
cranio_bound_shape <- cranio_matrix %>% 
  distinct(row, col) %>% 
  st_as_sf(coords = c("row", "col")) %>% 
  concaveman(concavity = 1)

# Create boundary as points
cranio_bound_matrix <- cranio_bound_shape %>% 
  # Add buffer so boundary is beyond data
  st_buffer(1) %>% 
  # Convert to line
  st_boundary() %>% 
  # Take points along this line at density = 1 unit
  st_line_sample(density = 1) %>% 
  st_coordinates() %>% 
  as_tibble() %>% 
  rename(row = X, col = Y)

# Visualize how boundary compares to all points
bound_viz <- ggplot() +
  geom_point(data = cranio_matrix %>% select(row, col),
             aes(x = row, y = col)) +
  geom_point(data = cranio_bound_matrix,
            aes(x = row, y = col), color = "blue") +
  coord_fixed() +
  labs(title = "Boundary Sanity Check") +
  theme_minimal()

bound_viz
```


As a first visualization, we plot the regions by pixel.

### Region Map

```{r}
#| label: plot regions

# Create dataset of only region assignment for each pixel
matrix_region <- cranio_matrix %>% 
  select(row, col, region) %>% 
  distinct()

# Identify regions' centroids for labels
matrix_region_center <- matrix_region %>% 
  group_by(region) %>% 
  summarize(row_center = mean(row),
            col_center = mean(col)) %>% 
  ungroup()

# Plot regions
plot_regions <- ggplot() +
  geom_raster(data = matrix_region, 
              aes(x = row, y = col, fill = factor(region))) +
  geom_text(data = matrix_region_center, 
            aes(x = row_center, y = col_center, label = factor(region))) +
  coord_fixed() +
  theme_void() +
  scale_fill_brewer(palette = "Set3", guide = "none") +
  labs(title = "Region Map", fill = "Region")

plot_regions
```

### Fusion Averages

Next, we summarize the growth by suture fusion and plot these averages on the brain shape.

```{r}
#| label: map fusions

# Clear memory space
rm(list = setdiff(ls(), ls(pattern = "DO|FOLDER|cranio"))); gc(verbose = F)

### Summarize differences by fusions

# Create dataset of average cell for each fusion
matrix_fusion <- cranio_matrix %>% 
  # Extract only fusion type and outcome from nested data
  hoist(data, "fusion_type", "diff") %>% 
  select(-c(data, region)) %>% 
  unnest_longer(col = c(fusion_type, diff)) %>% 
  # Summarize average by pixel and fusion type
  group_by(row, col, fusion_type) %>% 
  summarize(diff_avg = mean(diff)) %>% 
  ungroup()

# Plot brain map by fusion type
plot_fusion <- ggplot(matrix_fusion) +
  geom_raster(aes(x = row, y = col, fill = diff_avg)) +
  facet_wrap(~fusion_type) +
  coord_fixed() +
  theme_void() +
  scale_fill_viridis_c() +
  labs(title = "Fusion Average Growth",
       fill = "Avg. Growth") +
  theme(legend.position = "bottom")

# Print fusion maps
plot_fusion
```

### Fusion - Age Averages

We add to this summary by showing how these averages change on average over time, dividing age into 6 bins of roughly 2-month windows of ages. We also compare these results to the sample sizes in each of these age-fusion bins.

```{r}
#| label: plot fusion and age

### Investigate age

# Define age category cutoffs
# First, find quantiles of ages in craniosynostosis patients
age_fused_quant <- cranio_sub %>% 
  filter(fusion_type != "None") %$% 
  quantile(age, prob = seq(0, 1, 0.2))

# Next, use closest interpretable number of days
age_breaks <- c(0, 61, 122, 183, 366, 1830)

# Label these with spaces so they can wrap in plots
age_break_labels <- map_chr(2:length(age_breaks), 
                            ~paste0(age_breaks[.-1], 
                                    " - ", 
                                    if_else(. == length(age_breaks), "", "<"), 
                                    age_breaks[.]))

# Create dataset with fusion and age info
matrix_age_fusion <- cranio_matrix %>% 
  hoist(data, "fusion_type", "age", "diff") %>% 
  select(-c(data, region)) %>% 
  unnest_longer(col = c(fusion_type, age, diff)) %>% 
  # Add age categories
  mutate(age_bin = cut(age, 
                       breaks = age_breaks, labels = age_break_labels, 
                       right = FALSE, include.lowest = T)) %>% 
  # Summarize average in these bins
  group_by(row, col, fusion_type, age_bin) %>% 
  summarize(diff_avg = mean(diff)) %>% 
  ungroup()

# Plot brain map by fusion type and age
plot_fusion_age <- ggplot(matrix_age_fusion) +
  geom_raster(aes(x = row, y = col, fill = diff_avg)) +
  facet_grid(fusion_type ~ age_bin,
             labeller = label_wrap_gen(width = 10)) +
  coord_fixed() +
  scale_fill_viridis_c() +
  theme_void() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "Fusion - Age Average Growth",
       fill = "Avg. Growth") +
  theme(legend.position = "bottom")

# Print fusion maps
plot_fusion_age

## Determine sample size for these averages

# Count images per fusion and age bin
size_fusion_age <- cranio_sub %>% 
  mutate(age_bin = cut(age, 
                       breaks = age_breaks, labels = age_break_labels, 
                       right = FALSE, include.lowest = T)) %>% 
  group_by(fusion_type, age_bin) %>% 
  summarize(n = n()) %>% 
  ungroup() %>% 
  pivot_wider(names_from = age_bin, values_from = n, id_cols = fusion_type)

# Print table
kbl(size_fusion_age, caption = "Fusion-Age Sample Size") %>% 
  kable_styling(full_width = F)
```

### Age Trajectory {#sec-age-smooth}

In order to gain a better understanding of the specific shape of the age effect, we plot average growth by age, stratified by region and fusion type.

```{r}
#| label: age effect

# Check shape of age effect by region and fusion

# Summarize by age, region, and fusion type; have option to round to nearest month
matrix_age_region <- cranio_matrix %>% 
  hoist(data, "fusion_type", "age", "diff") %>% 
  select(-data) %>% 
  unnest_longer(col = c(fusion_type, age, diff)) %>% 
  mutate(age_round_month = round(age / 30)) %>% 
  group_by(fusion_type, age, region) %>% 
  summarize(diff_avg = mean(diff)) %>% 
  ungroup()

plot_age <- ggplot(matrix_age_region) +
  # Plot 1 line over age for each region-fusion combo
  geom_smooth(aes(x = age, y = diff_avg, 
                  color = fusion_type, fill = fusion_type),
              alpha = 0.2) +
  facet_wrap(~region) +
  labs(title = "Fusion - Region Average Growth by Age",
       x = "Age (Days)",
       y = "Avg. Growth",
       color = "Fusion Type",
       fill = "Fusion Type") +
  theme_minimal() +
  theme(legend.position = "bottom")

# Print plot
plot_age
```

Finally, we consider how each fusion type might differ between the two natal sexes recorded in the data.

```{r}
#| label: fusion by sex

# Take average pointwise differences in cell by fusion and sex
matrix_sex_fusion <- cranio_matrix %>% 
  hoist(data, "fusion_type", "sex", "diff") %>% 
  select(-c(data, region)) %>% 
  unnest_longer(col = c(fusion_type, sex, diff)) %>% 
  # Summarize average in these bins
  group_by(row, col, fusion_type, sex) %>% 
  summarize(diff_avg = mean(diff)) %>% 
  ungroup()

# Plot brain map by fusion type and age
plot_fusion_sex <- ggplot(matrix_sex_fusion) +
  geom_raster(aes(x = row, y = col, fill = diff_avg)) +
  facet_grid(fusion_type ~ sex,
             labeller = label_wrap_gen(width = 9)) +
  coord_fixed() +
  scale_fill_viridis_c() +
  theme_void() + 
  theme(axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "Fusion-Sex Average Growth",
       fill = "Avg. Growth") +
  theme(legend.position = "bottom")

# Print fusion maps
plot_fusion_sex
```

# Preliminary Models {#sec-prelim-models}

## Region-Level Independent

Next, we get a better understanding of how these demographics interact with fusion type through preliminary regression models. First, we regress the region-level average growth for each subject on fusion indicators, age, and sex. Based on the plot above, we include age as a smooth term; all other covariates are included as indicator variables. We also run a multiple linear regression with age as a linear term.

We denote the average growth for patient $i$ in region $r$ as $G_{ir}$. We let $f_r(\cdot)$ be a cubic regression spline estimated on observations from region $r$; we choose a cubic regression spline for faster estimation of many regressions, especially when we extend to a pixel-level model below. Then these models can be written:

$$
\begin{gather*}
\begin{split}
G_{ir} = {} & \beta_{0, r} + \beta_{1,r} [\text{Sagittal}]_i + \beta_{2,r} [\text{Metopic}]_i + \\
& \beta_{3,r} [\text{Right Coronal}]_i + \beta_{4,r} [\text{Left Coronal}]_i + \\
& \beta_{5,r} [\text{Left and Right Coronal}]_i + \\
& \beta_{6,r} [\text{Sex: Male}]_i + f_r(Age_i) + \epsilon_{ir}
\end{split} \\ \\
\begin{split}
G_{ir} = {} & \beta_{0, r} + \beta_{1,r} [\text{Sagittal}]_i + \beta_{2,r} [\text{Metopic}]_i + \\
&\beta_{3,r} [\text{Right Coronal}]_i + \beta_{4,r} [\text{Left Coronal}]_i + \\
& \beta_{5,r} [\text{Left and Right Coronal}]_i + \\
& \beta_{6,r} [\text{Sex: Male}]_i + \beta_{7,r}(Age_i) + \epsilon_{ir}
\end{split} \\ \\
\begin{split}
\epsilon_{ir} \,{\buildrel iid \over \sim}\,{} & N(0, \sigma^2_r)
\end{split}
\end{gather*}
$$

```{r}
#| label: region model

# Clear memory space
rm(list = setdiff(ls(), ls(pattern = "DO|FOLDER|cranio"))); gc(verbose = F)

# Add fusion variables for regression
fusion_dict <- tibble(
  fusion_type = c("None", 
                  "Sagittal", "Metopic", 
                  "Right Coronal", "Bicoronal", "Left Coronal"),
  fused_Sagittal = c(0, 1, 0, 0, 0, 0),
  fused_Metopic = c(0, 0, 1, 0, 0, 0),
  fused_RCoronal = c(0, 0, 0, 1, 1, 0),
  fused_LCoronal = c(0, 0, 0, 0, 1, 1)
)

# Recreate pixel map with region labels
matrix_region <- cranio_matrix %>% 
  select(row, col, region) %>% 
  distinct()

# Identify regions' centroids for labels
matrix_region_center <- matrix_region %>% 
  group_by(region) %>% 
  summarize(row_center = mean(row),
            col_center = mean(col)) %>% 
  ungroup()

if(DO_PRELIM_MODEL_REGION){
  
  ## Run models for average pointwise differences in each region
  region_models <- cranio_matrix %>% 
    # Rearrange to incorporate subject data
    unnest(data) %>% 
    # Take average in each region for each image
    group_by(fname, age, sex, fusion_type, region) %>% 
    summarize(diff_avg = mean(diff)) %>% 
    ungroup() %>%
    # Add fusion indicators
    left_join(fusion_dict, by = "fusion_type") %>% 
    # Nest region-specific data
    group_by(region) %>% 
    nest() %>% 
    ungroup() %>% 
    # Fit model with age as a spline (cubic regression basis for fast estimation)
    mutate(model = map(data,
                       ~gam(diff_avg ~ 
                              fused_Sagittal + fused_Metopic +
                              fused_RCoronal * fused_LCoronal +
                              sex + s(age, bs = "cr"),
                            data = .x, method = "REML")),
           # Also try linear age
           model_linear = map(data,
                              ~lm(diff_avg ~ 
                                    fused_Sagittal + fused_Metopic +
                                    fused_RCoronal * fused_LCoronal +
                                    sex + age,
                                  data = .x)))
  
  saveRDS(
    cranio_region_models, 
    file = here::here("intermediate", 
                      "prelim_models_region.rds"))
  
} else { 
  region_models <- readRDS(
    file = here::here("intermediate", 
                      "prelim_models_region.rds")) }

region_models %<>% 
  ## Add bicoronal variance to final coefficients; output coefficient summary
  mutate(across(matches("model"),
                ~map(.,
                     function(m){
                       
                       # Make constrast vector identifying all coronal variables
                       is_coronal <- as.numeric(str_detect(names(coef(m)), "Coronal"))                        
                       
                       # Compute variance of bicoronal combination
                       bicoronal_v <- t(is_coronal) %*% vcov(m) %*% is_coronal
                       bicoronal_e <- t(is_coronal) %*% coef(m)
                       
                       # Add bicoronal as row in tidy summary
                       bicoronal_row <- tibble(
                         term = "fused_Bicoronal",
                         estimate = bicoronal_e[1,1],
                         std.error = sqrt(bicoronal_v[1,1])) %>% 
                         mutate(p.value = 2 * pnorm(abs(estimate / std.error), 
                                                    lower.tail = FALSE))
                       
                       # Output coefficient table (smaller than model)
                       return(tidy(m, parametric = TRUE) %>% 
                                bind_rows(bicoronal_row))
                     }),
                .names = "{.col}_coeffs")) %>% 
  # Rename individual model terms before merging
  mutate(model_coeffs = map(model_coeffs, 
                            ~rename_with(., 
                                         .cols = -term,
                                         .fn = ~paste0(., "_smooth"))),
         model_linear_coeffs = map(model_linear_coeffs, 
                                   ~rename_with(., 
                                                .cols = -term,
                                                .fn = ~paste0(., "_linear")))) %>% 
  # Merge model summaries
  mutate(coeffs_combo = map2(model_coeffs, model_linear_coeffs,
                             ~full_join(.x, .y, by = "term"))) %>% 
  # Clean model summary
  mutate(coeffs_combo_cln = map(coeffs_combo, 
                                ~ filter(., term != "fused_RCoronal:fused_LCoronal") %>% 
                                  arrange(-str_detect(term, "fused")) %>% 
                                  mutate(term = fct_inorder(term))))

# Merge model summaries with matrix for plotting
model_region_matrix <- region_models %>% 
  select(region, coeffs_combo_cln) %>% 
  unnest(coeffs_combo_cln) %>% 
  full_join(matrix_region, relationship = "many-to-many")

# Plot parametric coefficients
plot_region_smooth <- ggplot(model_region_matrix %>% filter(term != "age")) +
  geom_raster(aes(x = row, y = col, fill = estimate_smooth)) +
  geom_text(data = matrix_region_center, 
            aes(x = row_center, y = col_center, label = region)) +
  facet_wrap(~term) +
  coord_fixed() +
  scale_fill_viridis_c(option = "turbo") +
  theme_void() +
  labs(title = "Region-Level Smooth Age Model",
       fill = "Coefficient") +
  theme(legend.position = "bottom")

# Plot parameters from alternate version of the model
plot_region_linear <- ggplot(model_region_matrix) +
  geom_raster(aes(x = row, y = col, fill = estimate_linear)) +
  geom_text(data = matrix_region_center, 
            aes(x = row_center, y = col_center, label = region)) +
  facet_wrap(~term) +
  coord_fixed() +
  scale_fill_viridis_c(option = "turbo") +
  theme_void() +
  labs(title = "Region-Level Linear Age Model",
       fill = "Coefficient") +
  theme(legend.position = "bottom")

# Print these plots
plot_region_linear

plot_region_smooth

## Plot smoothed effect
# Define range for predictions
region_smooth_newdata <- tibble(
  fused_Sagittal = 0,
  fused_Metopic = 0,
  fused_RCoronal = 0,
  fused_LCoronal = 0,
  sex = 0,
  age = seq(min(cranio_sub$age), 
            max(cranio_sub$age),
            length.out = 30)
)

model_region_smooth <- region_models %>% 
  select(region, model) %>% 
  # Add predictions from smooth term
  mutate(smooth_pred = map(
    model, 
    
    ~get_predictions(., 
                     cond = list(sex = 1,
                                 age = seq(min(cranio_sub$age), 
                                           max(cranio_sub$age),
                                           length.out = 30)), 
                     se = TRUE, f = 1.96,
                     print.summary = FALSE))) %>% 
  
  select(-model) %>% 
  unnest(smooth_pred) %>% 
  mutate(ci_ll = fit - CI,
         ci_ul = fit + CI)

# Add linear age effect to compare
model_region_linear <- region_models %>% 
  select(region, model_linear) %>% 
  mutate(newdata = list(tibble(fused_Sagittal = 0,
                          fused_Metopic = 0,
                          fused_RCoronal = 0,
                          fused_LCoronal = 0,
                          sex = 1,
                          age = seq(min(cranio_sub$age), 
                                    max(cranio_sub$age),
                                    length.out = 30))
  )) %>% 
  ungroup() %>% 
  mutate(lin_pred = map2(
    model_linear, newdata,
    ~predict(.x, newdata = newdata[[1]], se.fit = TRUE) %>% 
      as.data.frame())) %>% 
  select(-model_linear) %>% 
  unnest(c(newdata, lin_pred)) %>% 
  mutate(ci_ll = fit - 1.96 * se.fit,
         ci_ul = fit + 1.96 + se.fit)

# Combine smooth and linear predictions
model_region_combo <- model_region_smooth %>% 
  select(region, age, fit, ci_ll, ci_ul) %>% 
  mutate(type = "Smooth") %>% 
  bind_rows(
    model_region_linear %>% 
      select(region, age, fit, ci_ll, ci_ul) %>% 
      mutate(type = "Linear")
  )

# Plot 1 smoothed effect per region
plot_region_smooth_pred <- ggplot(model_region_combo) +
  geom_line(aes(x = age, y = fit, color = factor(region), linetype = rev(type))) +
  geom_ribbon(aes(x = age, ymin = ci_ll, ymax = ci_ul, 
                  fill = factor(region),
                  linetype = type),
              alpha = 0.1) +
  facet_wrap(~type, nrow = 1) +
  scale_linetype(guide = "none") +
  labs(x = "Age (Days)", y = "Predicted Growth (No Fusion Male)", 
       color = "Region", fill = "Region") +
  theme_minimal() +
  theme(legend.position = "bottom")

plot_region_smooth_pred

```

## Pixel-Level Independent

To get a more detailed spatial perspective, we then run the same regression model on each individual pixel in the data.

We denote the growth for patient $i$ in matrix row $x$ and column $y$ as $G_{i}(x,y)$ and let $f(x,y,\cdot)$ be a cubic regression spline function estimated on observations at pixel $[x,y]$; then this model can be written:

$$
\begin{gather*}
\begin{split}
G_{i}(x,y) = {} & \beta_{0}(x,y) + \beta_{1}(x,y) [\text{Sagittal}]_i + \beta_{2}(x,y) [\text{Metopic}]_i + \\
& \beta_{3}(x,y) [\text{Right Coronal}]_i + \beta_{4}(x,y) [\text{Left Coronal}]_i + \\
& \beta_{5}(x,y) [\text{Left and Right Coronal}]_i + \\
& \beta_{6}(x,y) [\text{Sex: Male}]_i + f(x,y, Age_i) + \epsilon_{i}(x,y)
\end{split} \\ \\
\epsilon_{i}(x,y) \,{\buildrel iid \over \sim}\,{} N(0, \sigma^2(x,y))
\end{gather*}
$$

```{r}
#| label: pixel model

# Clear memory space
rm(list = setdiff(ls(), ls(pattern = "DO|FOLDER|cranio"))); gc(verbose = FALSE)

if(DO_PRELIM_MODEL_PIXEL){
  
  ## Run models for pointwise differences in each pixel
  pixel_models <- cranio_matrix %>% 
    mutate(
      model = map(data,
                  function(d){
                    d_reg <- d %>% left_join(fusion_dict, by = "fusion_type")
                    
                    mdl <- bam(diff ~ 
                                 fused_Sagittal + fused_Metopic +
                                 fused_RCoronal * fused_LCoronal +
                                 sex + s(age, bs = "cr"),
                               data = d_reg)
                    
                    return(mdl)
                  })) %>% 
    mutate(coeffs = map(
      model,
      function(m){
        
        # Make constrast vector identifying all coronal variables
        is_coronal <- as.numeric(str_detect(names(coef(m)), "Coronal"))                        
        
        # Compute variance of bicoronal combination
        bicoronal_v <- t(is_coronal) %*% vcov(m) %*% is_coronal
        bicoronal_e <- t(is_coronal) %*% coef(m)
        
        # Add bicoronal as row in tidy summary
        bicoronal_row <- tibble(
          term = "fused_Bicoronal",
          estimate = bicoronal_e[1,1],
          std.error = sqrt(bicoronal_v[1,1])) %>% 
          mutate(statistic = estimate / std.error,
                 p.value = 2 * pnorm(abs(estimate / std.error), 
                                     lower.tail = FALSE))
        
        # Output coefficient table (smaller than model)
        return(tidy(m, parametric = TRUE) %>% 
                 bind_rows(bicoronal_row))
      }))
  
  # Combine pixel-level models and add region column
  model_pixel_summ <- pixel_models %>% 
    select(row, col, coeffs) %>% 
    unnest(coeffs) %>% 
    # Remove raw bicoronal interaction term (replaced with Bicoronal combo)
    filter(term != "fused_RCoronal:fused_LCoronal") %>% 
    arrange(row, col, -str_detect(term, "fused")) %>% 
    mutate(term = fct_inorder(term))
  
  # Extract smother predictions
  model_pixel_smooth <- pixel_models %>% 
    select(row, col, region, model) %>% 
    # Add plot of smoothed term
    mutate(smooth_pred = map(
      model, 
      ~get_predictions(., 
                       cond = list(sex = 1,
                                   age = seq(min(cranio_sub$age), 
                                             max(cranio_sub$age),
                                             length.out = 20)), 
                       se = TRUE, f = 1.96,
                       print.summary = FALSE))) %>% 
    
    select(-model) %>% 
    unnest(smooth_pred) %>% 
    mutate(ci_ll = fit - CI,
           ci_ul = fit + CI)
  
  save(model_pixel_summ, model_pixel_smooth,
       file = here::here("intermediate", 
                         "prelim_model_pixels_summ.rda"))
} else{ load(file = here::here("intermediate", 
                               "prelim_model_pixels_summ.rda")) }

# Plot pixel-level coefficients
plot_pixel <- ggplot(model_pixel_summ) +
  geom_raster(aes(x = row, y = col, fill = estimate)) +
  facet_wrap(~term) +
  coord_fixed() +
  scale_fill_viridis_c(option = "turbo") +
  theme_void() +
  labs(title = "Pixel-Level Model",
       fill = "Coefficient") +
  theme(legend.position = "bottom")

plot_pixel

# Plot 1 age smoothed effect per region
plot_pixel_pred <- ggplot(model_pixel_smooth) +
  geom_line(aes(x = age, y = fit, 
                group = interaction(row, col),
                color = factor(region)),
            alpha = 0.05, linewidth = 0.05) +
  labs(x = "Age (Days)", y = "Predicted Growth", 
       color = "Region") +
  theme_minimal() +
  scale_color_viridis_d() +
  guides(color = guide_legend(override.aes = list(alpha = 1, linewidth = 0.5))) +
  theme(legend.position = "bottom")

plot_pixel_pred
```

## Fusion-Independent GAM {#sec-prelim-gam}

Next, for a preliminary spatial model, we split the data only by fusion type. Based on the plots above, the age and sex effects are basically uniform across all of the pixels, so we include these as linear terms that do not vary spatially. Otherwise, we only include a tensor spline interaction between row and column indices. Our model does not reflect the brain boundary. It is a generalized additive model with an intercept and a tensor product smooth of row and column indices.

We denote the growth for patient $i$ with suture fusion type $s$ in matrix row $j$ and column $k$ as $Y_{sijk}$, let $f_{s}(\cdot)$ be a cubic regression spline estimated on observations from patients with suture fusion $s$, and let $g_s(\cdot, \cdot)$ be a tensor product smooth with a cubic regression spline basis; then this model can be written:

$$
\begin{gather*}
\begin{split}
G_{si}(x,y) = {} & \beta_{0, s} + \beta_{1,s} [\text{Sex: Male}]_i + \\
& f_{s}(Age_i) + h_s(x, y) + \epsilon_{si}(x,y)
\end{split} \\ \\
\epsilon_{si}(x,y) \,{\buildrel iid \over \sim}\,{} N(0, \sigma^2_s)
\end{gather*}
$$

```{r}
#| label: preliminary gam

# Clear memory space
rm(list = setdiff(ls(), ls(pattern = "DO|FOLDER|cranio"))); gc(verbose = FALSE)

if(DO_PRELIM_MODEL_FUSION){
  
  # Run preliminary spatial models
  fusion_models <- cranio_matrix %>% 
    unnest(data) %>% 
    # Leave out patients with no fusions, since this is a huge population
    filter(!fusion_type %in% c("None")) %>% 
    # Split data by fusion type
    group_by(fusion_type) %>% 
    nest() %>% 
    ungroup() %>% 
    # Fit model with linear sex effect, smoothed age term, and row-col tensor surface
    mutate(model = map(data,
                       ~bam(diff ~ sex + s(age, bs = "cr") + 
                              te(row, col),
                            data= .x)))
  
  saveRDS(
    fusion_models, 
    file = here::here("intermediate", "prelim_models_fusion_te.rds"))
} else { 
  fusion_models <- readRDS(
    file = here::here("intermediate", "prelim_models_fusion_te.rds")) }

fusion_models  %<>% 
  # Predict from this model for spatial pattern (1 year old male)
  mutate(pred_data = map(
    model,
    ~get_predictions(.x,
                     cond = list(
                       "age" = 365,
                       "sex" = 1,
                       "row" = seq(from = min(cranio_matrix$row), 
                                   to = max(cranio_matrix$row), 
                                   length.out = 50),
                       "col" = seq(from = min(cranio_matrix$col), 
                                   to = max(cranio_matrix$col), 
                                   length.out = 50)
                     ),
                     se = TRUE, f = 1.96,
                     print.summary = FALSE
    )))

# Flatten predictions into their own data
fusion_model_summ <- fusion_models %>% 
  select(fusion_type, pred_data) %>% 
  unnest(pred_data)

# Plot these predictions (in a rectangle shape with brain mask)
plot_model_fusion <- ggplot() +
  geom_raster(data = fusion_model_summ, 
              aes(x = row, y = col, fill = fit)) +
  geom_sf(data = cranio_bound_shape, fill = NA, linewidth = 1) +
  facet_wrap(~fusion_type) +
  scale_fill_viridis_c(option = "turbo") +
  scale_alpha(range = c(0.5, 1), guide = "none") +
  theme_void() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "Fusion-Specific GAM",
       fill = "Pred. Growth\n(1 y.o. M)") +
  theme(legend.position = "bottom")

plot_model_fusion

## Extract age smoothed predictions
model_fusion_smooth <- fusion_models %>% 
  select(fusion_type, model) %>% 
  # Add plot of smoothed term
  mutate(smooth_pred = map(
    model, 
    ~get_predictions(., 
                     cond = list(sex = 1,
                                 age = seq(min(cranio_sub$age), 
                                           max(cranio_sub$age),
                                           length.out = 30)), 
                     se = TRUE, f = 1.96,
                     print.summary = FALSE))) %>% 
  
  select(-model) %>% 
  unnest(smooth_pred) %>% 
  mutate(ci_ll = fit - CI,
         ci_ul = fit + CI)

# Plot 1 smoothed effect per fusion type
plot_fusion_pred <- ggplot(model_fusion_smooth) +
  geom_line(aes(x = age, y = fit, color = fusion_type)) +
  geom_ribbon(aes(x = age, ymin = ci_ll, ymax = ci_ul, 
                  group = fusion_type, fill = fusion_type),
              alpha = 0.5) +
  labs(x = "Age (Days)", y = "Predicted Growth (Male)", 
       color = "Fusion Type", fill = "Fusion Type") +
  theme_minimal() +
  theme(legend.position = "bottom")

plot_fusion_pred

## Compile parametric coefficients (Intercept and sex)
model_fusion_params <- fusion_models %>% 
  select(fusion_type, model) %>% 
  mutate(params = map(model, ~tidy(., parametric = TRUE))) %>% 
  select(-model) %>% 
  unnest(params)

# Print table of these parametric coefficients
kbl(model_fusion_params, 
    caption = "Preliminary 2D GAM Parametric Coefficients") %>% 
  kable_styling(full_width = F)

```

# Individual Soap Film Models {#sec-soap-film-1}

Next, we introduce the soap film smoother. To fit this model, we first need to create the boundary that determines the extent of the model and the set of knots in $\mathbb{R}^2$ that will define the smoothing functions.

```{r}
#| label: soap infrastructure

## Create boundary as list
cranio_bound <- list(
  list(row = cranio_bound_matrix$row,
       col = cranio_bound_matrix$col)
)

## Create knots
grid_n <- 10

# Identify row knot points
grid_row <- seq(min(cranio_bound_matrix$row), max(cranio_bound_matrix$row), 
                length.out = grid_n)

# Identify column knot points
grid_col <- seq(min(cranio_bound_matrix$col), max(cranio_bound_matrix$col), 
                length.out = grid_n)

# Expand to grid of knots
grid_all <- expand.grid("row" = grid_row, "col" = grid_col)

# Crop to cranial boundary
grid_k <- grid_all %>% 
  st_as_sf(coords = c("row", "col")) %>% 
  st_filter(cranio_bound_shape) %>% 
  st_coordinates() %>% 
  as_tibble() %>% 
  rename(row = X, col = Y)

# Visualize grid
grid_viz <- ggplot() +
  geom_point(data = cranio_bound_matrix, aes(x = row, y = col), color = "blue") +
  geom_point(data = grid_all, aes(x = row, y = col)) +
  geom_point(data = grid_k, aes(x = row, y = col), color ="red") +
  coord_fixed() +
  labs(title = "Grid Sanity Check")

grid_viz
```

Next, we use the set of row and column indices to create a design matrix of 2-dimensional smoothers that can be used for every individual image.

```{r}
#| label: soap film basis

# Keep only what we need to build smoother
matrix_generic <- cranio_matrix %>% 
  select(row, col)

# Generate soap film smoother object
cranio_so_smooth <- smoothCon(
  s(row, col, bs = "so", xt = list(bnd = cranio_bound)),
  data = matrix_generic, knots = grid_k
)
```


```{r}
#| label: prelim soap

## Filter to one patient for test model
single_matrix <- cranio_matrix %>% 
  unnest(data) %>% 
  filter(fusion_type == "Sagittal") %>% 
  filter(fname == first(fname))

## Fit soap film
single_soap <- gam(diff ~ cranio_so_smooth,
                   data = single_matrix, method = "REML")

## Plot predictions from soap film

# Predict on a denser grid than the knots
single_soap_pred <- get_predictions(
  single_soap, rm.ranef = FALSE, se = TRUE, f = 1.96,
  cond = list(
    row = seq(min(bound_matrix$row), max(bound_matrix$row), 
              length.out = 50),
    col = seq(min(bound_matrix$col), max(bound_matrix$col),
              length.out = 50)
  )) %>% 
  # Remove NA predictions outside boundary
  filter(!is.na(fit))

# Plot predictions
plot_single_soap <- ggplot(single_soap_pred) +
  geom_raster(aes(x = row, y = col, fill = fit)) +
  scale_fill_viridis_c(option = "turbo") +
  labs(x = NULL, y = NULL, fill = "Pred. Growth") +
  coord_fixed() +
  theme_void() +
  theme(legend.position = "bottom")

# Plot actual observations
plot_single_mx <- ggplot(single_matrix) +
  geom_raster(aes(x = row, y = col, fill = diff)) +
  scale_fill_viridis_c(option = "turbo") +
  labs(x = NULL, y = NULL, fill = "Actual Growth") +
  coord_fixed() +
  theme_void() +
  theme(legend.position = "bottom")

# Combine and print
plot_single_both <- ggarrange(plot_single_mx, plot_single_soap)
plot_single_both

## Get soap film model matrix
single_soap_mx <- predict(single_soap, type = "lpmatrix")
```
