---
title: "Malformation Analysis"
author: "Ryan Taylor"
format: 
  html:
    embed-resources: true
    theme: paper
    highlight: haddock
    code-fold: true
    toc: true
    toc-location: left
    code-overflow: wrap
    df-print: kable
    knitr:
      opts_chunk:
        echo: true
        include: true
        message: false
        warning: false
        results: markup
        fig.align: center
        fig.height: 5
        fig.width: 7
        R.options:
          scipen: 999
          knitr.kable.NA: ""
editor: source
---

```{r}
#| label: setup
#| include: false

# Clear memory
rm(list = ls()); gc()

# Load packages
# if(!require("pacman")) install.packages("pacman")
pacman::p_load(
  # Specialty
  itsadug,
  
  # Format
  latex2exp, table1, gtsummary,
  broom.helpers, broom.mixed,
  
  # Modeling
  mgcv,
  
  # Useful extensions
  gtools, readxl, data.table, biglm, reshape2,
  
  # Basics
  DescTools, tidyverse, magrittr, janitor, glue, skimr,
  knitr, kableExtra, ggpubr, ggpmisc
)

# Make it easier to turn answers into rounded text 
PrintNum <- function(x, acc = 0.01, percent = F, vec = F){
  if(percent){ 
    y <- scales::label_percent(accuracy = acc)(x)
  } else {
    y <- scales::label_number(accuracy = acc)(x)
  }
  if(vec)( y <- paste0("(",
                       paste(y, collapse = ", "),
                       ")"))
  return(y)
}
```

# Overview

This document consolidates our work on a Bayesian hierarchical model using a soap-film smoother to describe the patterns of volume abnormalities in children with craniosynostosis.

## Questions

* 7 patients in the data have *sphenoidal involvement* marked as 1. What effect might this have on volumes?
* 18 patients have *previous surgery* marked as 1. Is this specifically indicating a previous surgery related to craniosynostosis?
* Should we control for image type in later analyses?


# Covariates Data

Data were sent by collaborators at Colorado (primarily Antonio Porras, Alejandro Cruz Guerrero, and Connor Elkhill).

Covariates are stored in a "merged_covariates" csv file that Josh Lukemire created from these data. Each image's malformation map is stored as a $500 \times 500$ matrix in its own RData file saved in a data folder in Josh's OneDrive.

We clean the covariate data set for easier analysis and to identify rare subsets of the data that should be excluded from our early work.

```{r}
#| label: load data

# Set directory for references
here::i_am("2025.08_Malformation_Analysis.qmd")

# Set data folder
DATA_FOLDER <- here::here("../../data/Lukemire, Joshua's files - data/")

# Set subfolder with image data
MATRIX_FOLDER <- file.path(DATA_FOLDER, "unsigned_distances")

# List all files with any extension (so no folders)
matrix_files <- list.files(MATRIX_FOLDER, pattern = "\\.")

# Read in covariates data
cranio_raw <- read_csv(file.path(DATA_FOLDER, "merged_covariates.csv"))

# Establish constants

MAX_AGE <- 365
```

First, we ensure that all images in the covariate data are represented in the data folder and vice versa.

```{r}
#| label: clean file names

### Clean cranio covariates file

## Create new data set with easier names
cranio <- cranio_raw %>% 
  clean_names()

## Check if files in data were found in data folder
cranio %<>%
  mutate(
    fname = paste0(id_fmt, "_desc-unsigneddist.RData"),
    fname_exists = if_else(fname %in% matrix_files, 1, 0)
  )

# Print a table to check we have all files
fname_summ <- tibble(
  in_data_not_folder = sum(1 - cranio$fname_exists),
  in_folder_not_data = sum(!matrix_files %in% cranio$fname)
)

kbl(fname_summ, caption = "Check Files vs. Data Availability") %>% 
  kable_styling(full_width = F)

```

## Suture Fusion Categories

Next, we create a cleaner categorization of each image's suture fusions. Information about the suture fusions for each image are initially stored in a series of binary columns: *metopic_fused, sagittal_fused, left_coronal_fused, right_coronal_fused, left_lambdoid_fused, right_lambdoid_fused*. We combine these into one column and create categories for "Bicoronal" and "Multiple."

```{r}

#| label: clean sutures

## Classify sutures

# Currently stored in binary columns with names ending in "fused"
# Reshape data to identify all combinations of sutures
fusion_df <- cranio %>% 
  select(id_fmt, matches("fused")) %>% 
  pivot_longer(cols = -id_fmt) %>% 
  mutate(
    name = name %>% 
      str_remove("_fused") %>% 
      str_replace_all("_", " ") %>% 
      str_to_title()
  ) %>% 
  mutate(fusion = if_else(value == 1, name, NA_character_)) %>% 
  group_by(id_fmt) %>% 
  summarize(num_fusions = sum(value, na.rm = T),
            fusions = paste(na.omit(fusion), collapse = "/")) %>% 
  ungroup()

# (1) count number of sutures
# (2) identify bicoronal (the only useful instance of multiple sutures)
# (3) identify other instances of multiple sutures
# (3) assign remaining suture types
cranio %<>%
  left_join(fusion_df) %>% 
  mutate(fusion_type = case_when(
    num_fusions == 0 ~ "None",
    fusions == "Left Coronal/Right Coronal" ~ "Bicoronal",
    num_fusions > 1 ~ "Multiple",
    T ~ fusions
  )) %>% 
  mutate(fusion_type = fct_infreq(fusion_type))

# Summarize frequencies of each suture type
fusion_summ <- cranio %>% 
  group_by(fusion_type) %>% 
  summarize(pics = n(),
            patients = n_distinct(id_base)) %>% 
  arrange(desc(pics)) %>% 
  mutate(pics_pct = pics / sum(pics),
         patients_pct = patients / sum(patients)) %>% 
  adorn_totals("row")

kbl(fusion_summ, caption = "Suture Fusions: Full Data") %>% 
  kable_styling(full_width = F)
```

## Longitudinal Data

Next, we look at the availability of longitudinal data. No longitudinal data exist for patients with no suture fusions. Patients with suture fusions have an average of about 1.5 visits per person, and this average is fairly consistent even when we limit to the number of visits taken before the 1 year mark.

```{r}
#| label: visits

# Identify which visit each image is from and how many visits they had
cranio %<>%
  arrange(id_base, age) %>% 
  group_by(id_base) %>% 
  mutate(visit_num = 1:n()) %>% 
  # Count visits total and visits under age limit (currently 1 year)
  mutate(visits = max(visit_num),
         visits_max_age = max(visit_num[age <= MAX_AGE], na.rm = T)) %>% 
  ungroup()

# Summarize visits on average and max by fusion group
visits_summ <- cranio %>%
  group_by(fusion_type) %>% 
  summarize(patients = n_distinct(id_base),
            pics = n(),
            visits_avg = mean(visits),
            visits_max = max(visits),
            visits_avg_max_age = mean(if_else(is.infinite(visits_max_age),
                                              NA_real_,
                                              visits_max_age), 
                                      na.rm = T),
            visits_max_max_age = max(visits_max_age, na.rm = T)) %>% 
  arrange(desc(visits_avg))
  
kbl(visits_summ, caption = "Longitudinal Visits: Full Data") %>% 
  kable_styling(full_width = F)
```

## Exclusion Criteria

Next, we consider exclusion criteria.

First, there are a small number of patients who were labeled by the collaborators as "Craniosynostosis," but have no suture fusions. Close inspection reveals that these are 6 of the 7 patients in the data with "sphenoidal involvement." Therefore, we exclude these 7 images.

Second, it is worth investigating the category of "Multiple" fusions. We find that other than the "bicoronal" combination of fusions (left coronal and right coronal fusions but nothing else), no other combination has a sample size of more than 14. Therefore, we exclude 81 patients with "Multiple" fusions.

Third, we likewise find that *Right Lambdoid* and *Left Lambdoid* suture fusions are very rare in our data, with 10 and 7 images respectively. Therefore, we exclude lambdoid suture fusions.

Fourth, we find that 18 images in the data are associated with a "previous surgery." Since we are more interested in the patterns associated with suture fusions before surgical intervention, we exclude these images. Notably, 6 of these 18 images have multiple suture fusions, so they would already be excluded.

```{r}

#| label: exclusion criteria

## Investigate any issues with fusion type
# Sanity check: fusion category against collaborators' "group" column
fusion_group_summ <- tabyl(cranio, fusion_type, group)

kbl(fusion_group_summ, caption = "Collaborators' Group vs. Fusions") %>% 
  kable_styling(full_width = F)

# Check the 6 images in "Craniosynostosis" group with no sutures
cranio_none <- cranio %>% 
  filter(group == "Craniosynostosis" & fusion_type == "None")

# These children all have sphenoidal involvement
cranio_sphen <- cranio %>% 
  group_by(fusion_type, group, sphenoidal_involvement) %>% 
  summarize(n = n()) %>% 
  ungroup() %>% 
  arrange(desc(n))

kbl(cranio_sphen, caption = "Spehnoidal Involvement Summary") %>% 
  kable_styling(full_width = F)

## Investigate multiple sutures

fusion_multiple_summ <- cranio %>% 
  filter(fusion_type == "Multiple") %>% 
  group_by(num_fusions, fusions) %>% 
  summarize(pics = n(),
            patients = n_distinct(id_base)) %>% 
  ungroup() %>% 
  arrange(desc(pics))

kbl(fusion_multiple_summ, caption = "Summary of Multiple Fusions") %>% 
  kable_styling(full_width = F)

max_multiple_count <- max(fusion_multiple_summ$pics)

## Investigate previous surgery
cranio_surg <- cranio %>% 
  filter(previous_surgery == 1) %>% 
  group_by(fusion_type, group, sphenoidal_involvement) %>% 
  summarize(n = n()) %>% 
  ungroup() %>% 
  arrange(desc(n)) %>% 
  adorn_totals("row")

kbl(cranio_surg, caption = "Images with Previous Surgery") %>% 
  kable_styling(full_width = F)

### Create subset data
# Exclude rare fusion categories: multiple, lambdoid
# Exclude sphenoidal involvement (in Craniosynostosis group without suture)
# Exclude patients with previous surgery (10)
# Only keep last picture taken
cranio_sub <- cranio %>% 
  filter(!fusion_type %in% c("Multiple", 
                             "Left Lambdoid", "Right Lambdoid") &
           sphenoidal_involvement %in% c(NA, 0) &
           previous_surgery %in% c(NA, 0) &
           visit_num == visits) %>% 
  # Reduce number of columns
  select(id_base, fname, fusion_type,
         age, sex) %>% 
  arrange(id_base, age)
```

We generate a summary of the fusion types present in the remaining observations, to compare against the full data.

```{r}
#| label: summarize cleaned data

fusion_summ_sub <- cranio_sub %>% 
  group_by(fusion_type) %>% 
  summarize(pics = n(),
            patients = n_distinct(id_base)) %>% 
  arrange(desc(pics)) %>% 
  mutate(pics_pct = pics / sum(pics),
         patients_pct = patients / sum(patients)) %>% 
  adorn_totals("row")

kbl(fusion_summ_sub, caption = "Fusions in Subset Data") %>% 
  kable_styling(full_width = F)
```

# Pointwise Difference Matrix Data

Next, we load the malformation maps into one data set. Each malformation map is stored as a $500 \times 500$ matrix of pointwise absolute value differences, and is saved in its own RData file. We load each RData file in the remaining subset of images selected above. Then, we reshape the data from wide to long, so that each row contains one pixel's row, column, and pointwise difference value. We then store this long data as a tibble nested in the "data" column. Fnially, we rearrange these data so that we can work with a data set with one row for each pixel, containing all participants' pointwise differences in that pixel.

```{r}
#| label: load matrix files

# Clear memory space
rm(list = setdiff(ls(), c("cranio_sub", "DATA_FOLDER", "MATRIX_FOLDER"))); gc()

# Define function to read each file
load_matrix_data <- function(x){
  
  # Path to data file
  full_path <- file.path(MATRIX_FOLDER, x)
  
  # Load RData file (pointwise_diff)
  load(full_path)
  
  # Convert matrix to long format and remove values outside mask
  reduced_data <- pointwise_diff %>%
    reshape2::melt(varnames = c("row", "col"), value.name = "diff") %>%
    as_tibble() %>%
    drop_na(diff)
  
  return(list(reduced_data))
}

# Create minimal matrix data
cranio_matrix <- cranio_sub %>%
  select(fname, age, sex, fusion_type) %>% 
  rowwise() %>%
  mutate(
    raw_data = load_matrix_data(fname)
  ) %>%
  # Rearrange so each row is a pixel
  unnest(raw_data) %>%
  group_by(row, col) %>%
  nest()
```

## Mask Shape and Region Labels

Next, to reduce the size of the data, we merge in Josh's brain mask shape. We also add each pixel's region label to associate this analysis with previous analyses.

```{r}
#|label: regions and mask

# Load regions and mask
load(paste0(DATA_FOLDER, "region_labels_and_mask.RData"))

cranio_matrix %<>%
  left_join(region_labels,
            by = c("row" = "X", "col" = "Y")) %>% 
  filter(value != 0) %>% 
  rename(region = value)
```

## Pointwise Difference Visualizations

As a first visualization, we plot the regions by pixel. 

```{r}
#| label: plot regions

# Create dataset of only region assignment for each pixel
matrix_region <- cranio_matrix %>% select(row, col, region) %>% 
  distinct()

# Identify regions' centroids for labels
matrix_region_center <- matrix_region %>% 
  group_by(region) %>% 
  summarize(row_center = mean(row),
            col_center = mean(col)) %>% 
  ungroup()

# Plot regions
# Note: We need to reverse the row and column indices to match the overhead view
plot_regions <- ggplot() +
  geom_raster(data = matrix_region, 
              aes(x = -row, y = -col, fill = factor(region))) +
  geom_text(data = matrix_region_center, 
            aes(x = -row_center, y = -col_center, label = factor(region))) +
  coord_fixed() +
  theme_void() +
  scale_fill_brewer(palette = "Set3", guide = "none") +
  labs(title = "Region Map", fill = "Region")

plot_regions
```


Next, we summarize the average differences by suture fusion and plot these averages on the brain shape.

```{r}
#| label: map fusions

# Clear memory space
rm(list = setdiff(ls(), c("cranio_sub", "cranio_matrix",
                          "DATA_FOLDER", "MATRIX_FOLDER"))); gc()

### Summarize differences by fusions

# Create dataset of average cell for each fusion
matrix_fusion <- cranio_matrix %>% 
  # Extract only fusion type and outcome from nested data
  hoist(data, "fusion_type", "diff") %>% 
  select(-c(data, region)) %>% 
  unnest_longer(col = c(fusion_type, diff)) %>% 
  # Summarize average by pixel and fusion type
  group_by(row, col, fusion_type) %>% 
  summarize(diff_avg = mean(diff)) %>% 
  ungroup()

# Plot brain map by fusion type
plot_fusion <- ggplot(matrix_fusion) +
  geom_raster(aes(x = -row, y = -col, fill = diff_avg)) +
  facet_wrap(~fusion_type) +
  coord_fixed() +
  theme_void() +
  scale_fill_viridis_c() +
  labs(title = "Fusion Average Differences")

# Print fusion maps
plot_fusion
```

We add to this summary by showing how these averages change on average over time, dividing age into 6 bins of roughly 2-month windows of ages. We also compare these results to the sample sizes in each of these age-fusion bins.

```{r}
#| label: plot fusion and age

### Investigate age

# Define age category cutoffs
# First, find quantiles of ages in craniosynostosis patients
age_cranio_quant <- cranio_sub %>% 
  filter(fusion_type != "None") %$% 
  quantile(age, prob = seq(0, 1, 0.2))

# Next, use closest interpretable number of days
age_breaks <- c(0, 61, 122, 183, 366, 1830, 3663)

# Label these with spaces so they can wrap in plots
age_break_labels <- map_chr(2:length(age_breaks), 
                            ~paste0(age_breaks[.-1], 
                                    " - ", 
                                    if_else(. == length(age_breaks), "", "<"), 
                                    age_breaks[.]))

# Create dataset with fusion and age info
matrix_age_fusion <- cranio_matrix %>% 
  hoist(data, "fusion_type", "age", "diff") %>% 
  select(-c(data, region)) %>% 
  unnest_longer(col = c(fusion_type, age, diff)) %>% 
  # Add age categories
  mutate(age_bin = cut(age, 
                       breaks = age_breaks, labels = age_break_labels, 
                       right = FALSE, include.lowest = T)) %>% 
  # Summarize average in these bins
  group_by(row, col, fusion_type, age_bin) %>% 
  summarize(diff_avg = mean(diff)) %>% 
  ungroup()

# Plot brain map by fusion type and age
plot_fusion_age <- ggplot(matrix_age_fusion) +
  geom_raster(aes(x = -row, y = -col, fill = diff_avg)) +
  facet_grid(fusion_type ~ age_bin,
             labeller = label_wrap_gen(width = 10)) +
  coord_fixed() +
  scale_fill_viridis_c() +
  theme_void() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "Fusion - Age Average Differences")

# Print fusion maps
plot_fusion_age

## Determine sample size for these averages

# Count images per fusion and age bin
size_fusion_age <- cranio_sub %>% 
  mutate(age_bin = cut(age, 
                       breaks = age_breaks, labels = age_break_labels, 
                       right = FALSE, include.lowest = T)) %>% 
  group_by(fusion_type, age_bin) %>% 
  summarize(n = n()) %>% 
  ungroup() %>% 
  pivot_wider(names_from = age_bin, values_from = n, id_cols = fusion_type)

# Print table
kbl(size_fusion_age, caption = "Fusion-Age Sample Size") %>% 
  kable_styling(full_width = F)
```

In order to gain a better understanding of the specific shape of the age effect, we plot average pointwise differences by age, stratified by region and fusion type.

```{r}
#| label: age effect

# Check shape of age effect by region and fusion

# Summarize by age, region, and fusion type; have option to round to nearest month
matrix_age_region <- cranio_matrix %>% 
  hoist(data, "fusion_type", "age", "diff") %>% 
  select(-data) %>% 
  unnest_longer(col = c(fusion_type, age, diff)) %>% 
  mutate(age_round_month = round(age / 30)) %>% 
  group_by(fusion_type, age, region) %>% 
  summarize(diff_avg = mean(diff)) %>% 
  ungroup()

plot_age <- ggplot(matrix_age_region) +
  # Plot 1 line over age for each region-fusion combo
  geom_line(aes(x = age, y = diff_avg, 
                color = fusion_type),
            alpha = 0.4) +
  facet_wrap(~region)

# Print plot
plot_age
```

Finally, we consider how each fusion type might differ between the two natal sexes recorded in the data.

```{r}
#| label: fusion by sex

# Take average pointwise differences in cell by fusion and sex
matrix_sex_fusion <- cranio_matrix %>% 
  hoist(data, "fusion_type", "sex", "diff") %>% 
  select(-c(data, region)) %>% 
  unnest_longer(col = c(fusion_type, sex, diff)) %>% 
  # Summarize average in these bins
  group_by(row, col, fusion_type, sex) %>% 
  summarize(diff_avg = mean(diff)) %>% 
  ungroup()

# Plot brain map by fusion type and age
plot_fusion_sex <- ggplot(matrix_sex_fusion) +
  geom_raster(aes(x = -row, y = -col, fill = diff_avg)) +
  facet_grid(fusion_type ~ sex,
             labeller = label_wrap_gen(width = 9)) +
  coord_fixed() +
  scale_fill_viridis_c() +
  theme_void() + 
  theme(axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "Fusion-Sex Average Differences")

# Print fusion maps
plot_fusion_sex
```

# Preliminary Models

Next, we get a better understanding of how these demographics interact with fusion type through preliminary regression models. First, we regress the region-level average pointwise difference for each subject on fusion indicators, age, and sex. Based on the plot above, we include age as oen linear covariate as well as $log(age + 1)$.

```{r}
#| label: region model

# Clear memory space
rm(list = setdiff(ls(), c("cranio_sub", "cranio_matrix",
                          "DATA_FOLDER", "MATRIX_FOLDER"))); gc()

# Add fusion variables for regression
fusion_dict <- tibble(
  fusion_type = c("None", 
                  "Sagittal", "Metopic", 
                  "Right Coronal", "Bicoronal", "Left Coronal"),
  fused_Sagittal = c(0, 1, 0, 0, 0, 0),
  fused_Metopic = c(0, 0, 1, 0, 0, 0),
  fused_RCoronal = c(0, 0, 0, 1, 1, 0),
  fused_LCoronal = c(0, 0, 0, 0, 1, 1)
)

## Run models for average pointwise differences in each region
cranio_region_models <- cranio_matrix %>% 
  # Rearrange to incorporate subject data
  unnest(data) %>% 
  # Take average in each region for each image
  group_by(fname, age, sex, fusion_type, region) %>% 
  summarize(diff_avg = mean(diff)) %>% 
  ungroup() %>%
  # Add fusion indicators
  left_join(fusion_dict) %>% 
  # Nest region-specific data
  group_by(region) %>% 
  nest() %>% 
  # Add linear model with no intercept
  mutate(model = map(data,
                     ~lm(diff_avg ~ 0 + 
                           fused_Sagittal + fused_Metopic +
                           fused_RCoronal * fused_LCoronal +
                           age + I(log(age + 1)) + sex,
                         data = .x)),
         # Also try linear model with intercept
         model_intercept = map(data,
                               ~lm(diff_avg ~ 1 + 
                                     fused_Sagittal + fused_Metopic +
                                     fused_RCoronal * fused_LCoronal +
                                     age + I(log(age + 1)) + sex,
                                   data = .x))) %>% 
  ## Add bicoronal variance to final coefficients; output coefficient summary
  mutate(coeffs = map(model,
                      function(m){
                        
                        # Make constrast vector identifying all coronal variables
                        is_coronal <- as.numeric(str_detect(colnames(vcov(m)), "Coronal"))                        
                        
                        # Compute variance of bicoronal combination
                        bicoronal_v <- t(is_coronal) %*% vcov(m) %*% is_coronal
                        bicoronal_e <- t(is_coronal) %*% coef(m)
                        
                        # Add bicoronal as row in tidy summary
                        bicoronal_row <- tibble(
                          term = "fused_Bicoronal",
                          estimate = bicoronal_e,
                          std.error = sqrt(bicoronal_v)) %>% 
                          mutate(p.value = 2 * pnorm(abs(estimate / std.error), lower.tail = FALSE))
                        
                        # Output coefficient table (smaller than model)
                        return(tidy(m) %>% bind_rows(bicoronal_row))
                      }),
         ## Repeat with intercept added
         coeffs_intercept = map(model_intercept,
                                function(m){
                                  
                                  # Make constrast vector identifying all coronal variables
                                  is_coronal <- as.numeric(str_detect(colnames(vcov(m)), "Coronal"))                        
                                  
                                  # Compute variance of bicoronal combination
                                  bicoronal_v <- t(is_coronal) %*% vcov(m) %*% is_coronal
                                  bicoronal_e <- t(is_coronal) %*% coef(m)
                                  
                                  # Add bicoronal as row in tidy summary
                                  bicoronal_row <- tibble(
                                    term = "fused_Bicoronal",
                                    estimate = bicoronal_e,
                                    std.error = sqrt(bicoronal_v)) %>% 
                                    mutate(p.value = 2 * pnorm(abs(estimate / std.error), lower.tail = FALSE))
                                  
                                  # Output coefficient table (smaller than model)
                                  return(tidy(m) %>% bind_rows(bicoronal_row))
                                })
  )

# Combine region-level models and add region column
model_region_summ <- bind_rows(cranio_region_models$coeffs, .id = "region") %>% 
  # Remove raw bicoronal interaction term (replaced with Bicoronal combo)
  filter(term != "fused_RCoronal:fused_LCoronal") %>% 
  arrange(region, -str_detect(term, "fused")) %>% 
  mutate(region = as.numeric(region),
         term = fct_inorder(term)) %>% 
  data.table()

# Create same summary for model with an intercept
model_region_summ_intercept <- bind_rows(cranio_region_models$coeffs_intercept, .id = "region") %>% 
  # Remove raw bicoronal interaction term (replaced with Bicoronal combo)
  filter(term != "fused_RCoronal:fused_LCoronal") %>% 
  arrange(region, -str_detect(term, "fused")) %>% 
  mutate(region = as.numeric(region),
         term = fct_inorder(term)) %>% 
  data.table()

# Recreate pixel map with region labels
matrix_region <- cranio_matrix %>% 
  select(row, col, region) %>% 
  distinct()

# Identify regions' centroids for labels
matrix_region_center <- matrix_region %>% 
  group_by(region) %>% 
  summarize(row_center = mean(row),
            col_center = mean(col)) %>% 
  ungroup()

# Merge pixels to region-level coefficients
model_region_matrix <- matrix_region %>% 
  left_join(model_region_summ, relationship = "many-to-many")

model_region_matrix_intercept <- matrix_region %>% 
  left_join(model_region_summ_intercept, relationship = "many-to-many")

# Plot
plot_reg_region <- ggplot(model_region_matrix) +
  geom_raster(aes(x = -row, y = -col, fill = estimate)) +
  geom_text(data = matrix_region_center, 
            aes(x = -row_center, y = -col_center, label = region)) +
  facet_wrap(~term) +
  coord_fixed() +
  scale_fill_viridis_c(option = "turbo") +
  theme_void() +
  labs(title = "Region-Level No-Intercept Linear Model",
       fill = "Coefficient")

# Plot intercept version of the model
plot_reg_region_intercept <- ggplot(model_region_matrix_intercept) +
  geom_raster(aes(x = -row, y = -col, fill = estimate)) +
  geom_text(data = matrix_region_center, 
            aes(x = -row_center, y = -col_center, label = region)) +
  facet_wrap(~term) +
  coord_fixed() +
  scale_fill_viridis_c(option = "turbo") +
  theme_void() +
  labs(title = "Region-Level Intercept Linear Model",
       fill = "Coefficient")

plot_reg_region

plot_reg_region_intercept
```

To get a more detailed spatial perspective, we then run the same regression model on each individual pixel in the data. To save time, we do not calculate the variance estimate of these parameters; we only estimate the coefficients.

```{r}
#| label: pixel model

# Clear memory space
rm(list = setdiff(ls(), c("cranio_sub", "cranio_matrix", "fusion_dict",
                          "DATA_FOLDER", "MATRIX_FOLDER"))); gc()

## Run models for pointwise differences in each pixel
cranio_pixel_models <- cranio_matrix %>% 
  mutate(
    model = map(data,
                function(d){
                  
                  data_ind <- d %>% 
                    left_join(fusion_dict, by = "fusion_type")
                  
                  data_mx <- model.matrix(~ 0 + 
                                fused_Sagittal + fused_Metopic +
                                fused_RCoronal * fused_LCoronal +
                                age + I(log(age + 1)) + sex,
                                data_ind)
                  
                  model <- lm.fit(y = d$diff, x = data_mx)
                  
                  return(model)
                }
    )) %>% 
  ## Add bicoronal variance to final coefficients; output coefficient summary
  mutate(coeffs = map(model,
                      function(m){
                        
                        # Make constrast vector identifying all coronal variables
                        is_coronal <- as.numeric(str_detect(names(coef(m)), "Coronal"))                        
                        
                        # Compute estimate of bicoronal combination
                        bicoronal_e <- t(is_coronal) %*% coef(m)
                        
                        # Output coefficient table (smaller than model)
                        return(c(coef(m), "fused_Bicoronal" = bicoronal_e))
                      })
  )

# Combine pixel-level models and add region column
model_pixel_summ <- cranio_pixel_models %>% 
  select(row, col, coeffs) %>% 
  mutate(coeffs = map(coeffs, ~enframe(., name = "term", value = "estimate"))) %>% 
  unnest(coeffs) %>% 
  # Remove raw bicoronal interaction term (replaced with Bicoronal combo)
  filter(term != "fused_RCoronal:fused_LCoronal") %>% 
  arrange(row, col, -str_detect(term, "fused")) %>% 
  mutate(term = fct_inorder(term))

# Plot
plot_reg_pixel <- ggplot(model_pixel_summ) +
  geom_raster(aes(x = -row, y = -col, fill = estimate)) +
  facet_wrap(~term) +
  coord_fixed() +
  scale_fill_viridis_c(option = "turbo") +
  theme_void() +
  labs(title = "Pixel-Level Linear Model",
       fill = "Coefficient")

plot_reg_pixel
```

Next, for a preliminary spatial model, we split the data only by fusion type. Based on the plots above, the age and sex effects are basically uniform across all of the pixels, so we include these as linear terms that do not vary spatially. Otherwise, we only include a tensor spline interaction between row and column indices. Our model does not reflect the brain boundary. It is a generalized additive model with an intercept and a tensor product smooth of row and column indices.

```{r}
#| label: preliminary gam

# Clear memory space
rm(list = setdiff(ls(all.names = TRUE), 
                  c("cranio_sub", "cranio_matrix", "fusion_dict",
                    "DATA_FOLDER", "MATRIX_FOLDER"))); gc()

cranio_fusion_models <- cranio_matrix %>% 
  unnest(data) %>% 
  filter(!fusion_type %in% c("None")) %>% 
  group_by(fusion_type) %>% 
  nest() %>% 
  mutate(model = map(data,
                     ~bam(diff ~ age + I(log(age + 1)) + sex + te(row, col),
                     data= .x))) %>% 
  mutate(pred_data = map(
    model,
    ~get_predictions(.x,
                     cond = list(
                       "age" = median(cranio_sub$age),
                       "sex" = 0,
                       "row" = seq(from = min(cranio_matrix$row), 
                                   to = max(cranio_matrix$row), 
                                   length.out = 50),
                       "col" = seq(from = min(cranio_matrix$col), 
                                   to = max(cranio_matrix$col), 
                                   length.out = 50))
    )))

cranio_fusion_model_summ <- cranio_fusion_models %>% 
  select(fusion_type, pred_data) %>% 
  unnest(pred_data)

# Plot these predictions (in a rectangle shape)
plot_model_fusion <- ggplot(cranio_fusion_model_summ) +
  geom_raster(aes(x = -row, y = -col, fill = fit)) +
  facet_wrap(~fusion_type) +
  coord_fixed() +
  scale_fill_viridis_c(option = "turbo") +
  theme_void() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(title = "Fusion-Specific GAM",
       fill = "Pred. Diff.")

plot_model_fusion
```

