---
title: "Malformation Analysis"
author: "Ryan Taylor"
format: 
  html:
    embed-resources: true
    theme: paper
    highlight: haddock
    code-fold: true
    toc: true
    toc-location: left
    code-overflow: wrap
    df-print: kable
    knitr:
      opts_chunk:
        echo: true
        include: true
        message: false
        warning: false
        results: markup
        fig.align: center
        fig.height: 5
        fig.width: 7
        R.options:
          scipen: 999
          knitr.kable.NA: ""
number-sections: true
editor: source
---

::: {.hidden}
$$
\usepackage{xcolor}

\newcommand{\bX}{\mathbf X}
\newcommand{\bY}{\mathbf Y}
\newcommand{\by}{\mathbf y}
\newcommand{\bH}{\mathbf H}
\newcommand{\bh}{\mathbf h}
\newcommand{\bx}{\mathbf x}
\newcommand{\bG}{\mathbf G}
\newcommand{\bg}{\mathbf g}
\newcommand{\bS}{\mathbf S}
\newcommand{\bs}{\mathbf s}
\newcommand{\bP}{\mathbf P}
\newcommand{\bC}{\mathbf C}
\newcommand{\bI}{\mathbf I}
\newcommand{\hatbG}{\widehat \bG}
\newcommand{\ba}{\mathbf a}
\newcommand{\bz}{\mathbf z}
\newcommand{\bb}{\mathbf b}
\newcommand{\bB}{\mathbf B}
\newcommand{\bw}{\mathbf w}
\newcommand{\bW}{\mathbf W}
\newcommand{\bZ}{\mathbf Z}
\newcommand{\bV}{\mathbf V}
\newcommand{\bU}{\mathbf U}

\newcommand{\balpha}{\boldsymbol{\alpha}}
\newcommand{\bbeta}{\boldsymbol{\beta}}
\newcommand{\bgamma}{\boldsymbol{\gamma}}
\newcommand{\bepsilon}{\boldsymbol{\epsilon}}
\newcommand{\bphi}{\boldsymbol{\phi}}
\newcommand{\hatbbeta}{\hat{\bbeta}}
\newcommand{\hatbgamma}{\hat{\bgamma}}
\newcommand{\bSigma}{\boldsymbol{\Sigma}}
\newcommand{\bsigma}{\boldsymbol{\sigma}}
\newcommand{\hatVar}{\widehat{\Var}}
\newcommand{\hatCov}{\widehat{\Cov}}
\newcommand{\bLambda}{\boldsymbol{\Lambda}}
\newcommand{\hatbLambda}{\hat{\bLambda}}
\newcommand{\blambda}{\boldsymbol{\lambda}}
\newcommand{\hatblambda}{\hat{\blambda}}
\newcommand{\btheta}{\boldsymbol{\theta}}
\newcommand{\hatbSigma}{\widehat{\boldsymbol{\Sigma}}}
\newcommand{\bmu}{\boldsymbol{\mu}}
\newcommand{\bzeta}{\boldsymbol{\zeta}}
\newcommand{\bomega}{\boldsymbol{\omega}}
\newcommand{\bOmega}{\boldsymbol{\Omega}}

\newcommand{\diag}{\mathrm{diag}}
\newcommand{\tr}{\mathrm{tr}}

\newcommand{\bzero}{\boldsymbol{0}}
\newcommand{\bone}{\boldsymbol{1}}
\def\iid{\,{\buildrel iid \over \sim}\,} 
\newcommand{\reals}{\mathbb{R}}
$$
:::

```{r}
#| label: setup
#| include: false

# Clear memory
rm(list = ls()); gc(verbose = F)

# Load packages
# if(!require("pacman")) install.packages("pacman")
pacman::p_load(
  # Specialty
  itsadug, spdep, concaveman,
  
  # Bayes
  mvtnorm,
  
  # Format
  latex2exp, table1, gtsummary,
  broom.helpers, broom.mixed, colorspace,
  
  # Modeling
  mgcv,
  
  # Useful extensions
  gtools, readxl, reshape2, ggridges, Matrix, MASS, ggh4x,
  
  # Basics
  DescTools, tidyverse, magrittr, janitor, glue, skimr,
  knitr, kableExtra, ggpubr, ggpmisc, here
)

# Load my Rcpp functions
library(DevCranio)

# Load my R functions
source(here("source", "load_matrix_data_fn.R"))
source(here("source", "obtain_predictions_fn.R"))
source(here("source", "lm_penalized_gibbs_fn.R"))
source(here("source", "construct_smooth_ref_code_fns.R"))
source(here("source", "hierarchical_penalized_gibbs_fn.R"))
source(here("source", "simulate_data_fn.R"))
source(here("source", "mv_normal_matrix_fn.R"))

# Set switches
DO_LOAD_MATRICES <- F
DO_FILTER_MATRICES <- F
DO_PRELIM_MODEL_REGION <- F
DO_PRELIM_MODEL_PIXEL <- F
DO_PRELIM_MODEL_FUSION <- F
DO_INDIV_SOAPS <- F
DO_JOINT_FIT <- F

FIT_LOAD_DATE <- "2026-04-21"

# Set a flag for what not to delete
RM_IGNORE <- "DO|IGNORE|FOLDER|cranio|_dict|so_|_DATE"
```

```{r}
#| label: load-data-prep

# Set data folder
DATA_FOLDER <- here("../data/Lukemire, Joshua's files - data/")
COVARS_FOLDER <- here("../data/SphericalMaps_database/SphericalMaps_database/")

# Set subfolder with image data
MATRIX_FOLDER <- file.path(DATA_FOLDER, "unsigned_distances")
```

```{r}
#| label: set-constants

# Define number of knots along each dimension of soap film
so_GRID_N <- 10

# Define number of knots along age effect
cranio_knots_age <- 10

# Define maximum age
cranio_max_age <- 365

# Define age category cutoffs
cranio_age_breaks <- c(seq(0, 300, 60), cranio_max_age)

# Label these with spaces so they can wrap in plots
cranio_age_break_labels <- map_chr(2:length(cranio_age_breaks),
                            ~paste0(cranio_age_breaks[.-1],
                                    " - ",
                                    if_else(. == length(cranio_age_breaks), "", "<"),
                                    cranio_age_breaks[.]))

# Use data set to merge in fusion variables for regressions
fusion_dict <- tibble(
  fusion_type = c("Normative", 
                  "Sagittal", "Metopic", 
                  "Right Coronal", "Bicoronal", "Left Coronal"),
  fused_Sagittal = c(0, 1, 0, 0, 0, 0),
  fused_Metopic = c(0, 0, 1, 0, 0, 0),
  fused_RCoronal = c(0, 0, 0, 1, 1, 0),
  fused_LCoronal = c(0, 0, 0, 0, 1, 1)
)
```


# Overview

This document consolidates our work on a Bayesian hierarchical model using a soap-film smoother to describe the patterns of volume abnormalities in children with craniosynostosis.

## Updates

- Limited analysis to under 1 year old
- Removed duplicates from subset data
- Added linear terms for fusion type into model
- Removed Kronecker products from Gibbs sampler function (except for update of subject-level covariate coefficients)
- Added results from preliminary joint model fit to the document
- Updated 2D tensor product models to use 8-dimension functions, comparable to the dimension of our model

## Questions

- Can we add a 2nd level of scaling terms to the subject-level coefficients? This would change the assumption from $[\bbeta_k | \tau^2, \lambda_\alpha] \iid \mathcal{N}_Q (\bzero, \tau^2 \bP_T^{(\beta) -})$ to $[\bbeta_k | \tau^2, \lambda_\alpha] \iid \mathcal{N}_Q (\bzero, \tau^2 * \text{diag}(\psi_1, \ldots, \psi_Q) * \bP_T^{(\beta) -})$. Josh and I had talked about setting $\psi_q = 1$ for everything except the linear fusion terms, and setting one $\psi_q$ for these three parameters.

## Collaborator Questions

-   7 patients in the data have *sphenoidal involvement* marked as 1. What effect might this have on volumes?
-   18 patients have *previous surgery* marked as 1. Is this specifically indicating a previous surgery related to craniosynostosis?
-   Should we control for image type in later analyses?
-   Looking at a subset of the data, we can see that the images "sub-CNH_0002_dol-1295" and "sub-CNH_Cranio_00002_dol-61" have the exact same pointwise differences, even though the first is marked as having no fusion and the second is marked as having a Sagittal fusion. Are these the same image?

# Covariates Data

Data were sent by collaborators at Colorado (primarily Antonio Porras, Alejandro Cruz Guerrero, and Connor Elkhill).

Covariates are stored in a "merged_covariates" csv file that Josh Lukemire created from these data. Each image's malformation map is stored as a $500 \times 500$ matrix in its own RData file saved in a data folder in Josh's OneDrive.

```{r}
#| label: load-covariate-data

source(here("source", "01_load_covariate_data.R"))
```


We clean the covariate data set for easier analysis. While doing this, we need to summarize and visualize the data to identify rare subsets that should be excluded from our early work.

```{r}
#| label: clean-covariate-data

source(here("source", "02_clean_covariate_data.R"))
```

First, we ensure that all images in the covariate data are represented in the data folder and vice versa.

```{r}
#| label: clean-file-names

# Summary table of files in data vs. folder
kbl(fname_summ, caption = "Check Files vs. Data Availability") %>% 
  kable_styling(full_width = F)
```

## Suture Fusion Categories

Next, we create a cleaner categorization of each image's suture fusions. Information about the suture fusions for each image are initially stored in a series of binary columns: *metopic_fused, sagittal_fused, left_coronal_fused, right_coronal_fused, left_lambdoid_fused, right_lambdoid_fused*. We combine these into one column and create categories for "Bicoronal" and "Multiple."

```{r}
#| label: clean-sutures

# Summary table of pictures by fusion type
kbl(fusion_summ, caption = "Suture Fusions: Full Data") %>% 
  kable_styling(full_width = F)
```

## Longitudinal Visits {#sec-longitudinal}

Next, we look at the availability of longitudinal data. No longitudinal data exist for patients with no suture fusions. Patients with suture fusions have an average of about 1.5 visits per person, and this average is fairly consistent even when we limit to the number of visits taken before the 1 year mark.

```{r}
#| label: visits
  
# Summary table of longitudinal visits by fusion type
kbl(visits_summ, caption = "Longitudinal Visits: Full Data") %>% 
  kable_styling(full_width = F)
```

## Exclusion Criteria {#sec-exclusion}

Next, we consider exclusion criteria.

First, there are a small number of patients who were labeled by the collaborators as "Craniosynostosis," but have no suture fusions. Close inspection reveals that these are 6 of the 7 patients in the data with "sphenoidal involvement." Therefore, we exclude these 7 images.

```{r}
#| label: group-and-sphenoidal

# Summary table of group labels by fusion type
kbl(fusion_group_summ, caption = "Collaborators' Group vs. Fusions") %>% 
  kable_styling(full_width = F)

# Summary table of fusion type, group, and sphenoidal involvement
kbl(sphen_summ, caption = "Sphenoidal Involvement Summary") %>% 
  kable_styling(full_width = F)
```

Second, it is worth investigating the category of "Multiple" fusions. We find that other than the "bicoronal" combination of fusions (left coronal and right coronal fusions but nothing else), no other combination has a sample size of more than 14. Therefore, we exclude 81 patients with "Multiple" fusions.

Third, we likewise find that *Right Lambdoid* and *Left Lambdoid* suture fusions are very rare in our data, with 10 and 7 images respectively. Therefore, we exclude lambdoid suture fusions.

```{r}
#| label: multiple-sutures

# Summary table of possible multiple fusions
kbl(fusion_multiple_summ, caption = "Summary of Multiple Fusions") %>% 
  kable_styling(full_width = F)
```

Fourth, we find that 18 images in the data are associated with a "previous surgery." Since we are more interested in the patterns associated with suture fusions before surgical intervention, we exclude these images. Notably, 6 of these 18 images have multiple suture fusions, so they would already be excluded.

```{r}
#| label: previous-surgery

# Summary table of previous surgery with other exclusion criteria
kbl(surg_summ, caption = "Images with Previous Surgery") %>% 
  kable_styling(full_width = F)
```

### Age Summary {#sec-age-summary}

Fifth, we check the distribution of age in our data. We determine that over 98\% of the data falls under the age of 5 years, so we exclude those above this age, since we have very sparse observations to estimate an age effect.

```{r}
#| label: age-summary

# Plot of age distribution by fusion
age_plot

# High age percentiles
kbl(have_fusions_quant, 
    caption = "Max. Age Options: Patients with Fusions") %>% 
  kable_styling(full_width = F)

# Details on maximum age we use
kbl(max_age_summ, caption = "Age cap summary") %>% kable_styling(full_width = F)
```

# Pointwise Difference Matrix Data {#sec-load-matrices}

Next, we load the malformation maps into one data set. Each malformation map is stored as a $500 \times 500$ matrix of pointwise absolute value differences, and is saved in its own RData file. We load each RData file in the remaining subset of images selected above. Then, we reshape the data from wide to long, so that each row contains one pixel's row, column, and pointwise difference value. We then store this long data as a tibble nested in the "data" column. Finally, we rearrange these data so that we can work with a data set with one row for each pixel, containing all participants' pointwise differences in that pixel. In addition, we merge each pixel's region label to this pointwise matrix data, so that we can associate this analysis with previous analyses.



```{r}
#| label: load-matrix-files

# Toggle whether to create matrix data

# NB: filtering matrix also involves identifying duplicates in covariates
if(DO_LOAD_MATRICES & DO_FILTER_MATRICES){
  
  source(here("source", "03_load_Matrix_data.R"))
  source(here("source", "04_filter_matrix_data.R"))
  
} else if(DO_FILTER_MATRICES){
  
  cranio_matrix <- readRDS(file = here::here("analysis", "intermediate", 
                                             "covars_and_matrices.rds")) 
  
  source(here("source", "04_filter_matrix_data.R"))
  
  cranio_matrix <- cranio_matrix_sub
  
} else {
  cranio_matrix <- readRDS(file = here::here("analysis", "intermediate", 
                                             "covars_and_matrices_sub.rds"))
  
  cranio_sub <- readRDS(file = here::here("analysis", "intermediate",
                                          "demographics_filtered.rds"))
}
```

### Data Subset {#sec-subset-data}

We generate a summary of the fusion types present in the remaining observations, to compare against the full data.

```{r}
#| label: summarize-cleaned-data

# Summary table of fusion types in final data
kbl(fusion_summ_sub, caption = "Fusions in Subset Data") %>% 
  kable_styling(full_width = F)
```

We also create a Table 1, to identify what representation we have across key demographics.

```{r}
#| label: table-1

# Table 1 for final data
table1(~age + sex | fusion_type,
       data = cranio_sub)
```


## Growth Matrix Visualizations {#sec-matrix-viz}

First, to understand the first way we can aggregate the data spatially, we plot the collaborators' defined regions by pixel.

```{r}
#| label: matrix-analysis

source(here("source", "05_analyze_matrix_data.R"))
```


### Region Map

```{r}
#| label: plot-regions

# Plot of region locations
plot_regions
```

### Fusion Averages

Next, we summarize the growth by suture fusion and plot these averages on the brain shape.

```{r}
#| label: map-fusions

# Plot pixel average by fusion
plot_fusion
```

### Fusion - Age Averages

We add to this summary by showing how these averages change on average over time, dividing age into 6 bins of roughly 2-month windows of ages. We also compare these results to the sample sizes in each of these age-fusion bins.

```{r}
#| label: plot-fusion-and-age

# Plot pixel average by fusion and age bucket
plot_fusion_age

# Table of sample size by fusion and age bucket
kbl(size_fusion_age, caption = "Fusion-Age Sample Size") %>% 
  kable_styling(full_width = F)
```

### Age Trajectory {#sec-age-smooth}

In order to gain a better understanding of the specific shape of the age effect, we plot average growth by age, stratified by region and fusion type.

```{r}
#| label: age-effect

# Plot average age effect by fusion type by region
plot_age
```

Finally, we consider how each fusion type might differ between the two natal sexes recorded in the data.

```{r}
#| label: fusion-by-sex

# Plot pixel average by sex by fusion type
plot_fusion_sex
```

# Preliminary Models {#sec-prelim-models}

## Region-Level Independent

Next, we get a better understanding of how these demographics interact with fusion type through preliminary regression models. First, we regress the region-level average growth for each subject on fusion indicators, age, and sex. Based on the plot above, we include age as a smooth term; all other covariates are included as indicator variables. We also run a multiple linear regression with age as a linear term.

```{r}
#| label: region-models

source(here("source", "06_prelim_models_region.R"))
```


We denote the average growth for patient $i$ in region $v$ as $G_{i}^{(v)}$. We let $f_v(\cdot)$ be a cubic regression spline estimated on observations from region $v$; we choose a cubic regression spline for faster estimation of many regressions, especially when we extend to a pixel-level model below. Then these models can be written:

$$
\begin{gather*}
\begin{split}
G_{i}^{(v)} = {} & \beta_{0}^{(v)} + \beta_{1}^{(v)} [\text{Sagittal}]_i + \beta_{2}^{(v)} [\text{Metopic}]_i + \\
& \beta_{3}^{(v)} [\text{Right Coronal}]_i + \beta_{4}^{(v)} [\text{Left Coronal}]_i + \\
& \beta_{5}^{(v)} [\text{Left and Right Coronal}]_i + \\
& \beta_{6}^{(v)} [\text{Sex: Male}]_i + f^{(v)}(Age_i) + \epsilon_{i}^{(v)}
\end{split} \\ \\
\begin{split}
G_{i}^{(v)} = {} & \beta_{0}^{(v)} + \beta_{1}^{(v)} [\text{Sagittal}]_i + \beta_{2}^{(v)} [\text{Metopic}]_i + \\
&\beta_{3}^{(v)} [\text{Right Coronal}]_i + \beta_{4}^{(v)} [\text{Left Coronal}]_i + \\
& \beta_{5}^{(v)} [\text{Left and Right Coronal}]_i + \\
& \beta_{6}^{(v)} [\text{Sex: Male}]_i + \beta_{7}^{(v)}(Age_i) + \epsilon_{i}^{(v)}
\end{split} \\ \\
\begin{split}
\epsilon_{i}^{(v)} \iid {} & N(0, \sigma^{2 \ (v)})
\end{split}
\end{gather*}
$$

```{r}
#| label: region-model

# Plot parametric coefficients from region-level model using linear age
plot_region_linear

# Plot parametric coefficients from region-level model using smooth of age
plot_region_smooth

# Plot region-level age smooth effects
plot_region_smooth_pred

```

## Pixel-Level Independent

To get a more detailed spatial perspective, we then run the same regression model on each individual pixel in the data.

```{r}
#| label: pixel-models

source(here("source", "07_prelim_models_pixel.R"))
```


We denote the growth for patient $i$ in matrix row $x$ and column $y$ as $G_{i}(x,y)$ and let $f^{(x,y)}(\cdot)$ be a cubic regression spline function estimated on observations at pixel $(x,y)$; then this model can be written:

$$
\begin{gather*}
\begin{split}
G_{i}^{(x,y)} = {} & \beta_{0}^{(x,y)} + \beta_{1}^{(x,y)} [\text{Sagittal}]_i + \beta_{2}^{(x,y)} [\text{Metopic}]_i + \\
& \beta_{3}^{(x,y)} [\text{Right Coronal}]_i + \beta_{4}^{(x,y)} [\text{Left Coronal}]_i + \\
& \beta_{5}^{(x,y)} [\text{Left and Right Coronal}]_i + \\
& \beta_{6}^{(x,y)} [\text{Sex: Male}]_i + f^{(x,y)}(Age_i) + \epsilon_{i}^{(x,y)}
\end{split} \\ \\
\epsilon_{i}^{(x,y)} \iid{} N(0, \sigma^{2 \ (x,y)})
\end{gather*}
$$

```{r}
#| label: pixel-model

# Plot pixel-level parametric coefficients
plot_pixel

# Plot pixel-level age smooth effects, colored by region
plot_pixel_pred
```

## Fusion-Independent GAM {#sec-prelim-gam}

Next, for a preliminary spatial model, we split the data only by fusion type. Based on the plots above, the age and sex effects are basically uniform across all of the pixels, so we include these as linear terms that do not vary spatially. Otherwise, we only include a tensor spline interaction between row and column indices. Our model does not reflect the brain boundary. It is a generalized additive model with an intercept and a tensor product smooth of row and column indices.

```{r}
#| label: gam-2d-models

# Clear memory space
rm(list = setdiff(ls(), c(lsf.str(), ls(pattern = eval(RM_IGNORE)))))
gc(verbose = F)

source(here("source", "08_prelim_models_2dGAM.R"))
```


We denote the growth for patient $i$ with suture fusion type $r$ in matrix row $x$ and column $y$ as $G_{ri}(x,y)$. Let $f_{r}(\cdot)$ be a cubic regression spline estimated on observations from patients with suture fusion $r$ and let $h_r(\cdot, \cdot)$ be a tensor product smooth with a cubic regression spline basis; then this model can be written:

$$
\begin{gather*}
\begin{split}
G_{ri}(x,y) = {} & \beta_{0, r} + \beta_{1,r} [\text{Sex: Male}]_i + \\
& f_{r}(Age_i) + h_r(x, y) + \epsilon_{ri}(x,y)
\end{split} \\ \\
\epsilon_{ri}(x,y) \iid{} N(0, \sigma^2_r).
\end{gather*}
$$

Predictions from this model are shown below. In order to show how this model is inadequate for our data, we compare these predictions to the boundary of the matrix data.

```{r}
#| label: preliminary-gam

# Plot 2D GAM predictions on rectangle with brain shape superimposed
plot_model_fusion

# Plot fusion-specific age effects
plot_fusion_pred

# Print table of parametric coefficients from these models
kbl(model_fusion_params, 
    caption = "Preliminary 2D GAM Parametric Coefficients") %>% 
  kable_styling(full_width = F)

```

# Individual Soap Film Models

Next, we introduce the soap film smoother. To fit this model, we first need to create the boundary that determines the extent of the model and the set of knots in $\mathbb{R}^2$ that will define the smoothing functions.

```{r}
#| results: hide

# Clear memory space
rm(list = setdiff(ls(), c(lsf.str(), ls(pattern = eval(RM_IGNORE)))))
gc(verbose = F)
```

```{r}
#| label: soap-film-setup

source(here("source", "09_setup_soap_film.R"))
```


## Matrix Boundary

To construct this boundary, we need to address two issues: first, we need to define the boundary such that none of our data overlap with it; and second, we need to order the points that make up the boundary so that the boundary can be constructed by simply connecting these points with line segments *in order*. Therefore, we construct this boundary in a few steps. First, we take the concave hull of all points in our matrix; this approach is not be generalizable to the case where there are multiple loops to account for in the soap film, but it works for our case, where we have one continuous boundary. Next, we add a buffer of 1 unit, so that none of our data will overlap with the boundary. We then convert this shape to a boundary line and "sample" points at 1-unit intervals, ensuring that we have even coverage of the resulting boundary shape. These points are passed as a list to the soap film function. The boundary resulting from this process is depicted below.

```{r}
#| label: cranio-boundary

# Plot boundary shape as a sanity check
bound_viz
```

## Soap Film Grid  {#sec-soap-film-grid}

We construct the boundary shape above. Here, we set up a $10 \times 10$ grid of knots and crop this grid to only the knots within the boundary. We visualize these knots to ensure that the process works: every knot included in our model is within the boundary, and every knot not included is outside of it.



```{r}
#| label: soap-infrastructure

# Plot grid of all knots and those used in the model (in the boundary)
grid_viz
```

## Smoother Object {#sec-soap-film-smoothcon}

Next, we use the set of row and column indices (shared among all images) to create a design matrix containing 2-dimensional basis functions that make up a smoother that can be used for every individual image. We use the `mgcv::smoothCon` function.

From this smoother object, we extract the relevant matrices that will allow us to fit the model. Let $M$ be the number of pixels in each image and $K$ be the number of basis functions evaluated on these pixels. Then we extract the $M \times K$ design matrix $\bB$, the $K_c \times K_c$ penalty matrix for the boundary spline ($\bP_c$) and the $K_f \times K_f$ penalty matrix for the interior spline ($\bP_f$). In `mgcv`, The smoother object stores each penalty matrix as a block-diagonal size $K \times K$ matrix, where $K = K_c + K_f$. We denote these $\bP^{(c)}$ and $\bP^{(f)}$:

$$
\begin{aligned}
\\
\bP^{(c)} &= \begin{bmatrix}
\bP_c & \smash{\overbrace{\bzero}^{K_c \times K_f}} \\
\smash{\underbrace{\bzero}_{K_f \times K_c}} & \smash{\underbrace{\bzero}_{K_f \times K_f}}
\end{bmatrix} \\ \\ \\ \\
\bP^{(f)} &= \begin{bmatrix}
\smash{\overbrace{\bzero}^{K_c \times K_c}} & \smash{\overbrace{\bzero}^{K_c \times K_f}} \\
\smash{\underbrace{\bzero}_{K_f \times K_c}} & \bP_f
\end{bmatrix}
\\ \\
\end{aligned}
$$

In order to isolate the effect of each separate spline, we also remove the block zero matrices to convert each penalty matrix into the original individual matrices $P_c$ and $P_f$.

To estimate coefficients for this model, we multiply these penalty matrices by smoothing parameters $\lambda_c$ and $\lambda_f$. We then use the "total penalty matrix"


$$
\bP_T = \lambda_c \bP^{(c)} + \lambda_f \bP^{(f)} = 
\begin{bmatrix}
\lambda_c \bP_c & \smash{\overbrace{\bzero}^{K_c \times K_f}} \\
\smash{\underbrace{\bzero}_{K_f \times K_c}} & \lambda_f \bP_f
\end{bmatrix}.
\\ \\
$$

### Soap Film Basis Functions {#sec-soap-basis-fns}

After constructing the basis, it is useful to visualize the basis functions to understand the effect each has on predictions.

```{r}
#| label: basis-fn-analysis

source(here("source", "10_visualize_bases.R"))
```


First, we identify the outermost points in our data, on which the boundary splines are estimated. These points are the "nearest neighbors" to the boundary points defined earlier. We keep the boundary points in the order in which they are defined in the boundary object, so that they can be connected by line segments to form the boundary. For each boundary point, we calculate the Euclidean distance from the previous point and take the cumulative sum of these distances to determine the distance around the boundary. We use this measure as the X axis to plot the basis that is used to estimate the boundary condition, i.e. the value of the outcome on the boundary.

Then, we plot the "boundary-induced" basis functions: these are the 2-dimensional functions that are defined over the entire matrix. Each boundary-induced function essentially forms the minimal surface over the area that respects its single corresponding boundary basis function. This is described in section 4.1 in the soap film paper (p. 940):

First, suppose values on the boundary follow a function $\bG_i^{(c)}(d) = \Delta_i^{(c)}(d) + \epsilon_i^{(c)}(d)$, where $d$ is the distance along the boundary loop from an arbitrary starting point. 

Second, represent that function with the basis expansion $\Delta_i^{(c)}(d) = \sum_{k=1}^{K_c} \gamma^{(c)}_{ij} \delta_j(d)$. To estimate $\bgamma^{(c)}_{i}$, include a penalty functional $\bgamma_{i}^{(c) \top} \bP_c \bgamma^{(c)}_{i}$.

These basis functions $\boldsymbol{\delta}(d)$ are plotted below:

```{r}
#| label: boundary-cyclic-basis-viz

# Plot cyclic spline defined on distance along boundary
soap_basis_cyclic
```

Third, for each $k = 1, \ldots, K_c$, determine the basis-specific boundary condition by setting $\gamma^{(c)}_k = 1$ and $\gamma^{(c)}_{k'} = 0, \ k' \neq k$. *I think this means all we're left with is* $\delta_k(d)$. Solve for the function $ \phi^{(c)}_k(x,y)$ which satisfies this boundary condition based on $\delta_k(d)$ and satisfies the equation $\frac{\partial^2 \phi^{(c)}_k}{\partial x^2} + \frac{\partial^2 \phi^{(c)}_k}{\partial y^2} = 0$. 

Then the "boundary-condition-induced-function" $\Phi_i^{(c)}(x,y)$ has the basis function representation $\Phi_i^{(c)}(x,y) = \sum_{k=1}^{K_c} \gamma^{(c)}_{ik} \phi^{(c)}_k(x,y)$. The basis functions $\bphi^{(c)}(x,y)$ are plotted below:

```{r}
#| label: boundary-2d-function-viz

# Plot boundary-induced 2d bases
soap_basis_bd
```

Last, we plot a sample of interior basis functions. *I think* each interior basis function is defined in a similar form as the boundary-induced functions:

First, suppose the residual values after accounting for the boundary values follow a function $\bG_i(x,y) - \Phi^{(c)}(x,y) = \Phi_i^{(f)}(x,y) + \epsilon_i^{(f)}(x,y)$.

Second, represent that function with the basis expansion $\Phi_i^{(f)}(x,y) = \sum_{k=1}^{K_f} \gamma^{(f)}_{ik} \phi^{(f)}_k(x,y)$.

Third, to derive each basis function $\phi^{(f)}_k(x,y)$, set up the corresponding 2-dimensional function $\rho_{k}(x,y)$ such that $\rho_{k}(x,y) = 0$ on the boundary, $\frac{\partial^2 \rho_k}{\partial x^2} + \frac{\partial^2 \rho_k}{\partial y^2} = 0$, and $\rho_{k}(x,y)$ rises to a singularity at one corresponding knot $(x^*_{k}, y^*_{k})$.

Fourth, solve for the function $\phi^{(f)}_k(x,y)$ as the solution to the equation  $\frac{\partial^2 \phi^{(f)}_k}{\partial x^2} + \frac{\partial^2 \phi^{(f)}_k}{\partial y^2} = \rho_k(x,y)$.

Fifth, to estimate the basis coefficients $\bgamma_i^{(f)}$, include a penalty functional $\bgamma_i^{(f) \top} \bP_f \bgamma_i^{(f)}$.

Then the interior function has the basis function representation $\Phi_i^{(f)}(x,y) = \sum_{k=1}^{K_f} \gamma_{ik}^{(f)} \phi^{(f)}_k(x,y)$. A sample of the basis functions $\bphi^{(f)}(x,y)$ are plotted below:

```{r}
#| label: basis-interior-viz

# Plot selection of interior bases
soap_basis_int
```

Finally, for one soap film smoother, we have the basis representation 
$$
\begin{aligned}
\bG_i(x,y) &= \Phi_i^{(c)}(x,y) + \Phi_i^{(f)}(x,y) + \epsilon_i(x,y) \\
&= \sum_{k=1}^{K_c} \gamma^{(c)}_{ik} \phi^{(c)}_k(x,y) + \sum_{k=1}^{K_f} \gamma^{(f)}_{ik} \phi^{(f)}_k(x,y) + \epsilon_i(x,y) \\
&= \begin{bmatrix} \bphi^{(c)}(x,y) & \bphi^{(f)}(x,y) \end{bmatrix} \begin{bmatrix} \bgamma^{(c)}_{i} \\ \bgamma^{(f)}_{i} \end{bmatrix}
+ \epsilon_i(x,y) \\
&= \mathcal{B}(x,y) \bgamma_i + \epsilon_i(x, y) \\
\implies \bG_i &= \bB \bgamma_i + \bepsilon_i.
\end{aligned}
$$

Notably, this shows that the basis functions $\bB(x,y)$ can be common across all individuals. In this case, the basis coefficients $\bgamma_i$ are the only thing varying across subjects, and each coefficient has the same interpretation across all subjects.

### Smoother Model Components {#sec-soap-film-components}

Then, we estimate the coefficients for this model for patient $i$ using the closed-form expression

$$
\hatbgamma_i = (\bB^\top \bB + \bP_T)^{-1}\bB^\top \bG_i
$$

We run a grid search to find the optimal smoothing penalty constants $\lambda_b$ and $\lambda_f$ (called "sp" in mgcv). We use the range $10^{-9}, ..., 10^{-2}$. We then select the model that achieves the minimum generalized cross-validation ("GCV") score:
$$
\begin{gather*}
GCV = \frac{m \sum_{x,y} [G_i(x,y) - \hat{G}_i(x,y)]^2}{[m - tr(\bH)]^2} \\
\bH = \bB (\bB^\top \bB + \bP_T)^{-1}\bB^\top \\
\tr(\bH) = \tr((\bB^\top \bB + \bP_T)^{-1}\bB^\top \bB)
\end{gather*}
$$

Last, we reshape our data to include one row for each patient, with a nested data set of row indices, column indices, and pointwise differences.

```{r}
#| label: frequentist-soap-fits

source(here("source", "check_matrix_duplicates.R"))

source(here("source", "11_fit_soap_film_freq.R"))
```


## Model Fitting without Boundary

### Data subset nested Loop {#sec-film-loop}

To test the model fit process, we take a subset of the data, only keeping the first appearance of each suture fusion type. We then run a nested for loop to fit the soap film model on each image and for each potential smoothing penalty constant. Plots of the GCV and details for the optimal fits are below.

```{r}
#| label: prelim-film-loop

## Plot results from model using loop, without boundary

# Plot GCV from grid search
film_loop_gcv_plot

# Report results from lowest GCV model for each patient
kbl(film_loop_optim %>% 
      select(fname, fusion_type, sp, gcv), 
    caption = "First Subset: Optimal Fit for each Patient") %>% 
  kable_styling(full_width = F)

# Plot predictions from loop model without boundary next to actual data
loop_film_both

```

### Data subset vectorized {#sec-film-map}

Next, we try to replicate this process by nesting everything within a dataset.


```{r}
#| label: prelim-film-map

# Plot GCV for these models
film_map_gcv_plot

# Report best GCV models
kbl(film_map_optim %>% 
      select(fname, fusion_type, sp, gcv), 
    caption = "First Subset: Optimal Fit by Map/Mutate") %>% 
  kable_styling(full_width = F)

# Plot predictions for nested model fitting without boundary, and actual data
map_film_both
```

## Model Fitting with Boundary

### Data subset nested loop {#sec-soap-loop}

To test the model fit process, we take a subset of the data, only keeping the first appearance of each suture fusion type. We then run a nested for loop to fit the soap film model on each image and for each potential smoothing penalty constant. Plots of the GCV and details for the optimal fits are below.

```{r}
#| label: prelim-soap-loop

# Plot GCV from model with boundary fit in loop
soap_loop_gcv_plot

# Print model with lowest GCV
kbl(soap_loop_df %>% 
      select(fname, fusion_type, sp_bound, sp_film, gcv), 
    caption = "First Subset: Optimal Fit for each Patient",
    scientific = TRUE,
    digits = 8) %>% 
  kable_styling(full_width = F)

# Plot predictions from model with boundary fit in loop, and actual data
loop_soap_both

```

### Data subset vectorized {#sec-soap-map}

Next, we replicate this process by nesting everything within a dataset.

```{r}
#| label: prelim-soap-map

# Print results from nested model fits
kbl(soap_map_df %>% 
      select(fname, fusion_type, sp_bound, sp_film, gcv), 
    caption = "First Subset: Optimal Fit by Map/Mutate",
    digits = 8) %>% 
  kable_styling(full_width = F)

# Plot predictions from model with boundary fit in nested data, and actual data
map_soap_both

# Plot predictions from nested models with and without boundary, plus data
plot_map_build
```

For this subset, we make predictions using only the interior spline, only the boundary spline, and both, and compare all three to the actual data. This shows an interesting result: when we exclude the boundary spline entirely and estimate coefficients, the resulting predicted values are almost all non-negative. When we estimate the coefficients using both splines and then model their contributions separately, the contribution from the interior is often negative, resulting in predictions near 0. This is because the basis functions in the design matrix are not just defined on the boundary; they are defined over the entire space, based on the boundary condition described by a cyclic boundary spline.

## Fitting Coefficients for All Data {#sec-soap-film-individual}

Once we have set up the soap film model, we estimate coefficients for each individual, $\hatbgamma_i$. These are stored in a nested tibble, which we save as an RDS file.

```{r}
#| results: hide

# Clear memory space
rm(so_test_df)
rm(list = setdiff(ls(), c(lsf.str(), ls(pattern = eval(RM_IGNORE)))))
gc(verbose = F)
```

```{r}
#| label: individual-soap-models

source(here("source", "12_fit_soap_film_gcv_individuals.R"))

```

We create a dataset containing only the demographics and estimated coefficients for each individual, and save this along with the design matrix $\bB$, for predictions from these soap film models based on demographics.

```{r}
#| results: hide

# Clear memory space
rm(so_constants)
rm(list = setdiff(ls(), c(lsf.str(), ls(pattern = eval(RM_IGNORE)))))
gc(verbose = F)
```

# Bayesian Soap Film {#sec-bayes-soap}

Next, we fit a model on one individual using a Gibbs sampler. Recall that we defined the "total penalty matrix" as 
$$
\bP_T =
\begin{bmatrix}
\lambda_c \bP_c & \bzero \\
\bzero & \lambda_f \bP_f
\end{bmatrix}.
$$
Note that neither $P_c$ nor $P_f$ are full-rank, but we can define the generalized inverse of $\bP_T$ as
$$
\bP^{-}_T =
\begin{bmatrix}
\frac{1}{\lambda_c} \bP_c^- & \bzero\\
\bzero & \frac{1}{\lambda_f} \bP_f^-
\end{bmatrix}.
$$
This generalized inverse is not positive-definite, but we can use it in an improper prior.

We also use the notation 
$$
\bgamma_i = \begin{pmatrix} \smash{\overbrace{\bgamma^{(c)}_{i}}^{K_c \times 1}} \\ \smash{\underbrace{\bgamma^{(f)}_{i}}_{K_f \times 1}} \end{pmatrix},
\\ \\ \\
$$
where $\bgamma_c$ are the coefficients corresponding to the boundary spline basis and $\bgamma_f$ are the coefficients corresponding to the interior "film" spline basis.

Then we assume the (conjugate) priors:

$$
\begin{aligned}
\bG_i & \sim \mathrm{N}_m ( \bB^\top \bgamma_i, \sigma^2 \bI_m) \\
\bgamma_i & \sim \mathrm{N}_K \Big( \bzero, \sigma^2 \bP^{-}_T \Big) \\
\sigma^2 & \sim \mathrm{Inv. Gamma} (a_{\sigma}, b_{\sigma}) \\
\lambda_{c} & \sim \mathrm{Gamma} (a_c, b_c) \\
\lambda_{f} & \sim \mathrm{Gamma} (a_f, b_f) \\
\lambda_{c} & \perp \lambda_{f}
\end{aligned}
$$

We choose simple hyperparameters: $a_\sigma = 1, \ b_\sigma = 1,  \ a_c = a_f = 1, b_c = b_f = 1$. We have not yet checked how informative they are.

These priors lead to the full conditionals:

$$
\begin{aligned} 
\ [\lambda_{c} | \bgamma^{(c)}_{i}, \sigma^2] & \sim \mathrm{Gamma} \Bigg( \frac{K_c}{2} + a_c, \ \frac{1}{2 \sigma^2} \bgamma^{(c) \top}_{i} \bP_c \bgamma^{(c)}_{i} + b_c \Bigg) \\
[\lambda_{f} | \bgamma^{(f)}_{i}, \sigma^2] & \sim \mathrm{Gamma} \Bigg( \frac{K_f}{2} + a_f, \ \frac{1}{2 \sigma^2} \bgamma^{(f) \top}_{i} \bP_f \bgamma^{(f)}_{i} + b_f \Bigg) \\
[\sigma^2 | \bgamma_i, \lambda_{c}, \lambda_{f}] & \sim \mathrm{Inv. Gamma} \Bigg( \frac{m}{2} + \frac{K}{2} + a_\sigma, \ \frac{1}{2} \Big[ \bG_i^\top \bG_i  - 2 \bgamma_i^\top \bB^\top \bG_i + \bgamma_i^\top \Big( \bB^\top \bB + \bP_T \Big) \bgamma_i \Big] + b_\sigma \Bigg) \\
[\bgamma_i | \sigma^2, \lambda_{c}, \lambda_{f}] & \sim \mathrm{N}_K \Bigg( \Big( \bB^\top \bB + \bP_T \Big)^{-1} \bB^\top \bG_i, \ \sigma^2 \Big( \bB^\top \bB + \bP_T \Big)^{-1} \Bigg)
\end{aligned}
$$

## Fit on individual model {#sec-bayes-individual}

```{r}
#| label: bayes-soap-film-test

set.seed(978)

source(here("source", "13_fit_soap_film_bayes.R"))
```


### MCMC Diagnostics {#sec-mcmc-diagnostics}

Trace plots show that the parameters converge to a stable distribution quickly and are well-differentiated.

```{r}
#| label: MCMC-diagnostics

# Report timing of Bayesian soap film model
kbl(mcmc_time %>% mutate(pct = scales::label_percent(accuracy = 0.1)(pct)), 
    caption = "Loop Timing") %>% 
  kable_styling(full_width = F)

# Trace plot for variance term
sigma_trace

# Trace plot for smoothness constants
lambda_trace
```

```{r}
#| label: soap-beta-trace
#| fig.width: 10
#| fig.height: 50

# Trace plot of betas
beta_trace
```

### Visualize predictions {#sec-bayes-soap-plot}

```{r}
#| label: bayes-soap-predictions

# Plot predictions and actual data from 1 Bayesian soap film
soap_bayes1_plot
```

# Basis Coefficient Individual Models {#sec-basis-coeff-models}

Once we have a working Bayesian soap film model for one individual, we switch to the other level of the hierarchy: for each basis coefficient, we estimate how that basis coefficient can be estimated as a function of a patient's age and suture fusion type.

Specifically, we assume that each basis coefficient $\gamma_k$ for individual $i$, $i = 1, \ldots, n, \ k = 1, \ldots, K$, can be estimated as the sum of a sex effect added flexible function of age categorized by suture type:

$$
\begin{aligned}
E[\bgamma_{ik}] = {} & \beta_{0k} + \beta_{1k} [\text{Sex}_i] + \\
& f_{0k} ( \text{Age}_i, \bbeta_k^{(0)}, \lambda_{k}^{(0)} ) + \\ 
& f_{1k} ( \text{Age}_i, \bbeta_k^{(S)}, \lambda_{k}^{(S)} ) * \bone [\text{Sagittal}] + \\
& f_{2k} ( \text{Age}_i, \bbeta_k^{(M)}, \lambda_{k}^{(M)} ) * \bone [\text{Metopic}] + \\
& f_{3k} ( \text{Age}_i, \bbeta_k^{(RC)}, \lambda_{k}^{(RC)} ) * \bone [\text{Right Coronal}] + \\
& f_{4k} ( \text{Age}_i, \bbeta_k^{(LC)}, \lambda_{k}^{(LC)} ) * \bone [\text{Left Coronal}] + \\
& f_{5k} ( \text{Age}_i, \bbeta_k^{(BC)}, \lambda_{k}^{(BC)} ) * \bone [\text{Bicoronal}]\\
\end{aligned}
$$

```{r}
#| label: fit-coeff-demographic-model-test

set.seed(603)

source(here("source", "14_fit_demographic_model_test.R"))
```


## Basis Coefficient Individual Plots {#sec-basis-coeff-plots}

We can check these assumptions based on the coefficients we estimated for each individual using GCV.

The assumptions appear to be reasonable, although there may be significant variance $\eta_{ik}$. This could be especially problematic for larger ages, where there are significantly fewer observations and are often no observations for less common suture fusion types.

```{r}
#| label: plot-coeff-demographics
#| fig.width: 10
#| fig.height: 15

coeff_plots
```

## Coefficient Model Design Matrix {#sec-coefficient-design}

First, we set up the design matrix and penalty matrices, which are the same for each basis coefficient. We currently use 10 knots for each suture fusion-specific smooth. Therefore, the design matrix currently includes 2 parametric covariates (1 intercept and 1 Sex indicator), 9 basis functions for a reference smooth, and $9 * 5 = 45$ basis functions for the smooths associated with Sagittal, Metopic, Left Coronal, Right Coronal, and Bicoronal. In total, the design matrix $\bW$ is $n \times Q$, where $Q = 56$.

*TO UPDATE: Note that these smooths account for an identifiability constraint. Since we are including multiple basis expansions, each covariate in the smooths needs to sum to zero. In practice, this leads us to subtract the column mean from each column of the design matrix, which reduces the rank of each smooth's columns by 1. Therefore, though we use 10 knots for each smooth, we only keep 9 columns.*

We set up the design matrix and the penalty matrices to match the Gibbs sampler function we have already written to fit a penalized linear regression model, which in turn matches the penalty matrices in `mgcv`. Therefore, each penalty matrix is $Q \times Q$ and block-diagonal, with one $9 \times 9$ penalty matrix included and the remainder of the matrix left as 0.

First, we construct a full "reference" matrix and penalty matrix using `mgcv::smooth.construct2`.

Second, we write functions that takes this smooth object, data, the name of the factor variable we want to construct separate splines for, and the names of the unpenalized "parametric" covariates, and constructs a full design matrix and penalty matrix list for these.

## Single Basis Coefficient Bayesian Additive Model {#sec-coeff-gibbs}

We use this function to construct the design matrix and penalty matrix list for the Gibbs sampler function. 

To test this model, we use the 38th soap film basis coefficient. This coefficient's basis function is plotted above, and it is in a region that visibly differs between Normative, Sagittal, and Metopic presentations.

## Basis Coefficient MCMC Diagnostics {#sec-coeff-mcmc-diagnostics}

We plot the diagnostics for the parameters from this Gibbs sampler.

In particular, we can see that the variance of some of the coefficients is very large. These coefficients correspond to large ages for rarer suture fusions like Left Coronal, Right Coronal, and Bicoronal. The variance highlights a potential issue with sample size. For these rare suture fusions, we do not have many observations at higher ages. This is not a problem for estimating Normative growth or even Sagittal or Metopic growth. However, if we want the same basis expansion for each suture fusion type, we need to leave the knots in the same place, which takes away from our ability to estimate more detailed patterns at younger ages. On the other hand, we could add more knots, but this will add to the computation of estimating this model.

```{r}
#| label: coeff-MCMC-diagnostics

# Report timing of this MCMC
kbl(mcmc_coeff_time %>% mutate(pct = scales::label_percent(accuracy = 0.1)(pct)), 
    caption = "Loop Timing") %>% 
  kable_styling(full_width = F)

# Plot trace plot of variance
sigma_trace_coeff

# Plot trace of smoothing constant
lambda_trace_coeff
```

```{r}
#| label: coeff-MCMC-betas
#| fig.width: 10
#| fig.height: 20

# Trace plot of demographic effects
beta_trace_coeff
```

### Visualize coefficient predictions {#sec-coeff-pred-plot}

We plot the coefficient functions of age by suture (including the estimated intercept).

For each suture fusion type, we construct an evenly spaced 100-length sequence from the minimum age to the maximum age. We use the previously defined function to convert this to a design matrix that fits with our model.

To estimate a credible interval, we generate predictions for every draw of betas from the Gibbs sampler, and then take the 5th and 95th percentiles of these predictions for each row of our constructed data.

These credible intervals are very large for rare suture fusion types for high ages, which is consistent with the large variances of coefficients shown in the trace plots above. To get a better idea of the patterns estimated by the coefficient functions for smaller ages, we reduce the size of the y axis. We also include the locations of the knots in our model to gain a better understanding of what level of detail we currently have from our GAM.

```{r}
#| label: bayes-coefficient-GAM-predictions

# Plot age effects by fusion type
coeff_bayes1_plot

# Plot age effects by fusion type WITH y axis limits
coeff_bayes1_plot_lim
```

# Combined Hierarchical Model {#sec-joint-model}

Next, we design the joint model that estimates these two levels of effects simultaneously.

## Model Explanation {#sec-joint-model-spec}

### Data Description {#sec-joint-model-data}

First, let $\bS$ be a non-rectangular set of $M = 52,800$ pixels that describe the intracranial space projected onto 2 dimensions. We index each pixel $s_m$ to a pair of integer coordinates $(x_m, y_m) \in \mathbb{Z}^2$. Suppose $\bomega$ is the $279 \times 234$ rectangular grid of pixels circumscribing $\bS$, defined as $\{(x, y) \in \mathbb{R}^2 \ | \ \exists m : x = x_m \vee y = y_m \}$. Then we define $\mathcal{S}$ as the smallest compact set in the continuous space $\mathbb{R}^2$ containing $\bS$, and $\bOmega$ as the smallest rectangle in the continuous space $\mathbb{R}^2$ containing $\bomega$.

Second, let $G_i(s_m)$ represent the measurement of growth in image $i$ at location $s_m$. Therefore, each image consists of the observations $\bG_i(\bS) = [G_i(s_1), \ldots, G_i(s_M)]^\top$. For simplicity, we'll write $\bG_i = \bG_i(\bS)$, and $\bG_i \text{ is } M \times 1$. Next, suppose we observe growth measurements on the same set of locations $\bS$ for $N$ individuals, so that we have a full data matrix of $\bG = [\bG_1, \ldots, \bG_N]^\top$, where $\bG \text{ is } N \times M$:

$$
\bG = \begin{bmatrix}
\bG_1^\top \\
\vdots \\
\bG_N^\top
\end{bmatrix} = 
\begin{bmatrix}
G_1(s_1) & G_1(s_2) & \ldots & G_1(s_M) \\
G_2(s_1) & G_2(s_2) & \ldots & G_2(s_M) \\
\vdots & & \ddots & \vdots \\
G_N(s_1) & G_N(s_2) & \ldots & G_N(s_M)
\end{bmatrix}
$$

Third, suppose that along with the observed growth measurements, we also have a vector of demographic covariates related to each image, $\bU_i$, so that $\bU$ is an $N$-row matrix of fixed covariates.

### Mean Estimation {#sec-joint-model-mean}

We treat each growth measurements $\bG_i(\mathcal{S})$ as one functional outcome, and we would like to estimate a mean function $\mu_i(\mathcal{s_m})$ for each outcome based on demographic covariates. To estimate this model, we set up a Bayesian hierarchical framework:

$$
G_i(s_m) \iid \mathcal{N} \Big( \mu_i(s_m), \sigma^2 \Big).
$$

To estimate $\mu_i(s_m)$, we use a spline expansion: 
$$
\mu_i(s_m) = \sum_{k=1}^K \gamma_{ik} B_k(s_m).
$$

To define functions over the irregular, finite space $\mathcal{S}$, we use a soap film smoother expansion to define each $B_k(\cdot)$, and calculate the value of $B_k(s_m) = B_k(x_m, y_m)$. We define a $10 \times 10$ grid of knots on $\bOmega$, 60 of which fall within $\mathcal{S}$. We have no prior conditions for the values of $G$ along the boundary of $\mathcal{S}$, so we also estimate the boundary values with a cyclic spline of dimension 10. Together, this results in $K_f = 60$ and $K_c = 9$ basis functions, for a total of $K = 69$ parameters in the spline expansion. Therefore, we can rewrite the expectation

$$
\begin{aligned}
E[G_i(s_m)] &= \sum_{k=1}^{K = 69} \gamma_{ik} B_k(s_m) \\
\implies E[\bG_i(\bS)] &= \sum_{k=1}^{K = 69} \gamma_{ik} B_k(\bS) \\
& = \bB(\bS) \bgamma_i \\
& := \bB \bgamma_i \\
\implies [ \bG_i | \bgamma_i, \sigma^2 ] & \iid \mathcal{N}_M(\bB \bgamma_i, \ \sigma^2 \bI_M) \\
\implies [ vec(\bG^\top) | \bgamma, \sigma^2 ] &\sim \mathcal{N}_{N*M} ( \underbrace{[\bI_N \otimes \bB]}_{N * M \times N*K} \ vec(\bgamma^\top), \ \sigma^2 \bI_{N*M}),
\end{aligned}
$$

where

$$
\begin{aligned}
\underbrace{\bgamma}_{N \times K} &= \begin{bmatrix}
\bgamma_1^\top \\
\vdots \\
\bgamma_N^\top
\end{bmatrix} 
= 
\begin{bmatrix}
\gamma_{11} & \ldots & \gamma_{1K_c} & \gamma_{1, K_c +1} & \ldots & \gamma_{1K} \\
\vdots & & & & & \vdots \\
\gamma_{N1} & \ldots & \gamma_{NK_c} & \gamma_{N, K_c +1} & \ldots & \gamma_{NK}
\end{bmatrix}
=
\begin{bmatrix}
\bgamma^{(c)} & \bgamma^{(f)}
\end{bmatrix} \\
\text{ and } \\
\underbrace{\bB}_{M \times K} &= 
\begin{bmatrix}
\bB(s_1) \\
\vdots \\
\bB(s_M)
\end{bmatrix}
=
\begin{bmatrix}
B_1(s_1) & \ldots & B_{K_c}(s_1) & B_{K_c+1}(s_1) & \ldots & B_K(s_1) \\
\vdots & &  &  & & \vdots \\
B_1(s_M) & \ldots & B_{K_c}(s_M) & B_{K_c+1}(s_M) & \ldots & B_K(s_M)
\end{bmatrix}
=
\begin{bmatrix}
\bB^{(c)}(\bS) & \bB^{(f)}(\bS) \\
\end{bmatrix}
.
\end{aligned}
$$

### Soap Film Basis Coefficients Estimation {#sec-joint-model-coefficients}

At this level, we have modeled each individual image's spatial pattern across $\bS$ with the parameters $\bgamma$. Next, we account for the effect of the demographics $\bU$ on the growth $\bG$ by estimating each $\gamma_{ik}$ as a function of demographics. 

In our case, it is not appropriate to model $\gamma_{ik}$ as a linear function of the demographics $\bU_i$. Instead, we will model $\gamma_{ik}$ as a linear function of some untransformed covariates as well as a second spline expansion over Age, separated by suture fusion type. To make our model general enough to accomodate any linear combination of nonlinear functions of covariates $\bU_i$, let $\bW$ be the $N \times Q$ matrix containing untransformed covariates as well as realizations of basis spline expansions.

Suppose that there are $Q_\theta$ covariates in $\bU$ that we assume to be linearly related to $\bgamma$. Let $\tilde{\bU}$ be the $N \times Q_\theta$ matrix including only these linear terms. Note that in our case, the first of these $Q_\theta$ columns is a column vector of 1's corresponding to the intercept. 

Next, in our model, one covariate in $\bU$ is the Age of the subject in each image. We assume that age has a non-linear association with growth, and also that the non-linear association will vary by suture fusion type. To put this into practice, we will estimate one non-linear function $\Phi_{1k}(Age_i)$ for those with no suture fusion, and then will estimate a fusion-specific deviation function for each suture fusion type, $\mathcal{H}_{rk}(Age_i, Fusion_i)$.

Putting these together, we assume

$$ 
E[\gamma_{ik}] = \theta_{1k} + \theta_{2k} \tilde{U}_{i2} + \ldots + \theta_{Q_\theta, k} \tilde{U}_{iQ_\theta} + \mathcal{H}_{1k}(Age_i) + \sum_{r=2}^R \mathcal{H}_{rk}(Age_i, Fusion_i).
$$

Next, we convert this equation into matrix form. The linear terms are relatively straightforward to convert to matrix form:

$$
\theta_{1k} + \theta_{2k} \tilde{U}_{i2} + \ldots + \theta_{Q_\theta, k} \tilde{U}_{iQ_\theta} = 
\begin{bmatrix}
1 & \tilde{U}_{i2} & \ldots & \tilde{U}_{iQ_\theta}
\end{bmatrix} \begin{bmatrix}
\theta_{1k} \\ \vdots \\ \theta_{Q_\theta,k}
\end{bmatrix} = \tilde{\bU}_i \btheta_k.
$$

To estimate the first non-linear function, we use a spline expansion: $\mathcal{H}_{1k}(Age_i) = \sum_{l=1}^L \alpha_{1lk} h_l(Age_i)$. Then, for each of the $R-1$ remaining suture fusion types, we estimate a separate fusion-specific function $\mathcal{H}_{rk}(Age_i, Fusion_i)= I(Fusion_i = r) * [\sum_{l=1}^L \alpha_{rlk} h_l(Age_i)]$. As a simplifying assumption, we use the same spline expansion to estimate these $R$ different non-linear functions. the spline expansion representation of each non-linear function can also be converted to matrix multiplication:

$$
\begin{aligned}
\mathcal{H}_{1k}(Age_i) &= \sum_{l=1}^L \alpha_{1lk} h_l(Age_i) \\
&= \begin{bmatrix}
h_1(Age_i) & \ldots & h_L(Age_i)
\end{bmatrix} \begin{bmatrix}
\alpha_{11k} \\ \vdots \\ \alpha_{1Lk}
\end{bmatrix} \\
&= \bh_{i1} \balpha_{1k} \\ \\
\mathcal{H}_{rk}(Age_i, Fusion_i) &= \sum_{l=1}^L \alpha_{rlk} h_l(Age_i) * I(Fusion_i = r) \\
&= I(Fusion_i = r) * \begin{bmatrix}
h_1(Age_i) & \ldots & h_L(Age_i)
\end{bmatrix} \begin{bmatrix}
\alpha_{r1k} \\ \vdots \\ \alpha_{rLk}
\end{bmatrix} \\
&= \begin{bmatrix}
h_{r1}^*(Age_i, Fusion_i) & \ldots & h_{rL}^*(Age_i, Fusion_i)
\end{bmatrix} \begin{bmatrix}
\alpha_{r1k} \\ \vdots \\ \alpha_{rLk}
\end{bmatrix} \\
&= \bh_{ir}^* \balpha_{rk}, \\ \\
\text{where } h_{rl}^*(Age_i, Fusion_i) &= I(Fusion_i = r) * h_l(Age_i)
\end{aligned}
$$

Putting realizations of these functions together with $\tilde{\bU}$, we have:

$$
\begin{aligned}
\bW &= \begin{bmatrix}
\tilde{\bU} & \bh_1 & \bh_2^* & \ldots & \bh_{R}^*
\end{bmatrix} \\
&= 
\begin{bmatrix}
1 & \tilde{U}_{12} & \ldots & \tilde{U}_{1Q_\theta} & h_1(Age_1) & \ldots & h_L(Age_1) & h_{21}^*(Age_1, Fusion_1) & \ldots &  h_{RL}^*(Age_1, Fusion_1) \\
\vdots \\
1 & \tilde{U}_{N2} & \ldots & \tilde{U}_{NQ_\theta} & h_1(Age_N) & \ldots & h_L(Age_N) & h_{21}^*(Age_N, Fusion_N) & \ldots &  h_{RL}^*(Age_N, Fusion_N)
\end{bmatrix}
\end{aligned}.
$$

Then we define one coefficient vector relating each soap film basis coefficient $\gamma_k$ to all of these demographics in $\bW$. This coefficient vector $\bbeta_k$ contains $Q_\theta$ linear coefficients for untransformed covariates, which we denote $\btheta_k$, as well as $R*L$ coefficients defining one $L$-dimensional spline expansion of age for each of the $R$ fusion suture types, which we denote $\balpha_k$. Putting these together, we can write:

$$
\begin{aligned}
\bbeta_k^\top &= \begin{bmatrix}
\btheta_k^\top & \balpha_k^\top
\end{bmatrix} \\
&= \begin{bmatrix}
\theta_{1k} & \ldots & \theta_{Q_\theta,k} & \alpha_{11k} & \ldots & \alpha_{1Lk} & \alpha_{21k} & \ldots & \alpha_{RLk}
\end{bmatrix},
\end{aligned}
$$

Thus, $\bW$ is $N \times Q$ and $\bbeta$ is $Q \times K$, where $Q = Q_\theta + R*L$. 

Then altogether, we can write

$$
\begin{aligned}
E[\gamma_{ik}] = {} & \theta_{1k} + \theta_{2k} \tilde{U}_{i2} + \ldots + \theta_{Q_\theta, k} \tilde{U}_{iQ_\theta} +  \mathcal{H}_{1k}(Age_i) + \sum_{r=2}^R \mathcal{H}_{rk}(Age_i, Fusion_i) \\
= {} & \tilde{\bU}_i \btheta_k + \bh_{i1} \balpha_{1k} + \sum_{r = 2}^R \bh_{ir}^* \balpha_{rk} \\ \\
\implies E[\gamma_{ik}] = {} & \bbeta_k^\top \bW_i^\top \\
\implies E[\bgamma_i] = {} & \bbeta^\top \bW_i^\top \\
\implies E[\bgamma_i^\top] = {} & \bW_i \bbeta \\
\implies E[\bgamma] = {} & \bW \bbeta .
\end{aligned}
$$

Next, we determine the prior variance of $\bgamma$. We need to penalize the complexity of each $\mu_i$ function over $\bS$; recall $\mu_i(\bS) = \sum_{k=1}^K \gamma_{ik} B_k(\bS)$. This complexity can be quantified using one penalty matrix capturing the variation around the boundary, $\bP_c$, and one matrix capturing the variation within the interior space of the soap film, $\bP_f$. These can be combined to one "total" penalty matrix inducing smoothing across $\bgamma_i$ terms, along with smoothing constants $\lambda_c$ and $\lambda_f$:

$$
\bP_T^{(\gamma)} = \begin{bmatrix}
\lambda_c \bP_c & \bzero \\
\bzero & \lambda_f \bP_f
\end{bmatrix}
$$

To discourage the complexity captured by this matrix, we set the prior:

$$ 
\begin{aligned}
\ [\bgamma_i | \bbeta, \sigma^2, \lambda_c,  \lambda_f] & \iid \mathcal{N}_K (\bbeta^\top \bW_i^\top, \ \sigma^2 \bP_T^{(\gamma) -}) \\
\implies [ vec(\bgamma) | \bbeta, \sigma^2, \lambda_c,  \lambda_f] & \sim \mathcal{N}_{N*K} (\underbrace{[\bI_K \otimes \bW]}_{N*K \times Q*K} \ vec(\bbeta), \ \sigma^2 [\bI_N \otimes \bP_T^{(\gamma) -}]).
\end{aligned}
$$

Recall that $\bbeta$ has a complicated structure: 

$$
\begin{aligned}
\underbrace{\bbeta}_{Q \times K} &= \begin{bmatrix} 
\bbeta^{(c)} & \bbeta^{(f)}
\end{bmatrix} \\
&= 
\begin{bmatrix} 
\bbeta_1 & \ldots & \bbeta_{K_c} & \bbeta_{K_c + 1} & \ldots & \bbeta_K 
\end{bmatrix} \\
&= 
\begin{bmatrix}
\beta_{11} & \ldots & \beta_{1,K_c} & \bbeta_{1, K_c + 1} & \ldots &  \beta_{1K} \\
& \ddots & & & \ddots \\
\beta_{Q1} & \ldots & \beta_{Q,K_c} & \bbeta_{Q, K_c + 1} & \ldots &  \beta_{QK}
\end{bmatrix} \\
&= 
\begin{bmatrix}
\theta_{11} & \ldots & \theta_{1,K_c} & \theta_{1,K_c + 1} & \ldots & \theta_{1K} \\
 & \ddots & & & \ddots \\
\theta_{Q_\theta, 1} & \ldots & \theta_{Q_\theta, K_c} & \theta_{Q_\theta, K_c + 1} & \ldots & \theta_{Q_\theta, K} \\
\alpha_{111} & \ldots & \alpha_{11K_c} & \alpha_{1,1, K_c + 1} & \ldots & \alpha_{11K} \\
 & \ddots & & & \ddots & \\
\alpha_{R L 1} & \ldots & \alpha_{R L K_c} & \alpha_{R, L, K_c + 1} & \ldots & \alpha_{R L K}
\end{bmatrix} \\
&= \begin{bmatrix} 
\btheta^{(c)} & \btheta^{(f)} \\
\balpha^{(c)} & \balpha^{(f)} \\
\end{bmatrix} \\
&= \begin{bmatrix} 
\btheta \\ \balpha \\
\end{bmatrix}.
\end{aligned}
$$

### Demographic Effect Estimation {#sec-joint-model-demographics}

Last, we need to specify a prior distribution for $\bbeta$. 

We want to penalize the complexity of each nonlinear function defined by one set $\balpha_{kr} = \{\alpha_{kr1}, \ldots, \alpha_{krL}\}$. Similar to above, the complexity of the function defined by a nonlinear function can be quantified using an $L \times L$ penalty matrix $\bP_h$ and a corresponding penalty constant $\lambda_\alpha$. For our model, we assume that all of the age effects can be modeled using the same spline expansion, and all spline expansions share one $\lambda_\alpha$. Therefore, 

$$
\bP_T^{(\alpha)} = \lambda_\alpha \bP_\alpha = \lambda_\alpha [\bI_R \otimes \bP_h].
$$

To combine $\balpha_k$ with $\btheta_k$, we also need to establish a prior variance for $\btheta_k$ that leaves no penalty for complexity on $\btheta_k$. This is accomplished by using an identity matrix to represent the prior covariance of $\btheta_k$. Then:

$$
\bP_T^{(\beta)} = \begin{bmatrix}
\bI_{Q_\theta} & \bzero \\
\bzero & \bP_T^{(\alpha)}
\end{bmatrix} = 
\begin{bmatrix}
\bI_{Q_\theta} & \bzero & \ldots & \bzero \\
\bzero & \lambda_\alpha \bP_h & \ldots &\bzero \\
\vdots & & \ddots & \vdots \\
\bzero & \bzero & \ldots & \lambda_\alpha \bP_h
\end{bmatrix}
$$

Last, we assume no prior covariance between $\bbeta_k$ and $\bbeta_{k'}$, $k \neq k'$, and we add a prior scaling term $\tau^2$ to allow the magnitude of $\bbeta$ to match the magnitude of $\bgamma$.  Therefore, we assume the prior:

$$
\begin{aligned}
\ [\bbeta_k | \tau^2, \lambda_\alpha] & \iid \mathcal{N}_Q (\bzero, \tau^2 \bP_T^{(\beta) -}) \\
\implies \ [vec(\bbeta) | \tau^2, \lambda_\alpha] & \sim \mathcal{N}_{Q*K} (\bzero, \tau^2 [\bI_K \otimes \bP_T^{(\beta) -}])
\end{aligned}
$$

## Distributional Assumptions {#sec-joint-model-distributions}

We can now express all of our distributional assumptions. 

### Data-Generating Mechanism

We write the data-generating mechanism below. Given matrices are highlighted in $\color{green}{\text{green}}$; scalar parameters that are directly set in simulations are highlighted in $\color{red}{\text{red}}$ the first time they appear; other parameters that are recovered but are drawn from a distribution in a simulation are highlighted in $\color{blue}{\text{blue}}$; and the outcome data are highlighted in $\color{violet}{\text{violet}}$.

$$
\begin{aligned}
\underbrace{{\color{violet} \bG}}_{N \times M} &= {\color{blue}  \bgamma } {\color{green} \bB^\top } + \bepsilon^{(G)} \\
\bepsilon^{(G)}_{im} & \iid \mathcal{N} ( 0, {\color{red} \sigma^2 }) \\
\underbrace{\bgamma}_{N \times K} &= {\color{green} \bW} {\color{blue} \bbeta } + \bepsilon^{(\gamma)} \\
\bepsilon^{(\gamma)}_{i+} & \iid \mathcal{N} ( \bzero, \sigma^2 \bP_T^{(\gamma) -}) \\
\bP_T^{(\gamma)} &:= \begin{bmatrix}
\color{red}{ \lambda_c } {\color{green} \bP_c} & \bzero \\
\bzero & {\color{red} \lambda_f } {\color{green} \bP_f}
\end{bmatrix} \\
\underbrace{\bbeta}_{Q \times K} & := \begin{bmatrix}
\bbeta_{+1} & \ldots & \bbeta_{+K}
\end{bmatrix} \\
\bbeta_{+k} & \iid \mathcal{N} (\bzero, {\color{red} \tau^2 } \bP_T^{(\beta) -}) \\
\bP_T^{(\beta)} &:= \begin{bmatrix}
\bI_{Q_\theta} & \bzero \\
\bzero & {\color{red} \lambda_\alpha } \bP_\alpha
\end{bmatrix}
\end{aligned}
$$

### Prior Distributions {#sec-joint-model-priors}

We assume conjugate priors for the hyperparameters $\sigma^2, \ \lambda_c, \ \lambda_f, \, \tau^2, \text{ and } \lambda_\alpha$. 

Last, it is often necessary to express the distributions both of the vectorized coefficient matrices and the vectorized transposes of the same matrices. To do this, we use a property of Kronecker products:

$$
\begin{alignat*}{2}
& vec(ABC) && = (C^\top \otimes A) vec(B) \\ \\
\implies vec(AB) = {} & vec(ABI) && = (I \otimes A) vec(B) \\
= {} & vec(IAB) && = (B^\top \otimes I) vec(A).
\end{alignat*}
$$

Incorporating these into the assumptions described above, all distributional assumptions are listed below.

$$
\begin{alignat*}{3}
& (0) & [vec(\bG) | \bgamma, \sigma^2] & \sim {} && \mathcal{N}_{N*M} ( [\bB \otimes \bI_N] vec(\bgamma), \sigma^2 \bI_{N*M} ) \\
& & \iff [vec(\bG^\top) | \bgamma, \sigma^2] & \sim {} && \mathcal{N}_{N*M} ( [\bI_N \otimes \bB] vec(\bgamma^\top), \sigma^2 \bI_{N*M} ) \\ \\
& (1) & {} [vec(\bgamma) | \bbeta, \sigma^2, \lambda_c, \lambda_f] & \sim {} && \mathcal{N}_{N*K} ( [\bI_K \otimes \bW] vec(\bbeta), \sigma^2 [\bP_T^{(\gamma) -} \otimes \bI_N] ) \\
& & \iff [vec(\bgamma^\top) | \bbeta, \sigma^2, \lambda_c, \lambda_f] & \sim {} && \mathcal{N}_{N*K} ( [\bW \otimes \bI_K] vec(\bbeta^\top), \sigma^2 [\bI_N \otimes \bP_T^{(\gamma) -}] ) \\ \\
& (2) & [vec(\bbeta) | \tau^2, \lambda_\alpha] & \sim {} && \mathcal{N}_{Q*K} (\bzero, \tau^2 [\bI_K \otimes \bP_T^{(\beta) -}] ) \\
& & \iff [vec(\bbeta^\top) | \tau^2, \lambda_\alpha] & \sim {} && \mathcal{N}_{Q*K} (\bzero, \tau^2 [\bP_T^{(\beta) -} \otimes \bI_K]) \\ \\
& (3) & \sigma^2 & \sim {} && \text{Inv. Gamma}(a_\sigma, b_\sigma) \\
& (4) & \lambda_c & \sim {} && \text{Gamma}(a_{c}, b_{c}) \\
& (5) & \lambda_f & \sim {} && \text{Gamma}(a_{f}, b_{f}) \\
& (6) & \tau^2 & \sim {} && \text{Inv. Gamma}(a_\tau, b_\tau) \\
& (7) & \lambda_\alpha & \sim {} && \text{Gamma}(a_\alpha, b_\alpha)
\end{alignat*}
$$

### Full Conditionals {#sec-joint-model-conditionals}

These priors yield closed-form full conditional distributions, written below.

$$
\begin{alignat*}{4}
& (1) & {} [vec(\bgamma) | \cdot] & \sim {} && \mathcal{N}_{N*K} ( && [(\bB^\top \bB + \bP_T^{(\gamma)})^{-1} \otimes \bI_N] [(\bB^\top \otimes \bI_N) vec(\bG) + (\bP_T^{(\gamma)} \otimes \bW) vec(\bbeta)], \\
& & & && && \sigma^2 [ \underbrace{(\bB^\top \bB + \bP_T^{(\gamma)})^{-1}}_{K \times K} \otimes \bI_N ] \\ \\
& (2) & {} [vec(\bbeta) | \cdot ] & \sim {} && \mathcal{N}_{Q*K} ( &&  [(\bP_T^{(\gamma)} \otimes \bW^\top \bW) + \frac{\sigma^2}{\tau^2} (\bI_K \otimes \bP_T^{(\beta)})]^{-1} [(\bP_T^{(\gamma)} \otimes \bW^\top) vec(\bgamma)], \\
& & & && && \sigma^2 [(\bP_T^{(\gamma)} \otimes \bW^\top \bW) + \frac{\sigma^2}{\tau^2} (\bI_K \otimes \bP_T^{(\beta)})]^{-1}) \\ \\
& (3) & [\sigma^2 | \cdot ] & \sim {} && \Gamma^{-1} ( && a_\sigma + \frac{N*M}{2} + \frac{N*K}{2}, \\
& & & && && b_\sigma + \frac{1}{2} [vec(\bG) - (\bB \otimes \bI_N) vec(\bgamma)]^\top [vec(\bG) - (\bB \otimes \bI_N) vec(\bgamma)] + \\
& & & && && \frac{1}{2}[vec(\bgamma) - (\bI_K \otimes \bW) vec(\bbeta)]^\top [\bP_T^{(\gamma)} \otimes \bI_N] [vec(\bgamma) - (\bI_K \otimes \bW) vec(\bbeta)]) \\ \\
& (4) & [\lambda_c | \cdot] & \sim {} && \Gamma ( && a_c + \frac{N * K_c}{2}, \\
& & & && && b_c + \frac{1}{2 \sigma^2} [vec(\bgamma^{(c)}) - (\bI_{K_c} \otimes \bW) vec(\bbeta^{(c)})]^\top [\bP_c \otimes \bI_N] [vec(\bgamma^{(c)}) - (\bI_{K_c} \otimes \bW) vec(\bbeta^{(c)})] ) \\ \\
& (5) & [\lambda_f | \cdot] & \sim {} && \Gamma ( && a_f + \frac{N * K_f}{2}, \\
& & & && && b_f + \frac{1}{2 \sigma^2} [vec(\bgamma^{(f)}) - (\bI_{K_f} \otimes \bW) vec(\bbeta^{(f)})]^\top [\bP_f \otimes \bI_N] [vec(\bgamma^{(f)}) - (\bI_{K_f} \otimes \bW) vec(\bbeta^{(f)})] ) \\ \\
& (6) & [\tau^2 | \cdot] & \sim {} && \Gamma^{-1} ( && a_\tau + \frac{Q*K}{2}, \\
& & & && && b_\tau + \frac{1}{2} vec(\bbeta)^\top (\bI_K \otimes \bP_T^{(\beta)}) vec(\bbeta) ) \\ \\
& (7) & [\lambda_\alpha | \cdot] & \sim {} && \Gamma ( && a_\alpha + \frac{R*L*K}{2}, \\
& & & && && b_\alpha + \frac{1}{2 \tau^2} vec(\balpha^\top)^\top (\bP_\alpha \otimes \bI_K) vec(\balpha^\top) )
\end{alignat*}
$$

## Combined Model Fit {#sec-joint-model-fit}

```{r}
#| results: hide

# Clear memory space
rm(list = setdiff(ls(), c(lsf.str(), ls(pattern = eval(RM_IGNORE)))))
gc(verbose = F)
```

```{r}
#| label: bayesian-hierarchical-gibbs

set.seed(617)

# # ***DRAFT***: Fix sigma squared to improve scale
# sigmasq_true <- 50 # Similar to variance of all individuals' growth
# tausq_true <- 10
# lambdas_demo <- 1e-4

source(here("source", "15_run_gibbs_sampler.R"))
```

```{r}
#| label: show joint model results

# Plot patterns across age for representative points
outcomes_plot

# Plot patterns across shape for representative ages
joint_estimates_plot
```

