---
title: "Visualizing Hierarchical Simulations"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    toc-title: Contents
    toc-location: left
    df-print: kable
    theme: sandstone
    highlight: haddock
    code-fold: true
    code-overflow: wrap
    knitr:
      opts_chunk:
        echo: true,
        include: true
        message: false
        warning: false
        results: markup
        fig.align: center
        fig.height: 5
        fig.width: 7
        R.options:
          scipen: 999
          knitr.kable.NA: ""
editor: source
---

```{r}
#| label: setup
#| include: false

rm(list = ls()); gc()

# Load packages
# if(!require("pacman")) install.packages("pacman")
pacman::p_load(
  
  # Format
  latex2exp, table1, gtsummary, ggridges, forestplot,
  broom.helpers, broom.mixed, patchwork, colorRamps, pals,
  
  # Modeling
  mgcv,
  
  # Useful extensions
  gtools, corrplot, MASS, car, Matrix, mvtnorm, reshape2,
  
  # Basics
  DescTools, tidyverse, magrittr, janitor, glue, skimr,
  knitr, kableExtra, ggpubr, ggpmisc,
)

source(here::here("source", "construct_smooth_ref_code_fns.R"))
source(here::here("source", "mv_normal_matrix_fn.R"))
source(here::here("source", "hierarchical_penalized_gibbs_fn.R"))

DO_GIBBS <- T
LOAD_GIBBS_DATE <- "2026-02-17"
```

# Overview

This workbook explains and visualizes the data-generating mechanism or our combined hierarchical model. We simulate simplified data from our model and visualize the components of this simulation.

## Data-Generating Mechanism

We assume the following data-generating mechanism. 

Let $G_i(s_m)$ be an outcome of intracranial growth at location $s_m$ in image $i$, where $i = 1, \ldots, N$ and $m = 1, \ldots, M$. Let 
$$
\underbrace{\mathbf{G}}_{N \times M} = \begin{bmatrix}
G_1(s_1) & \ldots & G_1(s_M) \\
\vdots & \ddots & \vdots \\
G_N(s_1) & \ldots & G_N(s_M)
\end{bmatrix}
$$

be the matrix of growth outcomes for all images at all locations.

Let $\mathbf{W}_{i}$ be a $1 \times Q$ demographic row vector associated with image $i$, and let 
$$
\underbrace{\mathbf{W}}_{N \times Q} = \begin{bmatrix}
\mathbf{W}_{1} \\
\vdots \\
\mathbf{W}_{N}
\end{bmatrix}
$$

be the (known) matrix of demographics associated with all images.

Next, let $B_k(s_m)$ be the realization of soap film basis function $k$ at location $s_m$, where $k = 1, \ldots, K$, and let

$$
\underbrace{\mathbf{B}}_{M \times K} = \begin{bmatrix}
B_1(s_1) & \ldots & B_K(s_1) \\
\vdots & \ddots & \vdots \\
B_1(s_M) & \ldots & B_K(s_M)
\end{bmatrix}
$$

be the matrix of (known) realizations of all basis functions at all locations.

Then we assume that $\mathbf{G}$ is a function of these spatial basis functions and the demographics, and we require three different (known) penalty matrices to model these relationships: 

$\mathbf{P}_c$ is a penalty matrix on the boundary-induced soap film basis functions, which make up the first $K_c$ of the $K$ basis functions in $\mathbf{B}$.

$\mathbf{P}_f$ is a penalty matrix on the interior soap film basis functions, which make up the last $K_f$ of the $K$ basis functions in $\mathbf{B}$.

$\mathbf{P}_\alpha$ is a penalty matrix on nonlinear demographic effects modeled by an $L$-dimension basis expansion.

These known penalty matrices are combined with scalar smoothing parameters to form the total penalty matrices used in our model. For the full set of spatial basis functions, we define

$$
\mathbf{P}_T^{(\gamma)} := \begin{bmatrix}
\lambda_c \mathbf{P}_c & \boldsymbol{0} \\
\boldsymbol{0} & \lambda_f \mathbf{P}_f
\end{bmatrix}.
$$

For the full set of demographics, we assume that $Q_U$ demographics are unpenalized parameters, and that there are $R$ different suture fusions for which we want to estimate a nonlinear effect. Therefore,

$$
\mathbf{P}_T^{(\beta)} := \begin{bmatrix}
\mathbf{I}_{Q_U} & \boldsymbol{0} \\
\boldsymbol{0} & \lambda_\alpha (\mathbf{I}_R \otimes \mathbf{P}_\alpha)
\end{bmatrix}.
$$

Now, our data generating mechanism can be written:

$$
\begin{aligned}
\mathbf{G} &= \boldsymbol{\gamma} \mathbf{B}^\top + \boldsymbol{\epsilon}^{(G)} \\
\boldsymbol{\gamma} &= \mathbf{W} \boldsymbol{\beta} + \boldsymbol{\epsilon}^{(\gamma)} \\
\boldsymbol{\epsilon}^{(G)}_{im} & \overset{iid}{\sim} \mathcal{N}(0, \sigma^2) \\
\boldsymbol{\epsilon}^{(\gamma)}_{i \bullet} & \overset{iid}{\sim} \mathcal{N}_K(\boldsymbol{0}, \sigma^2 \mathbf{P}_T^{(\gamma) -}) \\
\boldsymbol{\beta}_{\bullet k} & \overset{iid}{\sim} \mathcal{N}_Q(\boldsymbol{0}, \tau^2 \mathbf{P}_T^{(\beta) -}).
\end{aligned}
$$

# Brain Shape Basis Functions

To implement this, we first need to reload $\mathbf{B}$, the matrix of soap film basis function realizations across the brain shape, which we create as part of the model fitting process. Recall that this shared across every image.

```{r}
#| label: load-brain-shape

# Load soap film object for basis functions
load(file = here::here("analysis", "intermediate", "soap_object.rda"))

# Load mapping from basis points to locations
load(file = here::here("analysis", "intermediate", "matrix_indices.rda"))
```

These basis functions are generated by `mgcv` using a soap film smoothing basis. The first nine ($K_c$) capture patterns around the boundary of the domain, and the remaining 60 ($K_f$) capture the values within the interior of the domain.

```{r}
#| label: basis-fn-viz
#| fig.width: 10
#| fig.height: 20

# Add row numbers to cell-location mapping
cell_loc_map <- so_matrix_pts %>% 
  rownames_to_column("loc") %>% 
  mutate(loc = as.integer(loc))

# Reshape basis function values to long df and merge with locations
soap_fn_df <- cranio_soap$X %>% 
  reshape2::melt() %>% 
  rename(loc = Var1, fn = Var2) %>% 
  left_join(cell_loc_map)

# Plot basis functions
soap_basis_plot <- ggplot(soap_fn_df, 
                          aes(x = row, y = col, fill = value)) +
  geom_raster() +
  facet_wrap(~fn, ncol = 9) +
  coord_fixed() +
  scale_fill_viridis_c() +
  theme_void() +
  theme(legend.position = "bottom")

soap_basis_plot
```

The associated penalty matrices $\mathbf{P}_c$ and $\mathbf{P}_f$ measuring the wiggliness of these basis functions over the domain are represented below.

```{r}
#| label: basis-penalty-viz

basis_s <- map(
  1:length(cranio_soap$S), 
  ~ cranio_soap$S[[.x]] %>% 
    reshape2::melt() %>% 
    rename(row = Var1, col = Var2) %>% 
    mutate(index = .x)
) %>% 
  bind_rows()

basis_s_plot <- basis_s %>% 
  filter(value != 0) %>% 
  group_split(index) %>% 
  map(
    ~ggplot(., aes(x = col, y = -row, fill = value)) +
      geom_tile() +
      facet_wrap(~index, nrow = 1) +
      scale_fill_gradient2() +
      scale_x_continuous(breaks = scales::breaks_extended(10)) +
      scale_y_continuous(breaks = scales::breaks_extended(10)) +
      theme_minimal() +
      theme(legend.position = "bottom") +
      coord_fixed()
  ) %>% 
  ggarrange(plotlist = ., nrow = 1) %>% 
  annotate_figure(top = "Soap Film Penalty Matrices")

basis_s_plot
```


# Individual Demographics

## Demographic Generation

Next, we generate a simplified set of demographics for each individual. In our model, we consider three characteristics or each individual: their sex (binary), age (continuous), and suture fusion type (categorical). We denote the matrix of these demographics for each individual as $\mathbf{U}$.

```{r}
#| label: demographic-hyperparameters

### Set parameters for simulating demographics

# Number of patients
n_pats <- 80

# Max age of these patients
age_max <- 365

# Proportion of sex = 1
prop_sex <- 0.5
```

For this simulation, we randomly generate `r n_pats` draws of age from a uniform distribution from 0 to `r age_max`. This is simplified from our data, where the age distribution appears closer to a Gamma distribution, especially when we only consider patients with suture fusions.

Then, we randomly assign $Sex = 1$ to these age values with probability `r prop_sex`. 

Finally, we assign the simulated suture fusion types of these patients: half are assigned to be "Normative," one quarter are assigned to be "Fusion 1," and one quarter are assigned to be "Fusion 2."

```{r}
#| label: create-demographics

# Set random seed
set.seed(978)

demo_test_df <- tibble(
  sex = rbinom(n_pats, 1, prop_sex),
  age = runif(n_pats, max = age_max),
  fusion_type = c(rep("Normative", n_pats / 2),
                  rep("Fusion_1", n_pats / 4),
                  rep("Fusion_2", n_pats / 4))
) %>%
  mutate(fusion_type = factor(fusion_type,
                              levels = c("Normative", "Fusion_1", "Fusion_2")),
         sex = as.factor(sex))
```

The resulting distribution of ages is:

```{r}
#| label: demographic-viz

demo_plot <- ggplot(demo_test_df) +
  geom_area(stat = "density", aes(x = age, fill = fusion_type)) +
  facet_grid(fusion_type ~ sex,
             labeller = labeller(.rows = label_value, .cols = label_both)) +
  theme_minimal() +
  theme(legend.position = "bottom")

demo_plot
```

## Demographic Model Matrix

Next, we use these simulated demographics to generate the demographic model matrix $\mathbf{W}$. We add an intercept column of all 1's and retain the sex column as it is, but we transform the age and fusion type columns with a basis spline expansion. 

```{r}
#| label: demographic-age-function-params

# Set number of knots to use for each age expansion
demo_age_knots <- 14
```

We use a cubic regression spline *with shrinkage* with `r demo_age_knots` basis functions for the nonlinear function of age. To construct the basis expansion, we use the `mgcv::smooth.construct2` function. We denote these functions as $h^*_1(Age), \ldots, h^*_{L+1}(Age)$.

```{r}
#| label: demographic-basis-expansion

# Specify smooth object
demo_smooth_obj <- s(age, bs = "tp", k = demo_age_knots)

demo_smooth_noshrink <- smooth.construct2(
  demo_smooth_obj, data = demo_test_df, knots = NULL
)

# Specify multiple of last eigenvalue to multiply by
shrink_eigen_mult <- 1e-5
attr(demo_smooth_obj, "shrink") <- shrink_eigen_mult

# Construct basis expansion and penalty matrix
demo_smooth_test <- smooth.construct2(
  demo_smooth_obj, data = demo_test_df, knots = NULL
)

# Adjust penalty matrix for underflow
scale_demo_penalty <- norm(cranio_soap$S[[1]], type = "O") / 
  norm(demo_smooth_test$S[[1]], type = "O")

cat("Scaling age effect penalty matrix by", format(scale_demo_penalty, nsmall = 2))

demo_smooth_test$S[[1]] <- demo_smooth_test$S[[1]] * scale_demo_penalty
demo_smooth_noshrink$S[[1]] <- demo_smooth_noshrink$S[[1]] * scale_demo_penalty
```

These basis functions and their associated penalty matrix $\mathbf{P}_\alpha$ take the following form:

```{r}
#| label: demo-expansion-viz

# Join basis function values with age
demo_expansion_df <- demo_smooth_test$X %>% 
  reshape2::melt() %>% 
  rename(row = Var1, fn = Var2) %>% 
  left_join(demo_test_df %>% mutate(row = 1:n()))

# Plot basis functions

demo_basis_plot <- ggplot(demo_expansion_df) +
  geom_line(aes(x = age, y = value, color = as.factor(fn))) +
  scale_color_manual(values = unname(pals::alphabet2())) +
  geom_point(inherit.aes = F,
             data = enframe(demo_smooth_test$xp),
             aes(x = value, y = 0),
             fill = "yellow", shape = 23) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(x = "Age", y = "Function Value", color = "Basis Function")

demo_basis_plot

# Plot penalty matrix
demo_expansion_s_plot <- demo_smooth_test$S[[1]] %>% 
  reshape2::melt() %>% 
  rename(row = Var1, col = Var2) %>% 
  mutate(type = "shrink") %>% 
  bind_rows(
    demo_smooth_noshrink$S[[1]] %>% 
      reshape2::melt() %>% 
      rename(row = Var1, col = Var2) %>% 
      mutate(type = "no shrink")
  ) %>% 
  ggplot(aes(x = col, y = -row, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(transform = "pseudo_log",
                       breaks = c(-1 * c(10^c(0:log10(max(demo_smooth_test$S[[1]])), 1)),
                                  c(10^c(0:log10(max(demo_smooth_test$S[[1]])), 1)))) +
  scale_x_continuous(breaks = scales::breaks_width(1)) +
  scale_y_continuous(breaks = scales::breaks_width(1)) +
  facet_wrap(~type) +
  labs(title = "Demographics Penalty Matrix")

demo_expansion_s_plot

demo_s_diff_plot <- (demo_smooth_test$S[[1]] - demo_smooth_noshrink$S[[1]]) %>% 
  reshape2::melt() %>% 
  rename(row = Var1, col = Var2) %>% 
  ggplot(aes(x = col, y = -row, fill = value)) +
  geom_tile() +
  scale_fill_gradient2() +
  scale_x_continuous(breaks = scales::breaks_width(1)) +
  scale_y_continuous(breaks = scales::breaks_width(1)) +
  labs(title = "Demographics Penalty Shrinkage Difference")

demo_s_diff_plot
```

For our model assumptions, we assume that these basis functions are added as covariates in an additive model that already contains an intercept. Therefore, we will have identifiability constraints if we keep all $L+1$ basis functions describing age. To address this issue, we center each of the basis functions and remove the last one. Likewise, we remove the last row and column of the penalty matrix from this smooth.

In addition, we want to estimate the effect of each suture fusion type as a deviation from Normative growth associated with age. Therefore, we transform the design matrix of basis functions into a "reference-coded" set of basis expansions. The nonlinear age effect for all patients is estimated using the normative set of basis functions, and a separate nonlinear age-related effect is estimated for each of the suture fusions as an additive effect on top of this normative effect.

```{r}
#| label: setup-reference-smooth

# Attach name of the dataset to this smooth
attr(demo_smooth_test, "data") <- "demo_test_df"

# Use function to reference-code smooth by fusion type
demo_ingredients_test <- setup_reference_model(
  sm = demo_smooth_test,
  by_var = "fusion_type", param_formula = ~sex
)

# Combine suture fusion penalties so we estimate the same lambda
demo_penalty_list_test <- list(demo_ingredients_test$S[[1]],
                               Reduce("+", demo_ingredients_test$S[-1]))

## Also identify which parameter is in which category

# Determine which penalty columns are non-zero
demo_params_cols <- map(demo_ingredients_test$S,
                        ~which(
                          apply(., 2,
                                function(x){ sum(x != 0)}) != 0))

# Set number of non-penalized terms
Q_u <- length(demo_params_cols[[1]])

# Map each parameter to a penalty number and classify
demo_params_dict <- tibble(
  block_num = 1:length(demo_params_cols), 
  param_num = demo_params_cols
  ) %>% 
  unnest(param_num) %>% 
  mutate(effect_cat = case_match(block_num,
                                 1 ~ "Theta",
                                 2 ~ "Normative",
                                 3 ~ "Fusion 1",
                                 4 ~ "Fusion 2")) %>% 
  select(-block_num)
```

To illustrate the result, we summarize the average basis function values by basis function and by suture type. We also plot the values in the penalty matrices representing the two parametric covariates and the 27 penalized covariates. The first penalty matrix will not multiplied by a smoothing constant, and the second will be multiplied by $\lambda_\alpha$.

```{r}
#| label: ref-smooth-viz

# Reshape design matrix
demo_smooth_x <- demo_ingredients_test$X %>% 
  reshape2::melt() %>% 
  rename(row = Var1, fn = Var2) %>% 
  left_join(demo_test_df %>% mutate(row = 1:n()))

# Summarize function means
demo_smooth_tbl <- demo_smooth_x %>% 
  group_by(fusion_type, fn) %>% 
  summarize(value = mean(value)) %>% 
  ungroup() %>% 
  pivot_wider(names_from = fusion_type, values_from = value) %>% 
  filter(fn >= 3)

kbl(demo_smooth_tbl, 
    caption = "Demographic Column Averages by Fusion", 
    digits = 2) %>% 
  kable_classic(full_width = F) %>% 
  column_spec(2, color = "white",
              background = spec_color(na_if(demo_smooth_tbl[[2]], 0))) %>% 
  column_spec(3, color = "white",
              background = spec_color(na_if(demo_smooth_tbl[[3]], 0))) %>% 
  column_spec(4, color = "white",
              background = spec_color(na_if(demo_smooth_tbl[[4]], 0)))

## Plot penalty matrix

# Reshape (total) covariance matrix
demo_smooth_s <- map(
  1:length(demo_penalty_list_test), 
  ~ demo_penalty_list_test[[.x]] %>% 
    reshape2::melt() %>% 
    rename(row = Var1, col = Var2) %>% 
    mutate(index = .x)
) %>% 
  bind_rows()

demo_s_plot <- demo_smooth_s %>% 
  group_split(index) %>% 
  map(
    ~ggplot(., aes(x = col, y = -row, fill = value)) +
      geom_tile() +
      facet_wrap(~index, nrow = 1) +
      scale_fill_gradient2(transform = "pseudo_log",
                       breaks = c(-1 * c(10^c(0:log10(max(demo_smooth_s$value)), 1)),
                                  c(10^c(0:log10(max(demo_smooth_s$value)), 1)))) +
      scale_x_continuous(breaks = scales::breaks_width(2)) +
      scale_y_continuous(breaks = scales::breaks_width(2)) +
      theme_minimal() +
      theme(legend.position = "bottom") +
      coord_fixed()
  ) %>% 
  ggarrange(plotlist = ., nrow = 1)

demo_s_plot
```

# Data Simulation

With the basis matrix and demographic matrix, we have most of the necessary ingredients to simulate the data. The last step to do so is to define the scalar parameters that will contribute to the distributions we draw from.

## Scalar Parameters

```{r}
#| label: scalar-params

# Scalar parameters for outcome data simulation
sigmasq <- 0.2

tausq <- 0.1

lambdas_basis <- c(0.03, 2)

lambdas_demo <- 5

# Set other arguments for the data simulation function
demo_has_unpenalized = T
basis_mx = cranio_soap$X
demo_mx = demo_ingredients_test$X
penalty_basis_list = cranio_soap$S
penalty_demo_list = demo_penalty_list_test
```

We use the scalar parameters:

$$
\begin{aligned}
\sigma^2 &= `r sigmasq` \quad \text{(Residual variance)} \\
\tau^2 &= `r tausq` \quad \text{(Demographic effect scale)} \\
\lambda_c &= `r lambdas_basis[[1]]` \quad \text{(Smoothing constant for soap film boundary)} \\
\lambda_f &= `r lambdas_basis[[2]]` \quad \text{(Smoothing constant for soap film interior)} \\
\lambda_\alpha &= `r lambdas_demo` \quad \text{(Smoothing constant for age effect functions)}
\end{aligned}
$$
```{r}
#| label: initialize-function-inputs

### Save dimensions
N <- nrow(demo_mx)
M <- nrow(basis_mx)
K <- ncol(basis_mx)
Q <- ncol(demo_mx)
```


## Total Penalty Matrices

Next, we combine the smoothing terms with the given penalty matrices to construct the "total" penalty matrices. This involves two steps:

$$
\begin{aligned}
\mathbf{P}_T^{(\gamma)} &= \lambda_c 
\begin{bmatrix}
\mathbf{P}_c & \boldsymbol{0} \\
\boldsymbol{0} & \boldsymbol{0}
\end{bmatrix} + 
\lambda_f
\begin{bmatrix}
\boldsymbol{0} & \boldsymbol{0} \\
\boldsymbol{0} & \mathbf{P}_f
\end{bmatrix} \\
\mathbf{P}_T^{(\beta)} &= 1
\begin{bmatrix}
\mathbf{I}_{Q_U} & \boldsymbol{0} & \boldsymbol{0} & \boldsymbol{0} \\
\boldsymbol{0} & \boldsymbol{0} & \boldsymbol{0} & \boldsymbol{0} \\
\boldsymbol{0} & \boldsymbol{0} & \boldsymbol{0} & \boldsymbol{0} \\
\boldsymbol{0} & \boldsymbol{0} & \boldsymbol{0} & \boldsymbol{0}
\end{bmatrix} + 
\lambda_\alpha
\begin{bmatrix}
\boldsymbol{0} & \boldsymbol{0} & \boldsymbol{0} & \boldsymbol{0} \\
\boldsymbol{0} & \mathbf{P}_\alpha & \boldsymbol{0} & \boldsymbol{0} \\
\boldsymbol{0} & \boldsymbol{0} & \mathbf{P}_\alpha & \boldsymbol{0} \\
\boldsymbol{0} & \boldsymbol{0} & \boldsymbol{0} & \mathbf{P}_\alpha
\end{bmatrix}.
\end{aligned}
$$

```{r}
#| label: total-penalty-matrices

### Combine parameters with given matrices

# Add 1 to multiply instead of lambda for unpenalized demographics
lambda_mult_demo <- if(demo_has_unpenalized){
  c(1, lambdas_demo)
} else {
  lambdas_demo
}

### Total Penalty Matrices

# Multiply lambdas by demographic penalty matrix
penalty_total_demo_list <- map2(lambda_mult_demo, penalty_demo_list,
                                ~.x * .y)

penalty_total_basis_list <- map2(lambdas_basis, penalty_basis_list,
                                 ~.x * .y)

# Sum block diagonal matrices together
penalty_total_demo_mx <- Reduce("+", penalty_total_demo_list)

penalty_total_basis_mx <- Reduce("+", penalty_total_basis_list)
```

## Prior Variance Terms

With the total variance matrices constructed, we take their pseudo-inverses. Each penalty matrix is not full-rank, so our prior assumptions for $\boldsymbol{\epsilon}^{(\gamma)}$ and $\boldsymbol{\beta}$ are improper, but *we think* they result in proper posterior distributions.

```{r}
#| label: prior-variance-matrices

### Moments for generated parameters

# Take generalized inverse of total penalty matrices
penalty_var_demo <- ginv(as.matrix(penalty_total_demo_mx))

penalty_var_basis <- ginv(as.matrix(penalty_total_basis_mx))

# Convert to efficient Matrix format
penalty_var_demo <- Matrix(penalty_var_demo)

penalty_var_basis <- Matrix(penalty_var_basis)
```

Last, we multiply each matrix by its respective scaling term, since the prior variance of $\boldsymbol{\epsilon}^{(\gamma)}_{i \bullet}$ is $\sigma^2 \mathbf{P}_T^{(\gamma) -}$, and the prior variance of $\boldsymbol{\beta}_{\bullet k}$ is $\tau^2 \mathbf{P}_T^{(\beta) -}$.

```{r}
#| label: prior-variance-scale

# Scale inverse penalty matrices by scalar variance terms
var_demo_tau <- tausq * penalty_var_demo

var_basis_sigma <- sigmasq * penalty_var_basis
```

## Generated Parameters

Finally, we are able to calculate parameters. We draw $\boldsymbol{\beta}, \ \boldsymbol{\epsilon}^{(\gamma)}, \text{ and } \boldsymbol{\epsilon}^{(G)}$ as described above. We use the function `mvtnorm::rmvnorm` and take draws from a multivariate normal distribution using Singular Value Decomposition.

```{r}
#| label: draw-params

set.seed(413)

### Draw first-level parameters

# Draw betas as K iid Q-length vectors
demo_effect_mx <- Matrix(t(
  rmvnorm(K, mean = rep(0, Q), sigma = as.matrix(var_demo_tau),
          method = "svd")
))

# Draw epsilon-gammas as N iid K-length vectors
epsilon_gamma <- rmvnorm(N, mean = rep(0, K), sigma = as.matrix(var_basis_sigma),
                         method = "svd")

# Draw epsilon-Gs as N*M iid scalars
epsilon_G <- rnorm(N*M, 0, sd = sqrt(sigmasq))

epsilon_G_mx <- Matrix(epsilon_G, N, M)
```

Next, we combine these parameters with the known matrices to generate $\boldsymbol{\gamma}$ and $\mathbf{G}$.

```{r}
#| label: data-and-gamma

# Calculate soap film coefficients
gamma_true <- demo_mx %*% demo_effect_mx

gamma_mx <- gamma_true + epsilon_gamma

# Calculate outcome data
outcome_true <- tcrossprod(gamma_mx, basis_mx)

outcome_mx_sim <- outcome_true + epsilon_G_mx

```

## Parameter Effects

With these generated parameters, we can visualize the implied effects of our demographics and the implied outcomes generated.

```{r}
#| label: without-errors

# Create dense demographic data to visualize effect of betas
demo_newdata <- demo_test_df %>%
  mutate(min_age = min(age),
         max_age = max(age)) %>%
  distinct(fusion_type, min_age, max_age) %>%
  rowwise() %>%
  mutate(age = list(seq(min_age, max_age, length.out = 50))) %>%
  ungroup() %>%
  mutate(sex = 0) %>%
  unnest(age) %>% 
  mutate(sex = factor(sex, levels = levels(demo_test_df$sex)))

# Create design matrix for this dense data
demo_new_design <- setup_reference_model(
  sm = demo_smooth_test, newdata = demo_newdata,
  by_var = "fusion_type", param_formula = ~sex
)$X

# Add rows for effect of fusion deviation curves
demo_newdata_fusions_rows <- which(demo_newdata$fusion_type != levels(demo_newdata$fusion_type)[1])

demo_newdata_dev <- demo_newdata %>% 
  slice(demo_newdata_fusions_rows) %>% 
  mutate(fusion_type = paste0(fusion_type, "_Only"))

demo_new_design_dev <- cbind(
  matrix(0, length(demo_newdata_fusions_rows), (Q_u + demo_age_knots - 1)),
  demo_new_design[demo_newdata_fusions_rows, (Q_u + demo_age_knots):ncol(demo_new_design)]
  )

demo_newdata <- bind_rows(demo_newdata, demo_newdata_dev)

demo_new_design <- rbind(demo_new_design, demo_new_design_dev)

# Calculate true gammas across age
beta_curve_true <- demo_new_design %*% demo_effect_mx

beta_curve_df <- beta_curve_true %>% 
  as.matrix() %>% 
  as.data.frame() %>% 
  mutate(subject = 1:n()) %>% 
  bind_cols(demo_newdata) %>% 
  pivot_longer(matches("V"), names_to = "Gamma") %>% 
  mutate(Gamma = as.numeric(str_remove(Gamma, "V")))

# Identify sample of gammas to plot
gammas_to_check <- c(1:2, 6:7, 10:12, 46:48, 64:65)

```

```{r}
#| label: gamma-viz

beta_effect_plot <- ggplot(beta_curve_df %>% 
                       filter(Gamma %in% gammas_to_check &
                                !str_detect(fusion_type, "Only")),
                     aes(x = age, y = value, color = fusion_type)) +
  geom_line() +
  facet_wrap(~Gamma, labeller = label_both) +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Age", y = "Gamma Value",
       title = "True Simulated Age Effects on Gammas")

beta_effect_plot
```

```{r}
#| label: outcome-viz

# Reshape outcome data without errors for plotting
outcome_long_df <- outcome_true %>% 
  as.matrix() %>% 
  reshape2::melt() %>% 
  rename(subj_num = Var1, loc_num = Var2) %>% 
  left_join(demo_test_df %>% mutate(subj_num = 1:n())) %>% 
  left_join(so_matrix_pts %>% mutate(loc_num = 1:n())) %>% 
  mutate(age_cut = quantcut(age, 5))

# Summarize mean outcomes by age, fusion, and location
outcome_summ_age_fusion <- outcome_long_df %>% 
  group_by(fusion_type, age_cut, loc_num, row, col) %>% 
  summarize(mean_val = mean(value)) %>% 
  ungroup()

# Plot summary by age and fusion
outcome_fusion_age_plot <- ggplot(outcome_summ_age_fusion, 
                                  aes(x = row, y = col, fill = mean_val)) +
  geom_raster() +
  facet_grid(age_cut~fusion_type) +
  coord_fixed() +
  scale_fill_viridis_c() +
  theme_void() +
  theme(legend.position = "bottom") +
  labs(title = "Age-Fusion Summaries of True Outcomes")

outcome_fusion_age_plot
```

# Hierarchical Gibbs Sampler

```{r}
#| label: test Gibbs sampler

# Set arguments
outcome_mx = outcome_mx_sim;
basis_mx = cranio_soap$X;
demo_mx = demo_ingredients_test$X;
pen_basis = cranio_soap$S;
pen_demo = demo_penalty_list_test;
basis_has_unpenalized = FALSE;
demo_has_unpenalized = TRUE;
gamma0 = 0;
beta0 = 0;
sigmasq0 = 1;
tausq0 = 1;
lambda_basis0 = 0.1;
lambda_demo0 = 0.01;
a_sigma_pri = 1;
b_sigma_pri = 1;
a_gamma_pri = 1;
b_gamma_pri = 1;
a_tau_pri = 1;
b_tau_pri = 1;
a_alpha_pri = 1;
b_alpha_pri = 1;
iters = 1e4;
burn_pct = 0.5
```

## Check Low Rank Penalty Matrix

```{r}
#| label: low rank penalty construction

# Steps to create total basis penalty matrix
num_pen_blocks_basis <- length(pen_basis)
num_penals_basis <- num_pen_blocks_basis - as.numeric(basis_has_unpenalized)

# non-zero columns in each penalty matrix
pen_block_cols_basis <- map(pen_basis, ~which(apply(., 2, sum) != 0))

# number of parameters for each penalty
pen_block_dims_basis <- map_dbl(pen_block_cols_basis, length)
  
lambda_basis <- rep_len(lambda_basis0, num_penals_basis)

lambda_basis_mult <- if(basis_has_unpenalized){
      c(1, lambda_basis)
} else { lambda_basis }

pen_basis_lambda <- map2(lambda_basis_mult, pen_basis, ~ .x * .y)

pen_basis_total <- Reduce("+", pen_basis_lambda)

# Steps to create total demographics penalty matrix
num_pen_blocks_demo <- length(pen_demo)
num_penals_demo <- num_pen_blocks_demo - as.numeric(demo_has_unpenalized)
  
lambda_demo <- rep_len(lambda_demo0, num_penals_demo)

lambda_demo_mult <- if(demo_has_unpenalized){
  c(1, lambda_demo)
} else { lambda_demo }

pen_demo_lambda <- map2(lambda_demo_mult, pen_demo, ~ .x * .y)

pen_demo_total <- Reduce("+", pen_demo_lambda)

# Calculation of demographics cross product
demo_sq <- crossprod(demo_mx)

# First matrix in the posterior variance
pen_basis_kron_demo_sq <- kronecker(pen_basis_total, demo_sq)

pen_demo_kron <- kronecker(Diagonal(K), pen_demo_total)

# Full posterior precision
penalties_kron <- pen_basis_kron_demo_sq + (sigmasq / tausq) * pen_demo_kron
  
```

```{r}
#| label: low rank penalty demonstration

# Determine dimension of posterior precision
post_prec_dim <- dim(penalties_kron)

# Rank of posterior precision
post_prec_rank <- qr2rankMatrix(qr(penalties_kron))

cat("Posterior Precision: Dimension - QR Rank =", 
    post_prec_dim[2] - post_prec_rank, "\n")

cat("Is Invertible:", 
    rcond(penalties_kron) >= .Machine$double.eps,
    "with rcond =", rcond(penalties_kron), "\n")

# Data piece of precision
post_data_dim <- dim(pen_basis_kron_demo_sq)

post_data_rank <- qr2rankMatrix(qr(pen_basis_kron_demo_sq))

cat("Posterior Data Kronecker: Dimension - QR Rank =", 
    post_data_dim[2] - post_data_rank)
```

```{r}
#| label: low rank penalty blocks

# Calculate dimensions
N <- nrow(outcome_mx)
M <- ncol(outcome_mx)
K <- ncol(basis_mx)
Q <- ncol(demo_mx)

# Check 2 diagonal blocks of full matrix

index <- 0

diag_block_df <- tibble(block_num = rep(NA, num_penals_basis), 
                        block = rep(NA, num_penals_basis),
                        dim = rep(NA, num_penals_basis), 
                        rank = rep(NA, num_penals_basis))

for(j in 1:num_penals_basis){
  
  first_index <- index + 1
  
  index <- index + pen_block_dims_basis[j] * Q
  
  block_j <- pen_basis_kron_demo_sq[first_index:index, first_index:index]
  
  row_j <- tibble_row(
    block_num <- j,
    block = list(block_j),
    dim = min(dim(block_j)),
    rank = qr2rankMatrix(qr(block_j))
  )
  
  diag_block_df[j,] <- row_j
  
}

diag_block_df %<>% mutate(diff = dim - rank)

kbl(diag_block_df %>% select(-block)) %>% 
  kable_classic(full_width = F)

# Check segment corresponding to cyclic spline

cyclic_kron_demo <- diag_block_df$block[[1]] + 
  kronecker(Diagonal(pen_block_dims_basis[1]), pen_demo_total)

cyclic_kron_dim <- dim(cyclic_kron_demo)

cyclic_kron_rank <- qr2rankMatrix(qr(cyclic_kron_demo))


cat("Posterior Data Cyclic Plus Demo: Dimension - QR Rank =", 
    cyclic_kron_dim[2] - cyclic_kron_rank)
```

## Run Sampler

```{r}
#| label: initialize

# Set initial gamma values
gamma_init <- as(gamma_true, "sparseVector")

set.seed(617)

gamma_init <- gamma_init + 
  rnorm(length(gamma_init), 0, sd(gamma_init))

gamma_init <- as.vector(gamma_init)

# Save true values of sigma squared and tau squared
sigmasq_true <- sigmasq
tausq_true <- tausq

# Set scaled lambda to be substituted in as fixed value
lambda_basis_fix <- c(1, 1) * lambdas_basis

# Replace basis penalty
pen_basis_id <- list(
  bdiag(Diagonal(9), Matrix(0, 60, 60)),
  cranio_soap$S[[2]]
)
```


```{r}
#| label: gibbs test

if(DO_GIBBS){
# Run Gibbs sampler
hierarchy_test <- hierarchical_penalized_gibbs(
  outcome_mx = outcome_mx_sim,
  basis_mx = cranio_soap$X,
  demo_mx = demo_ingredients_test$X,
  pen_basis = cranio_soap$S,
  pen_demo = demo_penalty_list_test,
  basis_has_unpenalized = FALSE,
  demo_has_unpenalized = TRUE,
  gamma0 = gamma_init,
  beta0 = 0,
  sigmasq0 = 1,
  tausq0 = 1,
  lambda_basis0 = 0.1,
  lambda_demo0 = 0.01,
  a_sigma_pri = 1,
  b_sigma_pri = 1,
  a_gamma_pri = 1,
  b_gamma_pri = 1,
  a_tau_pri = 1,
  b_tau_pri = 1,
  a_alpha_pri = 1,
  b_alpha_pri = 1,
  iters = 20,
  burn_pct = 0.5,
  update_pct = 0.1
)

save(hierarchy_test, 
     file = here::here("analysis", "intermediate", 
                       paste0("hierachical_test_", 
                              format(Sys.time(), "%Y-%m-%d"),
                              ".rda")))
} else {
  load(file = here::here("analysis", "intermediate",
                         paste0("hierachical_test_",
                                LOAD_GIBBS_DATE, 
                                ".rda")))
}
```

```{r}
#| label: trace data

# Turn parameters into long table for trace plotting
hier_gibbs_tr <- imap(
  # Remove timing from output
  hierarchy_test %>%
    list_modify("timing" = zap(),
                "start-stop" = zap()),
  function(x, xn){
    
    x_dim <- length(dim(x))
    
    # Turn array into long format
    x <- melt(x) %>% mutate(param_cat = xn)
    
    # Rename to identify iterations and parameter numbers
    if(x_dim == 2){
      
      x <- rename(x, param_num = Var1, iter = Var2)
      
    } else if (x_dim == 3){
      
      x <- rename(x, row_num = Var1, col_num = Var2, iter = Var3)
      
    }
    
    x
  }
) %>%
  # Combine
  list_rbind() %>%
  # Create ID for parameters to be identified
  group_by(param_cat) %>% 
  mutate(row_id = str_pad(row_num, 
                          ceiling(log10(max(row_num))), 
                          "left", "0"),
         col_id = str_pad(col_num, 
                          ceiling(log10(max(col_num))), 
                          "left", "0"),
         param1_id = str_pad(param_num, 
                             ceiling(log10(max(param_num))), 
                             "left", "0"),
         is_scalar = if_else(!is.na(max(param_num)) & max(param_num) == 1, 
                             1, 0)) %>% 
  ungroup() %>% 
  # Identify each unique parameter by row and column or by index in matrix
  mutate(param_id = if_else(is.na(param_num), 
                            paste0(row_id, "_", col_id),
                            param1_id)) %>% 
  select(-c(row_id, col_id, param1_id)) %>% 
  # Identify burn-in vs. posterior distribution
  mutate(draw_cat = if_else(str_detect(param_cat, "burn"), 
                            "Burn-In", "Post-Burn"),
         param_clean = str_remove(param_cat, "burn_")) %>% 
  mutate(parameter = if_else(is_scalar == 1, param_clean,
                             paste0(param_clean, "_", param_id))) %>% 
  select(-is_scalar) %>% 
  # Add true values
  rowwise() %>% 
  mutate(
    value_true = ifelse(
      param_clean == "sigma_sq",  sigmasq,
      ifelse(
        param_clean == "tau_sq", tausq,
        ifelse(
          param_clean == "lambda_basis", lambdas_basis[param_num],
          ifelse(
            param_clean == "lambda_demo", lambdas_demo[param_num],
            ifelse(
              param_clean == "beta", 
              as.matrix(demo_effect_mx)[row_num, col_num],
              ifelse(
                param_clean == "gamma", as.matrix(gamma_true)[row_num, col_num]
              ))))))
  ) %>% 
  ungroup()
```

```{r}
#| label: trace plots

# Trace plot of sigma
basis_trace_hier <- ggplot(hier_gibbs_tr %>%
                              filter(str_detect(parameter, "sigma_sq|lambda_basis"))) +
  geom_line(aes(x = iter, y = value)) +
  geom_hline(aes(yintercept = value_true), color = "blue") +
  facet_wrap(~interaction(draw_cat, parameter), 
             ncol = 2, scales = "free") +
  labs(title = "Soap Film Variance Traces")

basis_trace_hier

# Trace plot of tau
demo_trace_hier <- ggplot(hier_gibbs_tr %>%
                              filter(str_detect(parameter, "tau_sq|lambda_demo"))) +
  geom_line(aes(x = iter, y = value)) +
  geom_hline(aes(yintercept = value_true), color = "blue") +
  facet_wrap(~interaction(draw_cat, parameter), 
             ncol = 2, scales = "free") +
  labs(title = "Demographic Variance Traces")

demo_trace_hier
```

```{r}
#| label: coefficient scatter plots

# Identify gamma parameters and true values
gamma_vals <- hier_gibbs_tr %>%
  filter(param_clean == "gamma" & draw_cat != "Burn-In") %>% 
  group_by(parameter, row_num, col_num, value_true) %>% 
  summarize(post_mean = mean(value)) %>% 
  ungroup() %>% 
  left_join(demo_test_df %>% mutate(row_num = 1:n())) %>% 
  mutate(bias = post_mean - value_true,
         soap_cat = if_else(col_num <= 9, "Boundary", "Interior"))

# Plot gamma estimates against true values
gamma_scatter <- ggplot(gamma_vals, 
                        aes(x = value_true, 
                            y = post_mean, 
                            color = fusion_type,
                            shape = soap_cat)) +
  geom_point(alpha = 0.25) +
  scale_shape_manual(values = c("Boundary" = 17,
                                "Interior" = 16)) +
  coord_fixed() +
  labs(x = "True Value", y = "Estimate",
       title = "Gamma (Soap Film) Coefficients")

gamma_scatter

# Identify beta parameters and true values
beta_vals <- hier_gibbs_tr %>%
  filter(param_clean == "beta" & draw_cat != "Burn-In") %>% 
  group_by(parameter, row_num, col_num, value_true) %>% 
  summarize(post_mean = mean(value)) %>% 
  ungroup() %>% 
  mutate(bias = post_mean - value_true,
         soap_cat = if_else(col_num <= 9, "Boundary", "Interior")) %>%
  left_join(demo_params_dict, by = c("row_num" = "param_num")) %>% 
  mutate(effect_cat = fct_inorder(effect_cat))

beta_scatter <- ggplot(beta_vals, 
                       aes(x = value_true,
                           y = post_mean,
                           color = effect_cat,
                           shape = soap_cat)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, alpha = 0.5) +
  geom_point(alpha = 0.25) +
  scale_shape_manual(values = c("Boundary" = 17,
                                "Interior" = 16)) +
  facet_wrap(~soap_cat, scales = "free") +
  labs(x = "True Value", y = "Estimate",
       title = "Beta (Demographic) Coefficients") +
  ggpubr::stat_cor()

beta_scatter

beta_scatter_bound <- ggplot(beta_vals %>% 
                               filter(soap_cat == "Boundary"), 
                       aes(x = value_true, 
                           y = post_mean,
                           color = effect_cat)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, alpha = 0.5) +
  geom_point(alpha = 0.9) +
  geom_path(alpha = 0.5, arrow = arrow(length = unit(0.1, "inches"), type = "closed")) +
  facet_wrap(~col_num, scales = "free") +
  labs(x = "True Value", y = "Estimate",
       title = "Beta (Demographic) Coefficients")

beta_scatter_bound

beta_bias <- ggplot(beta_vals, aes(x = bias)) + 
  geom_density() +
  scale_x_continuous(breaks = scales::breaks_width(0.02)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| label: demographic curves
#| fig.height: 10
#| fig.width: 10

# Extract posterior mean estimates
beta_post <- apply(hierarchy_test$beta, c(1,2), mean)

beta_curve_est <- demo_new_design %*% beta_post

beta_curve_est_df <- beta_curve_est %>% 
  as.matrix() %>% 
  as.data.frame() %>% 
  mutate(subject = 1:n()) %>% 
  pivot_longer(matches("V"), names_to = "Gamma", values_to = "estimate") %>%
  mutate(Gamma = as.numeric(str_remove(Gamma, "V"))) %>% 
  left_join(beta_curve_df %>% rename(coefficient = value)) %>% 
  pivot_longer(c(estimate, coefficient), names_to = "value_type")

beta_curve_est_plot <- ggplot(beta_curve_est_df %>% 
                           filter(Gamma %in% gammas_to_check &
                                    str_detect(fusion_type, "Only|Norm")),
                         aes(x = age, y = value, 
                             color = fusion_type, linetype = value_type)) +
  geom_line() +
  facet_wrap(~Gamma, labeller = label_both) +
  scale_linetype_manual(values = c("coefficient" = 1,
                                   "estimate" = 2)) +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Age", y = "Gamma Value",
       title = "Simulated Age Effects")

beta_curve_est_plot

```

